### 2018-02-14 paper.Rmd
outline and papaja

---
title             : "childes-db: a flexible and reproducible interface to the Child Language Data Exchange System"
shorttitle        : "childes-db: an interface to CHILDES"

author: 
  - name          : "Alessandro Sanchez*"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Postal address"
    email         : "sanchez7@stanford.edu"
  - name          : "Stephan Meylan*"
    affiliation   : "2"
  - name          : "Mika Braginsky"
    affiliation   : "3"
  - name          : "Kyle E. MacDonald"
    affiliation   : "1"
  - name          : "Daniel Yurovsky"
    affiliation   : "4"
  - name          : "Michael C. Frank"
    affiliation   : "1"

affiliation:
  - id            : "1"
    institution   : "Stanford University"
  - id            : "2"
    institution   : "University of California, Berkeley"
  - id            : "3"
    institution   : "MIT"
  - id            : "4"
    institution   : "University of Chicago"

author_note: |
  Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.

  Enter author note here.

abstract: |
  The Child Language Data Exchange System (CHILDES) has played a critical role in research on child language development. However, traditional methods of inquiry into the data have often relied on a command line interface to extract developmental indices or custom parsers to extract data for computational modeling. Now, advances in open-source software allow for novel approaches in working with this dataset. We introduce childes-db, a tool that improves data accessibility and usability by offering novel interfaces to CHILDES, including browsable web applications and an R application programming interface.  Along with a versioned infrastructure that facilitates reproduction of past analyses, these interfaces lower barriers to analyzing naturalistic parent-child language, allowing for a wider range of researchers in language and cognitive development to easily leverage “data from beyond the lab” in their work.

keywords          : "keywords"
wordcount         : "X"

bibliography      : ["r-references.bib"]

figsintext        : no
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : yes

lang              : "english"
class             : "man"
output            : papaja::apa6_pdf
---

```{r load_packages, include = FALSE}
library("papaja")
```

# Introduction

Established in 1984, CHILDES archives tens of thousands of transcripts of children’s speech as well as that of their caregivers, making it an invaluable resource for characterizing children’s early productive language use as well as their language environment. Researchers have used the dataset for studies including estimating children’s latent vocabulary size, characterizing their lexical diversity, evaluating the prevalence of negative examples to children, and retrieving the frequencies of various syntactic constructions, among thousands of others. As the first major effort to consolidate and share transcripts of child language, CHILDES has been a pioneer in the move to curate and disseminate large-scale behavioral datasets publicly.

Uses
Uptake
Intake
Theory / growing interest in child language growth (roger brown etc)
Why dataset is successful in light of theoretical interests discussed
Challenges working with CHILDES
Expanding the range of use cases (beginner / expert)
Pulling out for modeling: 
People write their own parsers/pipelines
Duplicated effort
Versioning and reproducibility
Can’t easily integrate clan calls with pipeline
Solution (childes-db)

At its core is a series of linked tabular data stores, currently implemented in MySQL, which can be accessed from dozens of programming languages. The second is an R package called childesr that provides abstractions for the most common complex queries, obviating the need to write SQL. childesr exposes preprocessed, tabular data from CHILDES through its application programming interface (API), providing researchers with the opportunity to exploit conveniently formatted data in a familiar programming environment. Finally, we provide  a collection of browser-based interactive applications that enable visual exploration of CHILDES, allowing researchers to easily view and download useful information from the corpus including mean length of utterance (MLU) and word frequency counts.

# The Architecture of childes-db

High level discussion of features

## Childes-db

Database Architecture
MySQL
Extensibility
Versioning

## Cross-linguistic organization

## Interactive Analysis Tools

#Intended use cases

## Freq App 
Go into depth on color 
MLU App
Population App

# The childesr Package

Discuss more intensive use cases for the corpora

## Package Details

## Worked Examples

### Color frequency

non-stationarity

### Gender

# Teaching with childes-db

## Child-language class tutorials
## Data-science tutorials

# Conclusion

reiterate importance

\newpage

# References
```{r create_r-references}
r_refs(file = "r-references.bib")
```

\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

### 2018-03-13 paper.Rmd
color over time

@@
 library("papaja")
+library(tidyverse)
+library(childesr)
+library(directlabels)
 ```
@@
 
+```{r}
+colors <- c("black", "white", "red", "green", "yellow", "blue", "brown", 
+             "orange", "pink", "purple")
+
+graph_colors <- data_frame(color = colors,
+                           graph_color = if_else(color == "white", 
+                                                 "light gray", color))
+
+color_counts <- get_types(role = c("Mother", "Father"), age = c(12,60), 
+                              type = colors)
+
+count_estimates <- color_counts %>%
+  mutate(age = target_child_age / 30.5,
+         age_bin = as.integer(floor(age / 6) * 6),
+         color = tolower(gloss)) %>%
+  group_by(age_bin, color, target_child_id) %>%
+  summarise(count = sum(count)) %>%
+  summarise(mean = mean(count, na.rm = T),
+            sem = sd(count, na.rm = T) / sqrt(sum(!is.na(count)))) %>%
+  left_join(graph_colors)
+
+ggplot(count_estimates, aes(x = age_bin, y = mean, ymin = mean - sem, 
+                            ymax = mean + sem, color = color, label = color)) + 
+  geom_pointrange(position = position_dodge(.5)) + 
+  geom_smooth(se = F) +
+  facet_wrap(~ color) + 
+  theme_classic(base_size = 14) +
+  scale_color_manual(values = graph_colors$graph_color) +
+  scale_x_continuous(breaks = seq(12, 60, 6)) + 
+  geom_dl(method = list(dl.trans(x=x +.2), "last.qp", cex=1)) + 
+  theme(legend.position = "none")
+```
+
 non-stationarity
@@
 ```{r create_r-references}
-r_refs(file = "r-references.bib")
+r_refs(file = "childesd.bib")
 ```

### 2018-03-15 paper.Rmd
added architecture section

@@
 
-Database Architecture
-MySQL
-Extensibility
-Versioning
+Unlike the *parsing* approach in CLAN which entails the sequential processing of strings, childes-db treats CHILDES as a set of linked tables, with records corresponding to intuitive abstractions such as words, utterances, and transcripts.
+Users of data analysis languages like R or Julia, libraries like Pandas, or those familiar with Structured Query Language (SQL) will be familiar with operations on tables such as filtering (subsetting), sorting, aggregation (grouping), and joins (merges).
+We expect most users to interact with the database through the childesr API for the R language described below, we begin with a top-level description and motivation for the design of the database schema, details of the database’s current technical implementation, and our versioning scheme.
 
-## Cross-linguistic organization
+At its core, childes-db is a database consisting of a set of linked tabular data stores where rows corresponding to records.
+The smallest unit of abstraction tracked by the database is a token, treated here as the citation orthographic form of a word.
+Using the standardized written form of the word facilitates the computation of lexical frequency statistics, for comparison or aggregation across children or time periods.
+Deviations from the citation form—reflecting the deviations from citation forms that are particularly common in the course of language development— are kept as a separate (possibly null) field associated with each token.
+
+Other tables in the database comprise a hierarchical system of collections built on top of tokens—*utterance*, *transcription*, *corpus*, and *collection*—which stores attributes appropriate for each level of description.
+Every entity includes attributes that link it to all higher-order collections, e.g., an utterance lists the transcript, corpus, and collection to which it belongs.
+An *utterance* contains one or more words and includes fields such as the utterance type (declarative, question, etc.), total number of tokens, and the total number one morphemes.
+[[ How are we handling the long tail of tiers?]].
+A *transcription* consists of one or more utterances and includes the date collected, the name of the target child and age in days (if defined), and the filename from CHILDES.
+A *corpus* consists of one or more transcripts, corresponding to well-known collections like the Brown or Providence corpus.
+Finally, a *collection* is a superordinate collection of corpora generally corresponding to a geographic region, following the convention in CHILDES.
+Because every entity can be linked to a top-level collection, childes-db places content from all languages in the same table for a given level of abstraction.
+
+We chose to treat participants separate from the above hierarchy because children can occur across transcripts (e.g., the Brown Corpus).
+A participant is associated with every word and utterance, including a name, role, 3-letter CHILDES identifier (CHI, MOT, FAT, etc.), and the range of ages (or age of corresponding child) for which they are observed.
+For non-child participants, the record contains an identifier for the target child to associate caregivers and children.
+Technical Implementation
+
+The above tabular datastore is implemented as a MySQL database.
+As an industry-standard, open-source relational database server, the above database can be accessed directly from a wide range of programming languages.
+The childes-db project provides hosted read-only databases for direct access and for `childesr` (described below) as well as compressed .sql exports for local installation.
+While the former is appropriate for casual users, local installation can provide performance gains (allowing a user to access the database on their local machine or on their network) as well as write support in case users wish to store derived information in the same database.
 

### 2018-03-15 paper.Rmd
R package section

@@
 ```{r load_packages, include = FALSE}
-library("papaja")
+library(papaja)
 library(tidyverse)
@@
 library(directlabels)
+knitr::opts_chunk$set(cache = TRUE)
 ```
@@
 
-At its core is a series of linked tabular data stores, currently implemented in MySQL, which can be accessed from dozens of programming languages.
-The second is an R package called childesr that provides abstractions for the most common complex queries, obviating the need to write SQL.
-childesr exposes preprocessed, tabular data from CHILDES through its application programming interface (API), providing researchers with the opportunity to exploit conveniently formatted data in a familiar programming environment.
+At its core is a series of linked tabular data stores, currently implemented in `MySQL`, which can be accessed from dozens of programming languages.
+The second is an `R` package called `childesr` that provides abstractions for the most common complex queries, obviating the need to write `SQL`.
+`childesr` exposes preprocessed, tabular data from CHILDES through its application programming interface (API), providing researchers with the opportunity to exploit conveniently formatted data in a familiar programming environment.
 Finally, we provide  a collection of browser-based interactive applications that enable visual exploration of CHILDES, allowing researchers to easily view and download useful information from the corpus including mean length of utterance (MLU) and word frequency counts.
@@
 Users of data analysis languages like R or Julia, libraries like Pandas, or those familiar with Structured Query Language (SQL) will be familiar with operations on tables such as filtering (subsetting), sorting, aggregation (grouping), and joins (merges).
-We expect most users to interact with the database through the childesr API for the R language described below, we begin with a top-level description and motivation for the design of the database schema, details of the database’s current technical implementation, and our versioning scheme.
+We expect most users to interact with the database through the `childesr` API for the `R` language described below, we begin with a top-level description and motivation for the design of the database schema, details of the database’s current technical implementation, and our versioning scheme.
 
@@
 
-# The childesr Package
+# The `childesr` Package
 
-Discuss more intensive use cases for the corpora
+Although the interactive analysis tools described above cover some of the most common use cases of CHILDES data, researchers interested in more detailed and flexible analyses will want to interface directly with the data in childes-db.
+Making use of the `R` programming language [@R-base], we provide the `childesr` package to help researchers accomplish this task.
+`R` is an open-source, extensible statistical computing environment that is rapidly growing in popularity across fields and is increasing in use in child language research [e.g.
+@song2015; @norrman2015].
+The `childesr` package abstracts away the details of connecting to the database.
+Users can take advantage of the tools developed in the popular `dplyr` package [@R-dplyr], which make manipulating large datasets quick and easy.
+We describe the commands that the package provides and then give several worked examples of using the package for analyses.
 
 ## Package Details
+
+The `childesr` package is easily installed via CRAN, the comprehensive R archive network.
+To install, simply type: `install.packages("childesr")`.
+After installation, users have access to data loading functions that retrieve a variety of subsets of the data:
+
+- `get_collections()` gives the names of available collections of corpora ("Eng-NA", "Spanish", etc.)
+- `get_corpora()` gives the names of available corpora ("Brown", "Clark", etc.)
+- `get_transcripts()` gives information on available transcripts (language, date, target child demographics)
+- `get_participants()` gives information on transcript participants (name, role, demographics)
+- `get_speaker_statistics()` gives summary statistics for each participant in each transcript (number of utterances, number of types, number of tokens, mean length of utterance)
+- `get_utterances()` gives information on each utterance (glosses, stems, parts of speech, utterance type, number of tokens, number of morphemes, speaker information, target child information)
+- `get_types()` gives information on each type within each transcript (gloss, count, speaker information, target child information)
+- `get_tokens()` gives information on each token (gloss, stem, part of speech, number of morphemes, speaker information, target child information)
+
+Each of these functions can be supplied with arguments that restrict the query to a particular subset of the data (e.g.
+by collection, by corpus, by speaker role, by target child age, etc.) For more detailed documentation, see the package repository (http://github.com/langcog/childesr).
 

### 2018-03-15 paper.Rmd
interactive analysis tools

@@
 
-#Intended use cases
+childes-db offers a collection of browser-based interactive visualizations that enable exploration of the data through easy-to-use interfaces that don’t require any programming knowledge.
+These tools are intended for casual users who seek to extract useful developmental indices from CHILDES without the technical overhead.
+The visualizations available are the following: Frequency Counts (for lexical usage statistics), MLU (to chart syntactic growth), and Population Viewer (to get a better understanding of the size of various corpora in the database).
+All three of these visualizations were built using Shiny, an R package that enables easy building of web applications in R.
+For each visualization, a user specifies a subset of the data, in particular the collection, corpus, child (target child in the session), and speaker (mother, father, etc.).
+The user may also restrict the age range of the child with slider bars.
+Any number of corpora, children, and speakers may be selected, allowing for aggregate counts across a whole range of data, a task that could not have been done with ease using previous methods.
+Along with the visualizations, childes-db makes the raw data itself available for download under a separate “Table” tab; this allows researchers to further explore and analyze the data they see in the plots.
 
-## Freq App 
-Go into depth on color 
-MLU App
-Population App
+### Frequency Counts
+
+The Frequency Counts app allows users to enter a word in a text box and see the number of times this word was uttered across age for a particular subset of the data (in parts per million), whether it is an individual child in the Brown corpus or all Spanish speaking children and their mothers The ability to view the caregiver’s usage of a term alongside a child’s assists researchers in better understanding the relationship between input and uptake.
+Users can also plot word trajectories alongside each other (such as “mom” and “dad” or “front” and “back”) for convenient cross-word comparisons.
+
+### MLU
+
+MLU (mean length of utterance) is an important measure used to assess the syntactic complexity in children’s speech.
+The MLU app plots MLU curves across age for a given subset of data, again specified by collection, corpora, children, and speakers.
+Because of the underlying gaps in available CHILDES morphological data, MLUw (mean length of utterance in words) is computed and not MLUm (mean length of utterance in morphemes).
+In the English language, these two measures have been shown to be highly correlated.
+As with the Frequency Counts app, caregiver’s MLU can be plotted alongside children.
+
+### Population Viewer
+
+The population viewer is a tool that helps users understand the relative sizes of various corpora by displaying the total number of utterances or tokens in the available corpora at particular ages.
+In many instances the size of a given corpus may inform researchers’ decisions as to what is the best target dataset for analysis.
 

### 2018-03-16 paper.Rmd
gender

@@
 
+```{r}
+stats <- get_speaker_statistics(role = "Target_Child")
+
+data <- stats %>%
+  filter(!is.na(target_child_sex), collection_name == "Eng-NA") %>%
+  group_by(target_child_id, target_child_age, target_child_sex, language) %>%
+  summarise(ttr = sum(num_types) / sum(num_tokens))
+
+ggplot(data, aes(x = target_child_age, y = ttr, color = target_child_sex)) +
+  geom_smooth()
+```
+
 # Teaching with childes-db

### 2018-03-16 paper.Rmd
use language instead of collection

--- 2018-03-15 paper.Rmd
@@
 
+```{r}
+stats <- get_speaker_statistics(role = "Target_Child")
+
+data <- stats %>%
+  filter(!is.na(target_child_sex), language == "eng") %>%
+  group_by(target_child_id, target_child_age, target_child_sex, language) %>%
+  summarise(ttr = sum(num_types) / sum(num_tokens))
+
+ggplot(data, aes(x = target_child_age, y = ttr, color = target_child_sex)) +
+  geom_smooth()
+```
+
 # Teaching with childes-db

### 2018-03-16 paper.Rmd
super drafty intro

@@
 
-Established in 1984, CHILDES archives tens of thousands of transcripts of children’s speech as well as that of their caregivers, making it an invaluable resource for characterizing children’s early productive language use as well as their language environment.
+Children's language acquisition is a core puzzle in cognitive science.
+How are the representations underlying the generative use of language learned over their first years of life?
+Because a language is comprised of many individual pieces of information -- the names of particular objects, the marker for the past tense, the position of determiners -- studying language development is fundamentally about studying learning from environmental input, which is the only source of information on these specifics.
+Further, children are often reluctant experimental participants; often their most revealing and most sophisticated uses of language emerge in naturalistic play.
+Thus, language acquisition research must be grounded in the ecological behavior of children, both their language input and their language production in context.
+
+The foundational resource for studies of language acquisition is recordings and transcripts of children.
+Starting with Roger Brown's [-@brown1973] foundational work on Adam, Eve, and Sarah, longitudinal recordings of children's talk (and the talk they hear from their caregivers) has formed the basis for this field.
+But creating individual transcripts is extremely hard work.
+Beyond gathering the data, even orthographic transcriptions (without phonetic detail) can take ten times the length of the original recording to transcribe.
+Thus, researchers' desire for data from multiple ages or multiple children can very quickly outstrip their resources.
+
+Established in 1984 to address this issue, the Child Language Data Exchange System (CHILDES) aims to make transcripts and recording of children's language available for researchers via a free resource [@macwhinney2000].
+CHILDES archives tens of thousands of transcripts of children’s speech, making it an critical resource for characterizing children’s early productive language use as well as their language environment.
 Researchers have used the dataset for studies including estimating children’s latent vocabulary size, characterizing their lexical diversity, evaluating the prevalence of negative examples to children, and retrieving the frequencies of various syntactic constructions, among thousands of others.
@@
 
-Uses
-Uptake
-Intake
-Theory / growing interest in child language growth (roger brown etc)
-Why dataset is successful in light of theoretical interests discussed
-Challenges working with CHILDES
-Expanding the range of use cases (beginner / expert)
-Pulling out for modeling: 
-People write their own parsers/pipelines
-Duplicated effort
-Versioning and reproducibility
-Can’t easily integrate clan calls with pipeline
-Solution (childes-db)
+Yet there are some challenges working with CHILDES, both for students and for advanced.
+For beginners with fewer technical tools, mastering the system can be tricky.
+The CHILDES ecosystem uses a special file format (CHAT), which is read by a command-line tool (CLAN) with a host of specific arguments.
+Mastering the CHAT codes and the CLAN tool can be daunting to navigate for a beginning researcher or student, decreasing the ease of integrating hands-on exploration of the dataset into classes.
 
-At its core is a series of linked tabular data stores, currently implemented in `MySQL`, which can be accessed from dozens of programming languages.
+On the opposite end of the spectrum.
+for data-oriented researchers who are interested in doing large-scale analyses of CHILDES, the current tools are also not ideal.
+Many researchers who would like to ingest the entire corpus (or some large subset) into a computational analysis must write one-off parsers of the CHAT format to extract the subset of the data they would like to use [EXAMPLES].
+This practice is problematic for a number of reasons.
+Not only is effort wasted in rewriting the same parser again and again, but this process introduces a possible source of error in data handling.
+Further, these scripta are rarely shared, leading to much greater difficulty in reproducing the exact numerical results from papers using CHILDES [see e.g., @meylan2017].
+Finally, the CHILDES corpus itself is a moving target; thus computational work using the entire corpus at one time point may include a different set of data (due to new contributions or revisions).
+
+Our goal in the current manuscript is to describe a system for extending the functionality of CHILDES to address these issue.
+childes-db is a database-formatted copy of the CHILDES data that allows programmatic access through an application programming interface (API).
+This infrastructure allows the creation of web applications for browsing and visualizing the data easily, facilitating classroom use of the dataset.
+Further, it can be accessed programmatically by advanced researchers, obviating the need to write one-off parsers of the CHAT format.
+The database is versioned for access to previous images, allowing reproducibility of particular analyses.
+
+<!-- Uses -->
+<!-- Uptake -->
+<!-- Intake -->
+<!-- Theory / growing interest in child language growth (roger brown etc) -->
+<!-- Why dataset is successful in light of theoretical interests discussed -->
+<!-- Challenges working with CHILDES -->
+<!-- Expanding the range of use cases (beginner / expert) -->
+<!-- Pulling out for modeling:  -->
+<!-- People write their own parsers/pipelines -->
+<!-- Duplicated effort -->
+<!-- Versioning and reproducibility -->
+<!-- Can’t easily integrate clan calls with pipeline -->
+<!-- Solution (childes-db) -->
+
+<!-- At its core is a series of linked tabular data stores, currently implemented in `MySQL`, which can be accessed from dozens of programming languages.
 The second is an `R` package called `childesr` that provides abstractions for the most common complex queries, obviating the need to write `SQL`.
@@
 Finally, we provide  a collection of browser-based interactive applications that enable visual exploration of CHILDES, allowing researchers to easily view and download useful information from the corpus including mean length of utterance (MLU) and word frequency counts.
+-->
 

### 2018-03-16 paper.Rmd
color text

@@
 
-bibliography      : ["r-references.bib"]
+bibliography      : ["childesdb.bib"]
 
@@
 
+One common use of the CHILDES corpora is estimating the frequency with which children hear different words.
+These frequency estimates are used both in the development of theory (e.g.
+frequent words are learned earlier; @goodman2008), and in the construction of age-appropriate experimental stimuli.
+One benefit of the childesdb interface is that it allows for easy analysis of how the frequency of words changes over development.
+Many of our theories in which children learn the structure of language from its statistical properties implicitly assume that these statistics are *stationary*, i.e.
+unchanging over development (e.g., @saffran1996).
+However a number of recent analysis show that the frequencies with which infants encounter both linguistic and visual properties of their environment may change dramattically over development [@fausey2016], and these changing distributions may produce similarly dramatic changes in the ease or difficulty with which these regularities can be learned [@elman1993].
+
+To demonstrate how one might discover such a non-staionarity, we take as a case study the frequency with which children hear the color words of english (e.g.
+"blue", "green").
+Color words tend to be learned relatively late by children, potentially in part due to the abstractness of the meanings to which they refer (see @wagner2013).
+However, within the set of color words, the frequency with which these words are heard predicts a significant fraction of the variance in their order of acquisition [@yurovsky2015].
+But are these frequencies stationary--e.g.
+do children hear "blue" as often at 12 months as they do at 24 months?
+We answer this question in two ways--first using the childesb shiny apps, and then using the `childesr` package.
+
 ```{r}
@@
   scale_color_manual(values = graph_colors$graph_color) +
-  scale_x_continuous(breaks = seq(12, 60, 6)) + 
+  scale_x_continuous(breaks = seq(12, 60, 12),
+                     name = "age (months)") + 
   geom_dl(method = list(dl.trans(x=x +.2), "last.qp", cex=1)) + 
@@
 ```{r create_r-references}
-r_refs(file = "childesd.bib")
+r_refs(file = "childesdb.bib")
 ```

### 2018-03-16 paper.Rmd
recoloring color plot

@@
             sem = sd(count, na.rm = T) / sqrt(sum(!is.na(count)))) %>%
-  left_join(graph_colors)
+  left_join(graph_colors) %>%
+  mutate(color = factor(color, levels = colors))
 

### 2018-03-16 paper.Rmd
added details on versioning scheme

@@
 For non-child participants, the record contains an identifier for the target child to associate caregivers and children.
+
 Technical Implementation
@@
 While the former is appropriate for casual users, local installation can provide performance gains (allowing a user to access the database on their local machine or on their network) as well as write support in case users wish to store derived information in the same database.
+
+[[Details of extraction]]
+[[Details of testing: Why trust our parse over your in-house library?]]
+
+Versioning
+
+To reflect changes in the CHILDES database as new corpora are added or previous annotations revised, we produce a new versioned parse of the complete database every [[n]] months.
+By default, users connect to the most recent version.
+To support reproduction of results with previous versions of the database, we provide hosted versions of the last [[m]] versions so that researchers can refer to specific releases.
+Compressed .sql files are available for download for [[r]] years.
 

### 2018-03-16 paper.Rmd
knit settings

--- 2018-03-16 paper.Rmd
@@
 
-```{r load_packages, include = FALSE}
+```{r load_packages, echo = FALSE}
 library(papaja)
@@
 library(directlabels)
-knitr::opts_chunk$set(cache = TRUE)
+knitr::opts_chunk$set(cache = TRUE, echo = FALSE)
 ```

### 2018-03-16 paper.Rmd
added acknowledgements stub
intro - closed #2

@@
 
-  Enter author note here.
+  This work supported by a Jacobs Foundation Fellowship to MCF.
 
@@
 Children's language acquisition is a core puzzle in cognitive science.
-How are the representations underlying the generative use of language learned over their first years of life?
-Because a language is comprised of many individual pieces of information -- the names of particular objects, the marker for the past tense, the position of determiners -- studying language development is fundamentally about studying learning from environmental input, which is the only source of information on these specifics.
+How are the representations underlying the generative use of language learned over children's first years?
+Because a language is comprised of many individual pieces of information -- the names of particular objects, the marker for the past tense, the position of determiners -- language development is fundamentally a question of learning from environmental input, which is the only source of information on these specifics.
 Further, children are often reluctant experimental participants; often their most revealing and most sophisticated uses of language emerge in naturalistic play.
@@
 CHILDES archives tens of thousands of transcripts of children’s speech, making it an critical resource for characterizing children’s early productive language use as well as their language environment.
-Researchers have used the dataset for studies including estimating children’s latent vocabulary size, characterizing their lexical diversity, evaluating the prevalence of negative examples to children, and retrieving the frequencies of various syntactic constructions, among thousands of others.
 As the first major effort to consolidate and share transcripts of child language, CHILDES has been a pioneer in the move to curate and disseminate large-scale behavioral datasets publicly.
 
-Yet there are some challenges working with CHILDES, both for students and for advanced.
-For beginners with fewer technical tools, mastering the system can be tricky.
-The CHILDES ecosystem uses a special file format (CHAT), which is read by a command-line tool (CLAN) with a host of specific arguments.
-Mastering the CHAT codes and the CLAN tool can be daunting to navigate for a beginning researcher or student, decreasing the ease of integrating hands-on exploration of the dataset into classes.
+Since its inception, a tremendous corpus of research has made use of CHILDES data.
+Individual studies are too numerous to list, but  classics include studies of morphological over-regularization [@marcus1992], distributional learning [@redington1998], word segmentation [@goldwater2009], the role of frequency in word learning [@goodman2008] and many others.
+Some studies analyze individual examples in depth [e.g., @snider2007] while others work with the corpus as a whole [e.g., @meylan2017].
+These are just a few of the thousands of studies that take CHILDES corpora -- transcripts, annotations, and associated media -- as starting points.
 
-On the opposite end of the spectrum.
-for data-oriented researchers who are interested in doing large-scale analyses of CHILDES, the current tools are also not ideal.
-Many researchers who would like to ingest the entire corpus (or some large subset) into a computational analysis must write one-off parsers of the CHAT format to extract the subset of the data they would like to use [EXAMPLES].
+Yet there are some challenges working with CHILDES, both for students and for advanced users.
+For beginners with a lower level of technical experience, mastering the system can be tricky.
+The CHILDES ecosystem uses a special file format (CHAT), which is stored in a plain text format but includes structured annotations grouped into tiers (stored on separate lines) so that information about utterances can be stored with accompanying information such as the phonological or morphological structure of the utterance.
+These files are analyzed via by a command-line tool (CLAN) with a host of specific arguments that allow users to count word frequencies, compute statistics (e.g., mean length of utterance, or MLU).
+While this system is flexible and powerful, mastering the CHAT codes and the CLAN tool can be daunting to navigate for a beginner or student.
+These technical barriers decrease the ease of casual exploration by a novice researcher or in a classroom exercise.
+
+On the opposite end of the spectrum, for data-oriented researchers who are interested in doing large-scale analyses of CHILDES, the current tools are also not ideal.
+CLAN software is an excellent tool for interactive exploration, but -- as a free-standing application -- can be tricky to build into a processing pipeline written in a modern analytic platform such as python or R.
+Thus, researchers who would like to ingest the entire corpus (or some large subset) into a computational analysis typically write their own parsers of the CHAT format to extract the subset of the data they would like to use [e.g., @redington1998,@yang2013,@meylan2017].
+
 This practice is problematic for a number of reasons.
-Not only is effort wasted in rewriting the same parser again and again, but this process introduces a possible source of error in data handling.
+Effort wasted in rewriting the same parser again and again.
+In addition, this process introduces a major source of errors and inconsistencies in data handling due to difficulties dealing with the many special cases in the CHAT standard.
 Further, these scripta are rarely shared, leading to much greater difficulty in reproducing the exact numerical results from papers using CHILDES [see e.g., @meylan2017].
-Finally, the CHILDES corpus itself is a moving target; thus computational work using the entire corpus at one time point may include a different set of data (due to new contributions or revisions).
+This lack of computational reproducibility is widely recognized as problematic because it does not allow researchers to verify or build on published researcher [@donohoe2010,@stodden2016].
+Finally, the CHILDES corpus itself is a moving target;  computational work using the entire corpus at one time point may include a different set of data than subsequent work (due to new contributions or revisions).
 
-Our goal in the current manuscript is to describe a system for extending the functionality of CHILDES to address these issue.
-childes-db is a database-formatted copy of the CHILDES data that allows programmatic access through an application programming interface (API).
+In the current manuscript, we describe a system for extending the functionality of CHILDES to address these issues.
+Our systemn, childes-db, is a database-formatted mirror of CHILDES that allows programmatic access through an application programming interface (API).
 This infrastructure allows the creation of web applications for browsing and visualizing the data easily, facilitating classroom use of the dataset.
 Further, it can be accessed programmatically by advanced researchers, obviating the need to write one-off parsers of the CHAT format.
-The database is versioned for access to previous images, allowing reproducibility of particular analyses.
+The database is versioned for access to previous images, allowing computational reproducibility of particular analyses.
 
-<!-- Uses -->
-<!-- Uptake -->
-<!-- Intake -->
-<!-- Theory / growing interest in child language growth (roger brown etc) -->
-<!-- Why dataset is successful in light of theoretical interests discussed -->
-<!-- Challenges working with CHILDES -->
-<!-- Expanding the range of use cases (beginner / expert) -->
-<!-- Pulling out for modeling:  -->
-<!-- People write their own parsers/pipelines -->
-<!-- Duplicated effort -->
-<!-- Versioning and reproducibility -->
-<!-- Can’t easily integrate clan calls with pipeline -->
-<!-- Solution (childes-db) -->
-
-<!-- At its core is a series of linked tabular data stores, currently implemented in `MySQL`, which can be accessed from dozens of programming languages.
-The second is an `R` package called `childesr` that provides abstractions for the most common complex queries, obviating the need to write `SQL`.
-`childesr` exposes preprocessed, tabular data from CHILDES through its application programming interface (API), providing researchers with the opportunity to exploit conveniently formatted data in a familiar programming environment.
-Finally, we provide  a collection of browser-based interactive applications that enable visual exploration of CHILDES, allowing researchers to easily view and download useful information from the corpus including mean length of utterance (MLU) and word frequency counts.
--->
+We begin by describing the architecture of childes-db and the web applications that we provide.
+We next describe the `childesr` API, which allows for programmatic access to the data.
+We then present two worked examples of specific uses of the system using both the web app and the API.
+The first examines the frequency of color words in children's input; the second quantifies the growth of lexical diversity across genders.
+We end by presenting some suggestions for the use of the childes-db web frontend in classroom teaching.
 
@@
 
-High level discussion of features
+We first present the design and technical approach of childes-db, then describe the interactive analysis tools provided by our web-based frontend.
 
-## Childes-db
+## Design and technical approach
 
@@
 
-Technical Implementation
+### Technical Implementation
 
@@
 
-Versioning
+### Versioning
 
@@
 
-```{r}
+```{r warning=FALSE}
 stats <- get_speaker_statistics(role = "Target_Child")

### 2018-03-16 paper.Rmd
refs

@@
 
-```{r load_packages, echo = FALSE}
+```{r load_packages, echo = FALSE, warning=FALSE}
 library(papaja)
@@
 Individual studies are too numerous to list, but  classics include studies of morphological over-regularization [@marcus1992], distributional learning [@redington1998], word segmentation [@goldwater2009], the role of frequency in word learning [@goodman2008] and many others.
-Some studies analyze individual examples in depth [e.g., @snider2007] while others work with the corpus as a whole [e.g., @meylan2017].
+Some studies analyze individual examples in depth [e.g., @snyder2007] while others work with the corpus as a whole [e.g., @meylan2017].
 These are just a few of the thousands of studies that take CHILDES corpora -- transcripts, annotations, and associated media -- as starting points.
@@
 Further, these scripta are rarely shared, leading to much greater difficulty in reproducing the exact numerical results from papers using CHILDES [see e.g., @meylan2017].
-This lack of computational reproducibility is widely recognized as problematic because it does not allow researchers to verify or build on published researcher [@donohoe2010,@stodden2016].
+This lack of computational reproducibility is widely recognized as problematic because it does not allow researchers to verify or build on published researcher [@donoho2010,@stodden2016].
 Finally, the CHILDES corpus itself is a moving target;  computational work using the entire corpus at one time point may include a different set of data than subsequent work (due to new contributions or revisions).
@@
 
+# Acknowledgements
+- Brian MacWhinney for CHILDES
+- Melissa Kline for ClanToR
+
 \newpage

### 2018-03-30 paper.Rmd
started writing teaching in childes-db section

@@
 library(directlabels)
+library(koRpus)
 knitr::opts_chunk$set(cache = TRUE, echo = FALSE)
@@
 The foundational resource for studies of language acquisition is recordings and transcripts of children.
-Starting with Roger Brown's [-@brown1973] foundational work on Adam, Eve, and Sarah, longitudinal recordings of children's talk (and the talk they hear from their caregivers) has formed the basis for this field.
+Starting with Roger Brown's [-@brown1973] foundational work on Adam, Eve, and Sarah, longitudinal recordings of children's talk (and the talk they hear from their caregivers) have formed the basis for this field.
 But creating individual transcripts is extremely hard work.
@@
 
-At its core, childes-db is a database consisting of a set of linked tabular data stores where rows corresponding to records.
+At its core, childes-db is a database consisting of a set of linked tabular data stores where rows correspond to records.
 The smallest unit of abstraction tracked by the database is a token, treated here as the citation orthographic form of a word.
 Using the standardized written form of the word facilitates the computation of lexical frequency statistics, for comparison or aggregation across children or time periods.
-Deviations from the citation form—reflecting the deviations from citation forms that are particularly common in the course of language development— are kept as a separate (possibly null) field associated with each token.
+Deviations from the citation form -- reflecting the deviations from citation forms that are particularly common in the course of language development -- are kept as a separate (possibly null) field associated with each token.
 
-Other tables in the database comprise a hierarchical system of collections built on top of tokens—*utterance*, *transcription*, *corpus*, and *collection*—which stores attributes appropriate for each level of description.
+Other tables in the database comprise a hierarchical system of collections built on top of tokens -- *utterance*, *transcription*, *corpus*, and *collection* -- which stores attributes appropriate for each level of description.
 Every entity includes attributes that link it to all higher-order collections, e.g., an utterance lists the transcript, corpus, and collection to which it belongs.
@@
 
-childes-db offers a collection of browser-based interactive visualizations that enable exploration of the data through easy-to-use interfaces that don’t require any programming knowledge.
-These tools are intended for casual users who seek to extract useful developmental indices from CHILDES without the technical overhead.
-The visualizations available are the following: Frequency Counts (for lexical usage statistics), MLU (to chart syntactic growth), and Population Viewer (to get a better understanding of the size of various corpora in the database).
+childes-db offers a collection of browser-based interactive visualizations that enable data exploration through easy-to-use interfaces that don’t require any programming knowledge.
+These tools are intended for casual users who seek to extract useful developmental indices from CHILDES without any technical overhead.
+The visualizations available through childes-db are the following: Frequency Counts (lexical statistics), Derived Measures (lexical diversity), and Population Viewer (corpora statistics).
 All three of these visualizations were built using Shiny, an R package that enables easy building of web applications in R.
-For each visualization, a user specifies a subset of the data, in particular the collection, corpus, child (target child in the session), and speaker (mother, father, etc.).
-The user may also restrict the age range of the child with slider bars.
-Any number of corpora, children, and speakers may be selected, allowing for aggregate counts across a whole range of data, a task that could not have been done with ease using previous methods.
-Along with the visualizations, childes-db makes the raw data itself available for download under a separate “Table” tab; this allows researchers to further explore and analyze the data they see in the plots.
+For each visualization, a user specifies a subset of the data (in particular the collection, corpus, child, and speaker) and is also able to restrict the age range of the child with slider bars.
+Any number of corpora, children, or speakers may be selected for analysis, allowing for aggregate counts across a whole range of data in the CHILDES system, a task that could not have been easily done using previous methods.
+Along with the visualizations, childes-db makes the raw data itself available for download under a separate “Table” tab; this allows researchers to further explore the underlying data from the plots.
 
@@
 
-The Frequency Counts app allows users to enter a word in a text box and see the number of times this word was uttered across age for a particular subset of the data (in parts per million), whether it is an individual child in the Brown corpus or all Spanish speaking children and their mothers The ability to view the caregiver’s usage of a term alongside a child’s assists researchers in better understanding the relationship between input and uptake.
+The Frequency Counts app allows users to enter a word (or words) in a text box and see the frequency (in parts per million) in which this word was uttered across a child's age for a particular subset of the data, whether it is an individual child in the Brown corpus or all Spanish speaking children and their mothers.
+The ability to view the caregiver’s usage of a term alongside a child’s assists researchers in better understanding the relationship between input and uptake.
 Users can also plot word trajectories alongside each other (such as “mom” and “dad” or “front” and “back”) for convenient cross-word comparisons.
 
-### MLU
+### Derived Measures
 
-MLU (mean length of utterance) is an important measure used to assess the syntactic complexity in children’s speech.
-The MLU app plots MLU curves across age for a given subset of data, again specified by collection, corpora, children, and speakers.
-Because of the underlying gaps in available CHILDES morphological data, MLUw (mean length of utterance in words) is computed and not MLUm (mean length of utterance in morphemes).
+There are a number of ways to measure the lexical diversity of an individual's productive speech.
+Some of these measures, including MLU-w (mean length of utterance in words), TTR (type-token ratio), MTLD (measure of textual lexical diversity), and voc-d are available in the Derived Measures app, which plots these measures across age for a given subset of data, again specified by collection, corpora, children, and speakers.
+Because of the underlying gaps in available CHILDES morphological data, MLU-w is computed and not MLU-m (mean length of utterance in morphemes).
 In the English language, these two measures have been shown to be highly correlated.
-As with the Frequency Counts app, caregiver’s MLU can be plotted alongside children.
+As with the Frequency Counts app, caregiver’s lexical diversity measures can be plotted alongside children's.
 
@@
 
-color_counts <- get_types(role = c("Mother", "Father"), age = c(12,60), 
-                              type = colors)
+color_counts <- get_types(collection = "Eng-NA", role = c("Mother", "Father"),
+                          age = c(12,60), type = colors)
+
+parents <- color_counts %>%
+  distinct(collection_id, corpus_id, transcript_id, speaker_id) %>%
+  left_join(get_speaker_statistics(collection = "Eng-NA")) %>%
+  select(collection_id, corpus_id, transcript_id, speaker_id, num_tokens)
 
 count_estimates <- color_counts %>%
+  left_join(parents) %>%
   mutate(age = target_child_age / 30.5,
@@
          color = tolower(gloss)) %>%
-  group_by(age_bin, color, target_child_id) %>%
-  summarise(count = sum(count)) %>%
-  summarise(mean = mean(count, na.rm = T),
-            sem = sd(count, na.rm = T) / sqrt(sum(!is.na(count)))) %>%
+  group_by(age_bin, color, transcript_id) %>%
+  summarise(count = sum(count), tokens = sum(num_tokens)) %>%
+  mutate(prop = count/tokens * 1000000) %>%
+  tidyboot::tidyboot_mean(prop) %>%
   left_join(graph_colors) %>%
@@
 
-ggplot(count_estimates, aes(x = age_bin, y = mean, ymin = mean - sem, 
-                            ymax = mean + sem, color = color, label = color)) + 
+ggplot(count_estimates, aes(x = age_bin, y = empirical_stat, 
+                            ymin = ci_lower, ymax = ci_upper,
+                            color = color, label = color)) + 
   geom_pointrange(position = position_dodge(.5)) + 
-  geom_smooth(se = F) +
-  facet_wrap(~ color) + 
+  geom_point() +
+  geom_smooth(method = "loess", se = F, span = .5) +
+  facet_wrap(~ color, scales = "free_y") + 
   theme_classic(base_size = 14) +
@@
   scale_x_continuous(breaks = seq(12, 60, 12),
-                     name = "age (months)") + 
+                     name = "Age (months)") + 
+  scale_y_continuous(name = "Frequency (per million words)") +
   geom_dl(method = list(dl.trans(x=x +.2), "last.qp", cex=1)) + 
@@
 
-## Child-language class tutorials
-## Data-science tutorials
+## In class demonstrations
+
+Teachers of courses on early language acquisition often want to demonstrate interesting changes in children’s early productive language use as well as communicate the richness of information available in children's language environments.
+One method is to present static displays on slides or handouts that show either (a) text from parent-child conversations extracted from CHILDES or (b) data visualizations of changes in a variety of metrics related to production and/or input (e.g., MLU or Frequency).
+The challenge with static displays is that they cannot be modified during the course of teaching and they rely on the instructor predicting what examples will be compelling to their students.
+
+Interactive, in-class demonstrations offer a solution, and can be a powerful way to explain complex concepts.
+One of our goals in creating the browser-based interactive visualization tools in childes-db is to provide instructors with a new tool for communicating what can be learned by analyzing children's language production and input.
+In this section, we walk through one example of using childes-db to teach a key concept in first language acquisition: [todo: XXX].
+We also show how an instructor might make the in-class demonstration interactive via eliciting hypotheses from the class that are operationalized as queries to the interactive visualization tools.
+
+Potential topics: concrete vs. abstract nouns, development of question words, uncertainty terms, verbs vs. nouns vs. function words.
+
+Along with the interactive data visualizations, childes-db makes the raw data itself available for download under a separate "Table" tab.
+This allows researchers to further explore and analyze the data they see in the plots.
+This functionality provides a link to the second teaching application for childes-db: making interesting data accessible for tutorials and take-home programming assignments in more computationally-focused language courses.
+In the next section, we walk through one example of how an instructor could use childes-db to create a tutorial that scaffolds students' ability to ask interesting computational questions using child language data.
+
+## Tutorials and assignments
+
+[todo: pick a topic and write some r code with gaps for the students to fill in]
 

### 2018-04-03 paper.Rmd
added more content to worked example of in-class demo

@@
 
-Teachers of courses on early language acquisition often want to demonstrate interesting changes in children’s early productive language use as well as communicate the richness of information available in children's language environments.
-One method is to present static displays on slides or handouts that show either (a) text from parent-child conversations extracted from CHILDES or (b) data visualizations of changes in a variety of metrics related to production and/or input (e.g., MLU or Frequency).
-The challenge with static displays is that they cannot be modified during the course of teaching and they rely on the instructor predicting what examples will be compelling to their students.
+Teachers of courses on early language acquisition often want to demonstrate interesting changes in children’s early language.
+One teaching method is to present static displays on slides or handouts that show (a) text from parent-child conversations extracted from CHILDES or (b) data visualizations of changes in various metrics related to language production and/or input (e.g., MLU or Frequency).
+The challenge with static displays or handouts is that they cannot be modified during a lecture and thus rely on the instructor selecting examples that they think will be compelling to their students.
 
-Interactive, in-class demonstrations offer a solution, and can be a powerful way to explain complex concepts.
-One of our goals in creating the browser-based interactive visualization tools in childes-db is to provide instructors with a new tool for communicating what can be learned by analyzing children's language production and input.
-In this section, we walk through one example of using childes-db to teach a key concept in first language acquisition: [todo: XXX].
-We also show how an instructor might make the in-class demonstration interactive via eliciting hypotheses from the class that are operationalized as queries to the interactive visualization tools.
+Interactive, in-class demonstrations offer a solution and can be a powerful way to explain complex concepts while increasing student engagement with the course materials.
+One goal in creating the browser-based interactive visualization tools in childes-db is to provide instructors with a new tool for illustrating key concepts in children's language acqusition.
+In this section, we walk through one example using childes-db to show the developmental trajectories of different kinds of words that in children's early production.
+We also highlight ways that instructors could make the in-class demonstration interactive by eliciting hypotheses from students, which are then operationalized as queries to the interactive visualization tools.
 
-Potential topics: concrete vs. abstract nouns, development of question words, uncertainty terms, verbs vs. nouns vs. function words.
+### Worked example
 
-Along with the interactive data visualizations, childes-db makes the raw data itself available for download under a separate "Table" tab.
-This allows researchers to further explore and analyze the data they see in the plots.
-This functionality provides a link to the second teaching application for childes-db: making interesting data accessible for tutorials and take-home programming assignments in more computationally-focused language courses.
+What kinds of words do children first produce?
+This question is a good candidate for an interesting "hook" that gets students thinking about some key issues in child language acquisition.
+Diary studies and large-scale studies using parent report show that children’s first words tend to fall into a fairly small number of categories: people, food, body parts, clothing, animals, vehicles, toys, household objects, routines and activities or states [cite Clark and Fenson].
+
+The key insight is that young children talk about what is going on around them: people they see every day; toys and small household objects they can manipulate; food they themselves can control; clothing they can get off by themselves; animals and vehicles, both of which move and so attract attention; daily routines and activities; and some sound effects.
+
+Flow of the interactive demonstration:
+
+1. introduce the question
+2. ask students to reflect or do a pair-and-share discussion
+3. explain the visualization while showing the trajectory of a single lexical item
+4. elicit hypotheses from students about the kinds of words that children are likely to produce
+5. make real time queries to the interactive visualization tool and talk through the output
+
+```{r in-class-demo-flow}
+# series of plots building up the concept that early vocabularies often consist of 
+```
+
+Here are some ideas for other concepts in child language acqusition that teachers could demonstrate using the current interactive visualization tools in childes-db:
+
+* order of acquisition for poduction of quantifiers or number words
+* function vs. content words
+
+Some areas for future development are to add more information to the interactive visualizations such as part of speech tags, phonological information, and n-grams.
+This would allow teachers to use childes-db to demonstrates concepts such as when children start producing certain phonological patterns (e.g., consonant clusters), the debate about the extent of the noun/verb bias in children's early vocabulary, and the development of particular syntactic constructions.
+
+Finally, in addition to the interactive data visualizations, instructors can use childes-db to download the raw data used to generate the figures.
+This allows teachers an opportunity to have students further explore and analyze the data they see in the plots.
+Moreover, this functionality provides an example of another way childes-db can faciliate language acqusition courses by making data easily-accessible for tutorials and student research projects.
 In the next section, we walk through one example of how an instructor could use childes-db to create a tutorial that scaffolds students' ability to ask interesting computational questions using child language data.
@@
 
-[todo: pick a topic and write some r code with gaps for the students to fill in]
+One of the major goals of courses on applied natural language processing (NLP) is for students to get hands-on experience using state of the art NLP tools [@hearst2005teaching].
+A primary challenge for the instuctor is to decide how much time should be spent teaching the requisite programming skills for accessing and formatting language data, which are typically unstructured.
+One pedagogical strategy is to abstract away these details and avoid having students deal with accessing data and formatting text.
+This approach shifts students' effort towards programming analyses that encourage students to explore the language data and test out particular hypotheses.
+
+In section 2, we outlined the challenges of getting students started with analyzing data from CHILDES.
+Recall that these barriers include learning the details of a special file format (CHAT) and a command-line tool (CLAN) built to specifically for the CHAT format.
+This workflow can be daunting to navigate for a beginner.
+Thus, an instructor would have to spend course time teaching students to use CHAT and CLAN, leaving less time for students to try interesting analyses that get them excited about work on child language acqusition.
+Moreover, CLAN provides a nice tool for interactive exploration, but it does not conntect to a pipeline of modern natural language processing tools (e.g., models of word embeddings @mikolov2013distributed).
+
+Together, childes-db with the `childesr` API provide instructors with a solution to both of these challenges.
+The API gives students an easy-to-learn method for programmatic access to CHILDES data.
+This will reduce the time spent getting data out of CHILDES and allow more time for thinking about the interesting challenges and questions that can be answered by doing NLP with child language data.
+Second, by writing our API in the R programming language, students will gain access to general workflow making that can be built into a piepleine with state of the art NLP tools.
+
+### Worked example
+
+[todo: pick a topic and write some childesr code to get data out of childes and then show how a tutorial or assignment could be written in tidytext or NLTK or RNN to get get students hands on experience with child language data]
 

### 2018-04-04 paper.Rmd
edits

--- 2018-03-30 paper.Rmd
@@
 
-color_counts <- get_types(collection = "Eng-NA", role = c("Mother", "Father"),
-                          age = c(12,60), type = colors)
-
-parents <- color_counts %>%
-  distinct(collection_id, corpus_id, transcript_id, speaker_id) %>%
-  left_join(get_speaker_statistics(collection = "Eng-NA")) %>%
-  select(collection_id, corpus_id, transcript_id, speaker_id, num_tokens)
+color_counts <- get_types(role = c("Mother", "Father"), age = c(12,60), 
+                              type = colors)
 
 count_estimates <- color_counts %>%
-  left_join(parents) %>%
   mutate(age = target_child_age / 30.5,
@@
          color = tolower(gloss)) %>%
-  group_by(age_bin, color, transcript_id) %>%
-  summarise(count = sum(count), tokens = sum(num_tokens)) %>%
-  mutate(prop = count/tokens * 1000000) %>%
-  tidyboot::tidyboot_mean(prop) %>%
+  group_by(age_bin, color, target_child_id) %>%
+  summarise(count = sum(count)) %>%
+  summarise(mean = mean(count, na.rm = T),
+            sem = sd(count, na.rm = T) / sqrt(sum(!is.na(count)))) %>%
   left_join(graph_colors) %>%
@@
 
-ggplot(count_estimates, aes(x = age_bin, y = empirical_stat, 
-                            ymin = ci_lower, ymax = ci_upper,
-                            color = color, label = color)) + 
+ggplot(count_estimates, aes(x = age_bin, y = mean, ymin = mean - sem, 
+                            ymax = mean + sem, color = color, label = color)) + 
   geom_pointrange(position = position_dodge(.5)) + 
-  geom_point() +
-  geom_smooth(method = "loess", se = F, span = .5) +
-  facet_wrap(~ color, scales = "free_y") + 
+  geom_smooth(se = F) +
+  facet_wrap(~ color) + 
   theme_classic(base_size = 14) +
@@
   scale_x_continuous(breaks = seq(12, 60, 12),
-                     name = "Age (months)") + 
-  scale_y_continuous(name = "Frequency (per million words)") +
+                     name = "age (months)") + 
   geom_dl(method = list(dl.trans(x=x +.2), "last.qp", cex=1)) + 
@@
 
-## In class demonstrations
-
-Teachers of courses on early language acquisition often want to demonstrate interesting changes in children’s early productive language use as well as communicate the richness of information available in children's language environments.
-One method is to present static displays on slides or handouts that show either (a) text from parent-child conversations extracted from CHILDES or (b) data visualizations of changes in a variety of metrics related to production and/or input (e.g., MLU or Frequency).
-The challenge with static displays is that they cannot be modified during the course of teaching and they rely on the instructor predicting what examples will be compelling to their students.
-
-Interactive, in-class demonstrations offer a solution, and can be a powerful way to explain complex concepts.
-One of our goals in creating the browser-based interactive visualization tools in childes-db is to provide instructors with a new tool for communicating what can be learned by analyzing children's language production and input.
-In this section, we walk through one example of using childes-db to teach a key concept in first language acquisition: [todo: XXX].
-We also show how an instructor might make the in-class demonstration interactive via eliciting hypotheses from the class that are operationalized as queries to the interactive visualization tools.
-
-Potential topics: concrete vs. abstract nouns, development of question words, uncertainty terms, verbs vs. nouns vs. function words.
-
-Along with the interactive data visualizations, childes-db makes the raw data itself available for download under a separate "Table" tab.
-This allows researchers to further explore and analyze the data they see in the plots.
-This functionality provides a link to the second teaching application for childes-db: making interesting data accessible for tutorials and take-home programming assignments in more computationally-focused language courses.
-In the next section, we walk through one example of how an instructor could use childes-db to create a tutorial that scaffolds students' ability to ask interesting computational questions using child language data.
-
-## Tutorials and assignments
-
-[todo: pick a topic and write some r code with gaps for the students to fill in]
+## Child-language class tutorials
+## Data-science tutorials
 

### 2018-04-04 paper.Rmd
rudimentary color shiny analysis

--- 2018-04-04 paper.Rmd
@@
 One common use of the CHILDES corpora is estimating the frequency with which children hear different words.
-These frequency estimates are used both in the development of theory (e.g.
-frequent words are learned earlier; @goodman2008), and in the construction of age-appropriate experimental stimuli.
-One benefit of the childesdb interface is that it allows for easy analysis of how the frequency of words changes over development.
+These frequency estimates are used both in the development of theory [e.g.
+frequent words are learned earlier; @goodman2008], and in the construction of age-appropriate experimental stimuli.
+One benefit of the `childesdb` interface is that it allows for easy analysis of how the frequencies of words change over development.
 Many of our theories in which children learn the structure of language from its statistical properties implicitly assume that these statistics are *stationary*, i.e.
-unchanging over development (e.g., @saffran1996).
-However a number of recent analysis show that the frequencies with which infants encounter both linguistic and visual properties of their environment may change dramattically over development [@fausey2016], and these changing distributions may produce similarly dramatic changes in the ease or difficulty with which these regularities can be learned [@elman1993].
+unchanging over development [e.g., @saffran1996].
+However a number of recent analysis show that the frequencies with which infants encounter both linguistic and visual properties of their environment may change dramatically over development [@fausey2016], and these changing distributions may produce similarly dramatic changes in the ease or difficulty with which these regularities can be learned [@elman1993].
 
-To demonstrate how one might discover such a non-staionarity, we take as a case study the frequency with which children hear the color words of english (e.g.
+To demonstrate how one might discover such a non-stationary, we take as a case study the frequency with which children hear the color words of English (e.g.
 "blue", "green").
-Color words tend to be learned relatively late by children, potentially in part due to the abstractness of the meanings to which they refer (see @wagner2013).
+Color words tend to be learned relatively late by children, potentially in part due to the abstractness of the meanings to which they refer [see @wagner2013].
 However, within the set of color words, the frequency with which these words are heard predicts a significant fraction of the variance in their order of acquisition [@yurovsky2015].
@@
 
-```{r}
+#### Using shiny
+
+To investigate whether the frequency of color words is stationary over development, a user can navigate to the Frequency app, and enter a set of color words into the `Word` selector separated by a comma: here "blue, red, green." Because the question of interest is about the frequency of words in the input (rather than produced by children), the `Speaker` field can be set to reflect this.
+In this example we select "Mother" because, across corpora in CHIDLES, the speaker tagged with the Mother role tends to produce most of the words that children hear.
+Because children learn most of their basic color words by the age of 5, the age range 1-5 years is a reaonable choice for `Ages to include`.
+The results of these selections are shown in Figure \ref{fig:color-shiny}.
+We can also create a hyperlink to store these set of choices so that we can share these results with others, or with ourselves in the future, byby clicking on the `Bookmark` button in the bottom left corner.
+
+From this figure, it seems likely that children hear "blue" more frequently early in development, but the trajectories of the "red" and "green" are less clear.
+We also do not have a good sense of the errors of these measurements, are limited to just a few colors at a time before the plot becomes too busy, and cannot combine frequencies across speakers.
+To perform this analysis in a more compelling and complete way, a user can use the `childesr` interface to get more direct access to the CHILDES corpora.
+
+```{r color-shiny, fig.width = 10, fig.height = 6, fig.cap = "An example of using the Frequency shiny app to explore how children's color input changes over development"}
+knitr::include_graphics("images/color-shiny.png")
+```
+
+#### Using childesr
+
+Figure \ref{fig:color-plot}.
+
+```{r color-plot, fig.height = 5, fig.width = 10, fig.cap ="Color frequency as a function of age.
+Points represent means across transcripts, error bars represent 95% confidence intervals computed by nonparametric bootstrap", cache = TRUE, message = FALSE, warning = FALSE}
 colors <- c("black", "white", "red", "green", "yellow", "blue", "brown", 

### 2018-04-04 paper.Rmd
finished draft of teaching section

--- 2018-04-04 paper.Rmd
@@
 
-## Child-language class tutorials
-## Data-science tutorials
+## In class demonstrations
+
+Teachers of courses on early language acquisition often want to demonstrate interesting changes in children’s early language.
+One teaching method is to present static displays on slides or handouts that show (a) text from parent-child conversations extracted from CHILDES or (b) data visualizations of changes in various metrics related to language production and input (e.g., MLU or Frequency).
+The challenge with static displays or handouts is that they cannot be modified during a lecture and thus rely on the instructor selecting examples that they think will be compelling to their students.
+
+Interactive, in-class demonstrations offer a solution and can be a powerful way to explain complex concepts while increasing student engagement with the course materials.
+One goal in creating the browser-based interactive visualization tools in childes-db is to provide instructors with a new tool for illustrating key concepts in children's language acquisition.
+In this section, we walk through one example using childes-db to show the developmental trajectories of different kinds of words that in children's early production.
+We also highlight ways that instructors could make the in-class demonstration interactive by eliciting hypotheses from students, which become operationalized as queries to the interactive visualization tools.
+
+### Worked example
+
+What kinds of words do children first produce?
+This question is a good candidate for an interesting "hook" that gets students thinking about some key issues in child language acquisition.
+Diary studies and large-scale studies using parent report show that children’s first words tend to fall into a fairly small number of categories: people, food, body parts, clothing, animals, vehicles, toys, household objects, routines, and activities or states [@clark2009first; @fenson1994variability].
+The key insight is that young children talk about what is going on around them: people they see every day, e.g., toys and small household objects they can manipulate or food they can control.
+
+A teacher could illustrate children's early vocabularies by showing a table with columns for the lexical category, examples of words from that category, and children's average age of production.
+To make this an interactive demonstration using childes-db, the teacher could:
+
+1. introduce the question: What kinds of words do children first produce?
+2. allow students to reflect or do a pair-and-share discussion with their neighbor
+3. show the trajectory of a single lexical item while explaining the key parts of the visualization (see Panel A of Fig. XX)
+4. elicit hypotheses from students about the kinds of words that children are likely to produce 
+5. make real-time queries to the interactive visualization tool to add students' suggestions and talk through the updated plots 
+6. finish by entering a pre-selected set of words that communicate the critical takeaway (e.g., see Panel C of Fig. XX)
+
+```{r in-class-demo-flow, fig.pos="h", out.width = "100%", fig.cap="Worked example of using the interactive visualization tools for in-class teaching.
+Panels A-D show how an instructor could dynamically build a plot during a lecture to interactively demonstrate a key concept in language acquisition."}
+library(here)
+library(png)
+png::readPNG(here::here("imgs/in_class_teach_viz.png")) %>% grid::grid.raster()
+```
+
+\noindent
+Another benefit of this workflow is that the instructor could easily reuse it to demonstrate other concepts or questions that come up during the lecture, e.g., order of production for different types of words such as quantifiers vs.
+number words or function vs.
+content words.
+Moreover, future work on childes-db interactive visualization tools could add part of speech tags, phonological information, and n-grams.
+This would allow instructors to use childes-db to demonstrate concepts such as when children start producing certain phonological patterns (e.g., consonant clusters), the debate about the extent of the noun/verb bias in children's early vocabulary, and the development of particular syntactic constructions.
+
+In addition to the interactive data visualizations, instructors can use childes-db to download the raw data used to generate the figures.
+This allows teachers an opportunity to have students further explore and analyze the data they see in the plots.
+Moreover, this functionality provides an example of another way childes-db can facilitate language acquisition courses by making data easily-accessible for tutorials and student research projects.
+In the next section, we walk through one example of how an instructor could use childes-db to create a tutorial that scaffolds students' ability to ask interesting computational questions using child language data.
+
+## Tutorials and assignments
+
+One of the goals for courses on applied natural language processing (NLP) is for students to get hands-on experience using state of the art NLP tools to analyze real-world language data [@hearst2005teaching].
+A primary challenge for the instructor is to decide how much time should be spent teaching the requisite programming skills for accessing and formatting language data, which are typically unstructured.
+One pedagogical strategy is to abstract away these details and avoid having students deal with obtaining data and formatting text.
+This approach shifts students' effort away from data cleaning and towards programming analyses that encourage students to explore language data and test interesting hypotheses.
+
+In the introduction, we outlined the challenge of getting students started with analyzing data from CHILDES.
+These barriers include learning the details of a special file format (CHAT) and a command-line tool (CLAN) built to specifically for the CHAT format.
+This workflow can be daunting to navigate for a beginner.
+Thus, an instructor would have to spend course time teaching students to use CHAT and CLAN, leaving less time for students to try analyses that might increase their interest in child language acquisition.
+Moreover,  while CLAN provides a useful tool for interactive exploration,  it does not connect to a pipeline of modern natural language processing tools (e.g., models of word embeddings @mikolov2013distributed), making it difficult for students to explore the data using new NLP tools.
+
+We hope that childes-db combined with the `childesr` API will provide instructors with a solution to both of these challenges.
+The API gives students an easy-to-learn method for programmatic access to CHILDES data.
+This will reduce the time spent getting data out of CHILDES and allow more time for thinking about interesting questions that can be answered with NLP and child language data.
+Second, by writing our API in the R programming language, students will gain access to a general workflow can be integrated into a pipeline that includes the newest NLP tools.
+
+For example, an instructor could create a programming assignment with the goal of reproducing the key findings in the case studies that we presented earlier -- color words or gender.
+Depending on the students' knowledge of R, the instructor could decide how much of the `childesr` code to provide before asking students to generate their own plots and writeups.
+
+Moreover, the instructor could take advantage of recent advances in tools for teaching literate programming such as R Markdown [@xie2015dynamic; @baumer2014r].
+Writing the assignment in an R Markdown notebook would allow the interleaving of prose with executable chunks of code.
+Instructors could then write prose that is directly related to the course content and the assignment while building in pre-written blocks of executable code that demonstrate key programming concepts for students.
+To push students' understanding, instructors could write incomplete or blank code blocks that students can use to write their own code.
+In short, `childesr` combined with R Markdown provides a powerful teaching tool, allowing students to start analyzing language data quickly and providing instructors with a way to build customized lessons for students at different levels of programming experience.
 

### 2018-04-04 paper.Rmd
SM comments on viz section

--- 2018-04-04 paper.Rmd
@@
 The CHILDES ecosystem uses a special file format (CHAT), which is stored in a plain text format but includes structured annotations grouped into tiers (stored on separate lines) so that information about utterances can be stored with accompanying information such as the phonological or morphological structure of the utterance.
-These files are analyzed via by a command-line tool (CLAN) with a host of specific arguments that allow users to count word frequencies, compute statistics (e.g., mean length of utterance, or MLU).
+These files are analyzed using a command-line tool (CLAN) with a host of specific arguments that allow users to count word frequencies, compute statistics (e.g., mean length of utterance, or MLU).
 While this system is flexible and powerful, mastering the CHAT codes and the CLAN tool can be daunting to navigate for a beginner or student.
@@
 
-childes-db offers a collection of browser-based interactive visualizations that enable data exploration through easy-to-use interfaces that don’t require any programming knowledge.
+childes-db offers a collection of browser-based interactive visualizations that enable data exploration through easy-to-use interfaces that do not require any programming knowledge.
 These tools are intended for casual users who seek to extract useful developmental indices from CHILDES without any technical overhead.
@@
 
+<!-- sm: intro sentence here for the role of frequency in CLA research  --> 
 The Frequency Counts app allows users to enter a word (or words) in a text box and see the frequency (in parts per million) in which this word was uttered across a child's age for a particular subset of the data, whether it is an individual child in the Brown corpus or all Spanish speaking children and their mothers.
+<!-- sm: "whether..." is awk.
+-->
 The ability to view the caregiver’s usage of a term alongside a child’s assists researchers in better understanding the relationship between input and uptake.
@@
 
-There are a number of ways to measure the lexical diversity of an individual's productive speech.
+<!-- sm: intro sentence here for the role of developmental metrics in CLA research and in downstream applications like SLP--> 
+There are a number of ways to measure the lexical diversity of an individual's productive speech. <!-- sm: different intro; MLU isn't a lexical diversity measure -->
 Some of these measures, including MLU-w (mean length of utterance in words), TTR (type-token ratio), MTLD (measure of textual lexical diversity), and voc-d are available in the Derived Measures app, which plots these measures across age for a given subset of data, again specified by collection, corpora, children, and speakers.
+<!-- sm: MLU is not a measure of lexical diversity-->
 Because of the underlying gaps in available CHILDES morphological data, MLU-w is computed and not MLU-m (mean length of utterance in morphemes).
-In the English language, these two measures have been shown to be highly correlated.
-As with the Frequency Counts app, caregiver’s lexical diversity measures can be plotted alongside children's.
+<!-- sm: I think we should go ahead and compute this for transcript with MOR tiers?
+Given that this was one of the first uses, I don't think Brian will like the 'underlying gaps' explanation-->
+In the English language, these two measures have been shown to be highly correlated. <!-- sm: citation needed -->
+As with the Frequency Counts app, caregiver’s lexical diversity measures can be plotted alongside those of children.
 
 ### Population Viewer
-
-The population viewer is a tool that helps users understand the relative sizes of various corpora by displaying the total number of utterances or tokens in the available corpora at particular ages.
+The population viewer is a tool that helps users understand the relative sizes and temporal extents of various corpora by displaying the total number of utterances or tokens in the available corpora at particular ages.
 In many instances the size of a given corpus may inform researchers’ decisions as to what is the best target dataset for analysis.
+<!-- note that the graphs don't currently include a grouping factor, so adding multiple corpora generates one line, so you can't actually compare two corpora in the same visualization -->
 
@@
 @song2015; @norrman2015].
-The `childesr` package abstracts away the details of connecting to the database.
+The `childesr` package abstracts away the details of connecting to <!-- sm: and querying-->  the database.
 Users can take advantage of the tools developed in the popular `dplyr` package [@R-dplyr], which make manipulating large datasets quick and easy.
@@
 To install, simply type: `install.packages("childesr")`.
-After installation, users have access to data loading functions that retrieve a variety of subsets of the data:
+After installation, users have access to data loading functions that retrieve a variety of subsets <!-- sm: entities?
+variety of subsets sounds non-hierarchical-->  of the data:
 
@@
 One common use of the CHILDES corpora is estimating the frequency with which children hear different words.
-These frequency estimates are used both in the development of theory [e.g.
-frequent words are learned earlier; @goodman2008], and in the construction of age-appropriate experimental stimuli.
-One benefit of the `childesdb` interface is that it allows for easy analysis of how the frequencies of words change over development.
+These frequency estimates are used both in the development of theory (e.g.
+frequent words are learned earlier; @goodman2008), and in the construction of age-appropriate experimental stimuli.
+One benefit of the childes-db interface is that it allows for easy analysis of how the frequency of words changes over development.
 Many of our theories in which children learn the structure of language from its statistical properties implicitly assume that these statistics are *stationary*, i.e.
-unchanging over development [e.g., @saffran1996].
-However a number of recent analysis show that the frequencies with which infants encounter both linguistic and visual properties of their environment may change dramatically over development [@fausey2016], and these changing distributions may produce similarly dramatic changes in the ease or difficulty with which these regularities can be learned [@elman1993].
+unchanging over development (e.g., @saffran1996).
+However a number of recent analysis show that the frequencies with which infants encounter both linguistic and visual properties of their environment may change dramattically over development [@fausey2016], and these changing distributions may produce similarly dramatic changes in the ease or difficulty with which these regularities can be learned [@elman1993].
 
-To demonstrate how one might discover such a non-stationary, we take as a case study the frequency with which children hear the color words of English (e.g.
+To demonstrate how one might discover such a non-staionarity, we take as a case study the frequency with which children hear the color words of English (e.g.
 "blue", "green").
-Color words tend to be learned relatively late by children, potentially in part due to the abstractness of the meanings to which they refer [see @wagner2013].
+Color words tend to be learned relatively late by children, potentially in part due to the abstractness of the meanings to which they refer (see @wagner2013).
 However, within the set of color words, the frequency with which these words are heard predicts a significant fraction of the variance in their order of acquisition [@yurovsky2015].
@@
 
-#### Using shiny
-
-To investigate whether the frequency of color words is stationary over development, a user can navigate to the Frequency app, and enter a set of color words into the `Word` selector separated by a comma: here "blue, red, green." Because the question of interest is about the frequency of words in the input (rather than produced by children), the `Speaker` field can be set to reflect this.
-In this example we select "Mother" because, across corpora in CHIDLES, the speaker tagged with the Mother role tends to produce most of the words that children hear.
-Because children learn most of their basic color words by the age of 5, the age range 1-5 years is a reaonable choice for `Ages to include`.
-The results of these selections are shown in Figure \ref{fig:color-shiny}.
-We can also create a hyperlink to store these set of choices so that we can share these results with others, or with ourselves in the future, byby clicking on the `Bookmark` button in the bottom left corner.
-
-From this figure, it seems likely that children hear "blue" more frequently early in development, but the trajectories of the "red" and "green" are less clear.
-We also do not have a good sense of the errors of these measurements, are limited to just a few colors at a time before the plot becomes too busy, and cannot combine frequencies across speakers.
-To perform this analysis in a more compelling and complete way, a user can use the `childesr` interface to get more direct access to the CHILDES corpora.
-
-```{r color-shiny, fig.width = 10, fig.height = 6, fig.cap = "An example of using the Frequency shiny app to explore how children's color input changes over development"}
-knitr::include_graphics("images/color-shiny.png")
-```
-
-#### Using childesr
-
-Figure \ref{fig:color-plot}.
-
-```{r color-plot, fig.height = 5, fig.width = 10, fig.cap ="Color frequency as a function of age.
-Points represent means across transcripts, error bars represent 95% confidence intervals computed by nonparametric bootstrap", cache = TRUE, message = FALSE, warning = FALSE}
+```{r}
 colors <- c("black", "white", "red", "green", "yellow", "blue", "brown", 

### 2018-04-04 paper.Rmd

--- 2018-04-04 paper.Rmd
@@
 frequent words are learned earlier; @goodman2008), and in the construction of age-appropriate experimental stimuli.
-One benefit of the childesdb interface is that it allows for easy analysis of how the frequency of words changes over development.
+One benefit of the childes-db interface is that it allows for easy analysis of how the frequency of words changes over development.
 Many of our theories in which children learn the structure of language from its statistical properties implicitly assume that these statistics are *stationary*, i.e.
@@
 
-To demonstrate how one might discover such a non-staionarity, we take as a case study the frequency with which children hear the color words of english (e.g.
+To demonstrate how one might discover such a non-staionarity, we take as a case study the frequency with which children hear the color words of English (e.g.
 "blue", "green").

### 2018-04-04 paper.Rmd

@@
 One common use of the CHILDES corpora is estimating the frequency with which children hear different words.
-These frequency estimates are used both in the development of theory (e.g.
-frequent words are learned earlier; @goodman2008), and in the construction of age-appropriate experimental stimuli.
-One benefit of the childes-db interface is that it allows for easy analysis of how the frequency of words changes over development.
+These frequency estimates are used both in the development of theory [e.g.
+frequent words are learned earlier; @goodman2008], and in the construction of age-appropriate experimental stimuli.
+One benefit of the `childesdb` interface is that it allows for easy analysis of how the frequencies of words change over development.
 Many of our theories in which children learn the structure of language from its statistical properties implicitly assume that these statistics are *stationary*, i.e.
-unchanging over development (e.g., @saffran1996).
-However a number of recent analysis show that the frequencies with which infants encounter both linguistic and visual properties of their environment may change dramattically over development [@fausey2016], and these changing distributions may produce similarly dramatic changes in the ease or difficulty with which these regularities can be learned [@elman1993].
+unchanging over development [e.g., @saffran1996].
+However a number of recent analysis show that the frequencies with which infants encounter both linguistic and visual properties of their environment may change dramatically over development [@fausey2016], and these changing distributions may produce similarly dramatic changes in the ease or difficulty with which these regularities can be learned [@elman1993].
 
-To demonstrate how one might discover such a non-staionarity, we take as a case study the frequency with which children hear the color words of English (e.g.
+To demonstrate how one might discover such a non-stationary, we take as a case study the frequency with which children hear the color words of English (e.g.
 "blue", "green").
-Color words tend to be learned relatively late by children, potentially in part due to the abstractness of the meanings to which they refer (see @wagner2013).
+Color words tend to be learned relatively late by children, potentially in part due to the abstractness of the meanings to which they refer [see @wagner2013].
 However, within the set of color words, the frequency with which these words are heard predicts a significant fraction of the variance in their order of acquisition [@yurovsky2015].

### 2018-04-04 paper.Rmd
ref typos

--- 2018-04-04 paper.Rmd
@@
 library(koRpus)
+library(here)
 knitr::opts_chunk$set(cache = TRUE, echo = FALSE)
@@
 The foundational resource for studies of language acquisition is recordings and transcripts of children.
-Starting with Roger Brown's [-@brown1973] foundational work on Adam, Eve, and Sarah, longitudinal recordings of children's talk (and the talk they hear from their caregivers) have formed the basis for this field.
+Starting with Roger Brown's [-@brown1973] work on Adam, Eve, and Sarah, longitudinal recordings of children's talk (and the talk they hear from their caregivers) have formed the basis for this field.
 But creating individual transcripts is extremely hard work.
@@
 CLAN software is an excellent tool for interactive exploration, but -- as a free-standing application -- can be tricky to build into a processing pipeline written in a modern analytic platform such as python or R.
-Thus, researchers who would like to ingest the entire corpus (or some large subset) into a computational analysis typically write their own parsers of the CHAT format to extract the subset of the data they would like to use [e.g., @redington1998,@yang2013,@meylan2017].
+Thus, researchers who would like to ingest the entire corpus (or some large subset) into a computational analysis typically write their own parsers of the CHAT format to extract the subset of the data they would like to use [e.g., @redington1998;@yang2013;@meylan2017].
 
@@
 Further, these scripta are rarely shared, leading to much greater difficulty in reproducing the exact numerical results from papers using CHILDES [see e.g., @meylan2017].
-This lack of computational reproducibility is widely recognized as problematic because it does not allow researchers to verify or build on published researcher [@donoho2010,@stodden2016].
+This lack of computational reproducibility is widely recognized as problematic because it does not allow researchers to verify or build on published researcher [@donoho2010;@stodden2016].
 Finally, the CHILDES corpus itself is a moving target;  computational work using the entire corpus at one time point may include a different set of data than subsequent work (due to new contributions or revisions).
@@
 Many of our theories in which children learn the structure of language from its statistical properties implicitly assume that these statistics are *stationary*, i.e.
-unchanging over development (e.g., @saffran1996).
+unchanging over development [e.g., @saffran1996].
 However a number of recent analysis show that the frequencies with which infants encounter both linguistic and visual properties of their environment may change dramattically over development [@fausey2016], and these changing distributions may produce similarly dramatic changes in the ease or difficulty with which these regularities can be learned [@elman1993].
@@
 "blue", "green").
-Color words tend to be learned relatively late by children, potentially in part due to the abstractness of the meanings to which they refer (see @wagner2013).
+Color words tend to be learned relatively late by children, potentially in part due to the abstractness of the meanings to which they refer [see @wagner2013].
 However, within the set of color words, the frequency with which these words are heard predicts a significant fraction of the variance in their order of acquisition [@yurovsky2015].

### 2018-04-04 paper.Rmd
closed #7

@@
 We introduce childes-db, a tool that improves data accessibility and usability by offering novel interfaces to CHILDES, including browsable web applications and an R application programming interface.
-Along with a versioned infrastructure that facilitates reproduction of past analyses, these interfaces lower barriers to analyzing naturalistic parent-child language, allowing for a wider range of researchers in language and cognitive development to easily leverage “data from beyond the lab” in their work.
+Along with a versioned infrastructure that facilitates reproduction of past analyses, these interfaces lower barriers to analyzing naturalistic parent-child language, allowing for a wider range of researchers in language and cognitive development to easily leverage CHILDES in their work.
 
@@
 In addition, this process introduces a major source of errors and inconsistencies in data handling due to difficulties dealing with the many special cases in the CHAT standard.
-Further, these scripta are rarely shared, leading to much greater difficulty in reproducing the exact numerical results from papers using CHILDES [see e.g., @meylan2017].
+Further, these scripts are rarely shared, leading to much greater difficulty in reproducing the exact numerical results from papers using CHILDES [see e.g., @meylan2017].
 This lack of computational reproducibility is widely recognized as problematic because it does not allow researchers to verify or build on published researcher [@donoho2010;@stodden2016].

### 2018-04-04 paper.Rmd
updates to outline during call

@@
 Now, advances in open-source software allow for novel approaches in working with this dataset.
-We introduce childes-db, a tool that improves data accessibility and usability by offering novel interfaces to CHILDES, including browsable web applications and an R application programming interface.
+We introduce childes-db, a database-formatted mirror of CHILDES that improves data accessibility and usability by offering novel interfaces, including browsable web applications and an R application programming interface (API).
 Along with a versioned infrastructure that facilitates reproduction of past analyses, these interfaces lower barriers to analyzing naturalistic parent-child language, allowing for a wider range of researchers in language and cognitive development to easily leverage CHILDES in their work.
@@
 Yet there are some challenges working with CHILDES, both for students and for advanced users.
-For beginners with a lower level of technical experience, mastering the system can be tricky.
 The CHILDES ecosystem uses a special file format (CHAT), which is stored in a plain text format but includes structured annotations grouped into tiers (stored on separate lines) so that information about utterances can be stored with accompanying information such as the phonological or morphological structure of the utterance.
@@
 
-# The Architecture of childes-db
+# Design and technical approach
 
-We first present the design and technical approach of childes-db, then describe the interactive analysis tools provided by our web-based frontend.
+<!-- Remind readers why they're reading this: understanding the structure helps you use and understand the tools -->
 
-## Design and technical approach
+<!-- We first present the design and technical approach of childes-db, then describe the interactive analysis tools provided by our web-based frontend.  -->
+
+## Database format
 
@@
 An *utterance* contains one or more words and includes fields such as the utterance type (declarative, question, etc.), total number of tokens, and the total number one morphemes.
-[[ How are we handling the long tail of tiers?]].
+
 A *transcription* consists of one or more utterances and includes the date collected, the name of the target child and age in days (if defined), and the filename from CHILDES.
@@
 
-### Technical Implementation
+## Technical implementation
+
+<!-- [[mention using XML childes representation]] -->
 
@@
 The childes-db project provides hosted read-only databases for direct access and for `childesr` (described below) as well as compressed .sql exports for local installation.
-While the former is appropriate for casual users, local installation can provide performance gains (allowing a user to access the database on their local machine or on their network) as well as write support in case users wish to store derived information in the same database.
+While the former is appropriate for most users, local installation can provide performance gains (allowing a user to access the database on their local machine or on their network) as well as allowing users to store derived information in the same database.
 
-[[Details of extraction]]
-[[Details of testing: Why trust our parse over your in-house library?]]
+<!-- [[Details of extraction]] -->
+<!-- [[Details of testing: Why trust our parse over your in-house library? CLAN TESTING DETAILS - also mention versioning as a reason to trust]] -->
 
 ### Versioning
+
+<!-- Continue to make the case that it's important to version! childes changes without warning, corpora get added, etc. -->
 
@@
 
+### Current annotation coverage
+
+<!-- [[ How are we handling the long tail of tiers? first priority is lexical statistics & morphology, but there's other stuff too]].  -->
+
+<!-- no media links currently, no timing information -->
+
+<!-- this stuff could be added at some point without changing format -->
+
+# Interfaces for Accessing childes-db
+
+We first discuss the childes-db web apps and then introduce the `childesr` package.
+
 ## Interactive Analysis Tools
+
+<!-- need a good shorthand for the interactive analysis tools: web apps -->
 
@@
 ### Population Viewer
+
 The population viewer is a tool that helps users understand the relative sizes and temporal extents of various corpora by displaying the total number of utterances or tokens in the available corpora at particular ages.
@@
 
-# The `childesr` Package
+## The `childesr` Package
 
@@
 We describe the commands that the package provides and then give several worked examples of using the package for analyses.
-
-## Package Details
 
@@
 
-## Worked Examples
+# Using childes-db: Worked Examples
+
+In this section we give a number of examples of how childes-db can be used in both research and teaching, using both the web apps and the R API.
+
+## Research applications
 
@@
 
-#### Using shiny
+#### Using web apps
 
@@
 
-# Teaching with childes-db
+## Teaching with childes-db
 
-## In class demonstrations
+### In class demonstrations
 
@@
 
-## Tutorials and assignments
+### Tutorials and assignments
 
@@
 
-reiterate importance
+<!-- reiterate importance -->
+
+<!-- adding more tiers is a goal for the future -->
 
 # Acknowledgements
-- Brian MacWhinney for CHILDES
-- Melissa Kline for ClanToR
 
-\newpage
+Thanks to Brian MacWhinney for advice and guidance, and to Melissa Kline for her work on ClanToR, which formed a starting point for our work.
+This work supported by a Jacobs Advanced Research Fellowship to MCF.
 

### 2018-04-04 paper.Rmd
further updates from call

@@
     corresponding : yes    # Define only one corresponding author
-    address       : "Postal address"
+    address       : "Department of Psychology, 450 Serra Mall, Stanford, CA 94305"
     email         : "sanchez7@stanford.edu"
@@
 author_note: |
-  Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.
-
-  This work supported by a Jacobs Foundation Fellowship to MCF.
+  Thanks to Brian MacWhinney for advice and guidance, and to Melissa Kline for her work on ClanToR, which formed a starting point for our work.
+This work supported by a Jacobs Advanced Research Fellowship to MCF.
 
@@
 
-figsintext        : no
+figsintext        : yes
 figurelist        : no
@@
 
-\noindent
 Another benefit of this workflow is that the instructor could easily reuse it to demonstrate other concepts or questions that come up during the lecture, e.g., order of production for different types of words such as quantifiers vs.
@@
 Thus, an instructor would have to spend course time teaching students to use CHAT and CLAN, leaving less time for students to try analyses that might increase their interest in child language acquisition.
-Moreover,  while CLAN provides a useful tool for interactive exploration,  it does not connect to a pipeline of modern natural language processing tools (e.g., models of word embeddings @mikolov2013distributed), making it difficult for students to explore the data using new NLP tools.
+Moreover,  while CLAN provides a useful tool for interactive exploration,  it does not connect to a pipeline of modern natural language processing tools [e.g., models of word embeddings; @mikolov2013distributed], making it difficult for students to explore the data using new NLP tools.
 
@@
 
-# Acknowledgements
-
-Thanks to Brian MacWhinney for advice and guidance, and to Melissa Kline for her work on ClanToR, which formed a starting point for our work.
-This work supported by a Jacobs Advanced Research Fellowship to MCF.
-
 # References

### 2018-04-06 paper.Rmd
typos and other edits

@@
   Thanks to Brian MacWhinney for advice and guidance, and to Melissa Kline for her work on ClanToR, which formed a starting point for our work.
-This work supported by a Jacobs Advanced Research Fellowship to MCF.
+This work is supported by a Jacobs Advanced Research Fellowship to MCF.
 
@@
 Established in 1984 to address this issue, the Child Language Data Exchange System (CHILDES) aims to make transcripts and recording of children's language available for researchers via a free resource [@macwhinney2000].
-CHILDES archives tens of thousands of transcripts of children’s speech, making it an critical resource for characterizing children’s early productive language use as well as their language environment.
+CHILDES archives tens of thousands of transcripts of children’s speech, making it a critical resource for characterizing children’s early productive language use as well as their language environment.
 As the first major effort to consolidate and share transcripts of child language, CHILDES has been a pioneer in the move to curate and disseminate large-scale behavioral datasets publicly.
@@
 On the opposite end of the spectrum, for data-oriented researchers who are interested in doing large-scale analyses of CHILDES, the current tools are also not ideal.
-CLAN software is an excellent tool for interactive exploration, but -- as a free-standing application -- can be tricky to build into a processing pipeline written in a modern analytic platform such as python or R.
+CLAN software is an excellent tool for interactive exploration, but -- as a free-standing application -- can be tricky to build into a processing pipeline written in a modern analytic platform such as Python or R.
 Thus, researchers who would like to ingest the entire corpus (or some large subset) into a computational analysis typically write their own parsers of the CHAT format to extract the subset of the data they would like to use [e.g., @redington1998;@yang2013;@meylan2017].
@@
 This practice is problematic for a number of reasons.
-Effort wasted in rewriting the same parser again and again.
+Effort is wasted in rewriting the same parser again and again.
 In addition, this process introduces a major source of errors and inconsistencies in data handling due to difficulties dealing with the many special cases in the CHAT standard.
 Further, these scripts are rarely shared, leading to much greater difficulty in reproducing the exact numerical results from papers using CHILDES [see e.g., @meylan2017].
-This lack of computational reproducibility is widely recognized as problematic because it does not allow researchers to verify or build on published researcher [@donoho2010;@stodden2016].
+This lack of computational reproducibility is widely recognized as problematic because it does not allow researchers to verify or build on published research [@donoho2010;@stodden2016].
 Finally, the CHILDES corpus itself is a moving target;  computational work using the entire corpus at one time point may include a different set of data than subsequent work (due to new contributions or revisions).
@@
 In the current manuscript, we describe a system for extending the functionality of CHILDES to address these issues.
-Our systemn, childes-db, is a database-formatted mirror of CHILDES that allows programmatic access through an application programming interface (API).
+Our system, childes-db, is a database-formatted mirror of CHILDES that allows programmatic access through an application programming interface (API).
 This infrastructure allows the creation of web applications for browsing and visualizing the data easily, facilitating classroom use of the dataset.
@@
 Users of data analysis languages like R or Julia, libraries like Pandas, or those familiar with Structured Query Language (SQL) will be familiar with operations on tables such as filtering (subsetting), sorting, aggregation (grouping), and joins (merges).
-We expect most users to interact with the database through the `childesr` API for the `R` language described below, we begin with a top-level description and motivation for the design of the database schema, details of the database’s current technical implementation, and our versioning scheme.
+Although we expect most users to interact with the database through the `childesr` API for the `R` language described below, we begin with a top-level description and motivation for the design of the database schema, details of the database’s current technical implementation, and our versioning scheme.
 
 At its core, childes-db is a database consisting of a set of linked tabular data stores where rows correspond to records.
-The smallest unit of abstraction tracked by the database is a token, treated here as the citation orthographic form of a word.
+<!-- what did you mean here?
+don't rows = records--> The smallest unit of abstraction tracked by the database is a token, treated here as the citation orthographic form of a word.
 Using the standardized written form of the word facilitates the computation of lexical frequency statistics, for comparison or aggregation across children or time periods.
@@
 Every entity includes attributes that link it to all higher-order collections, e.g., an utterance lists the transcript, corpus, and collection to which it belongs.
-An *utterance* contains one or more words and includes fields such as the utterance type (declarative, question, etc.), total number of tokens, and the total number one morphemes.
+An *utterance* contains one or more words and includes fields such as the utterance type (declarative, question, etc.), total number of tokens, and the total number of morphemes (if the morphological structure is available in the original CHAT).
 
-A *transcription* consists of one or more utterances and includes the date collected, the name of the target child and age in days (if defined), and the filename from CHILDES.
+A *transcription* consists of one or more utterances and includes the date collected, the name of the target child, and age in days (if defined), and the filename from CHILDES.
 A *corpus* consists of one or more transcripts, corresponding to well-known collections like the Brown or Providence corpus.
 Finally, a *collection* is a superordinate collection of corpora generally corresponding to a geographic region, following the convention in CHILDES.
-Because every entity can be linked to a top-level collection, childes-db places content from all languages in the same table for a given level of abstraction.
+Because every entity can be linked to a top-level collection, childes-db places content from all languages in the same table for a given level of abstraction <!-- a little confused by this last sentence -->.
 
@@
 
-To reflect changes in the CHILDES database as new corpora are added or previous annotations revised, we produce a new versioned parse of the complete database every [[n]] months.
+<!-- Potentially put some R code to show easy it is to access previous versions? -->
+
+To reflect changes in the CHILDES database as new corpora are added or previous annotations revised, we produce a new versioned parse of the complete database every 6 months.
 By default, users connect to the most recent version.
-To support reproduction of results with previous versions of the database, we provide hosted versions of the last [[m]] versions so that researchers can refer to specific releases.
-Compressed .sql files are available for download for [[r]] years.
+To support reproduction of results with previous versions of the database, we provide access to hosted versions through our `childesr` API so that researchers can refer to specific releases.
+Compressed .sql files are available for download.
 
@@
 frequent words are learned earlier; @goodman2008], and in the construction of age-appropriate experimental stimuli.
-One benefit of the `childesdb` interface is that it allows for easy analysis of how the frequencies of words change over development.
+One benefit of the childes-db interface is that it allows for easy analysis of how the frequencies of words change over development.
 Many of our theories in which children learn the structure of language from its statistical properties implicitly assume that these statistics are *stationary*, i.e.
 unchanging over development [e.g., @saffran1996].
-However a number of recent analysis show that the frequencies with which infants encounter both linguistic and visual properties of their environment may change dramatically over development [@fausey2016], and these changing distributions may produce similarly dramatic changes in the ease or difficulty with which these regularities can be learned [@elman1993].
+However a number of recent analyses show that the frequencies with which infants encounter both linguistic and visual properties of their environment may change dramatically over development [@fausey2016], and these changing distributions may produce similarly dramatic changes in the ease or difficulty with which these regularities can be learned [@elman1993].
 
@@
 However, within the set of color words, the frequency with which these words are heard predicts a significant fraction of the variance in their order of acquisition [@yurovsky2015].
-But are these frequencies stationary--e.g.
+But are these frequencies stationary -- e.g.
 do children hear "blue" as often at 12 months as they do at 24 months?
-We answer this question in two ways--first using the childesb shiny apps, and then using the `childesr` package.
+We answer this question in two ways -- first using the childesb shiny apps, and then using the `childesr` package.
 
@@
 To investigate whether the frequency of color words is stationary over development, a user can navigate to the Frequency app, and enter a set of color words into the `Word` selector separated by a comma: here "blue, red, green." Because the question of interest is about the frequency of words in the input (rather than produced by children), the `Speaker` field can be set to reflect this.
-In this example we select "Mother" because, across corpora in CHIDLES, the speaker tagged with the Mother role tends to produce most of the words that children hear.
-Because children learn most of their basic color words by the age of 5, the age range 1-5 years is a reaonable choice for `Ages to include`.
+In this example we select "Mother" because, across corpora in CHILDES, the speaker tagged with the Mother role tends to produce most of the words that children hear.
+Because children learn most of their basic color words by the age of 5, the age range of 1-5 years is a reaonable choice for `Ages to include`.
 The results of these selections are shown in Figure \ref{fig:color-shiny}.
-We can also create a hyperlink to store these set of choices so that we can share these results with others, or with ourselves in the future, byby clicking on the `Bookmark` button in the bottom left corner.
+We can also create a hyperlink to store these set of choices so that we can share these results with others, or with ourselves in the future, by clicking on the `Bookmark` button in the bottom left corner.
 
@@
   group_by(target_child_id, target_child_age, target_child_sex, language) %>%
-  summarise(ttr = sum(num_types) / sum(num_tokens))
+  summarise(measure = mean(mtld))
 
-ggplot(data, aes(x = target_child_age, y = ttr, color = target_child_sex)) +
+ggplot(data, aes(x = target_child_age, y = measure, color = target_child_sex)) +
   geom_smooth()

### 2018-04-10 paper.Rmd
gender plot changes

@@
 ### Gender
+```{r}
+stats <- get_speaker_statistics(role = "Target_Child")
+```
 
 ```{r warning=FALSE}
-stats <- get_speaker_statistics(role = "Target_Child")
+lang_threshold <- 575
+age_max <- 4
+
+lang_counts <- stats %>% 
+  filter(target_child_age < age_max * 365) %>%
+  group_by(language) %>%
+
+  summarise(n = n())
+
+included_languages <- filter(lang_counts, n > lang_threshold)$language
 
 data <- stats %>%
-  filter(!is.na(target_child_sex), language == "eng") %>%
+  filter(!is.na(target_child_sex), 
+         language %in% included_languages) %>%
   group_by(target_child_id, target_child_age, target_child_sex, language) %>%
-  summarise(measure = mean(mtld))
+  filter(mtld != 0) %>%
+  summarise(measure = mean(mtld)) %>%
+  ungroup() %>%
+  mutate(age_years = target_child_age / 365, 
+         target_child_sex = factor(target_child_sex, levels = c("male","female"))) %>%
+  filter(age_years < age_max)
 
-ggplot(data, aes(x = target_child_age, y = measure, color = target_child_sex)) +
-  geom_smooth()
+mod <- lme4::lmer(measure ~ target_child_sex * poly(age_years, 2) - 1 + 
+                    (1 | target_child_id) +
+                    (target_child_sex + age_years | language), 
+  data = data)
+
+mod_data <- expand.grid(age_years = seq(0,age_max, .1), 
+                                          target_child_sex = c("male","female"),
+                                          language = included_languages) %>%
+  as_data_frame()
+
+mod_data$measure <- predict(mod, newdata = mod_data, 
+                           re.form = ~(target_child_sex + age_years | language))
+
+ggplot(data, aes(x = age_years, y = measure, 
+                 color = target_child_sex)) +
+  geom_point(alpha = .3) + 
+  geom_line(data = mod_data, 
+            aes(x = age_years, y = measure, color = target_child_sex)) +
+  # geom_smooth(span = 1, method = "loess") + 
+  facet_wrap(~language) + 
+  ggthemes::theme_few() + 
+  ggthemes::scale_color_ptol(name = "Sex") + 
+  xlim(0,age_max) + 
+  ylim(0,15) + 
+  xlab("Age (years)") + 
+  ylab("MTLD") + 
+  theme(legend.position = "bottom")
 ```

### 2018-04-11 paper.Rmd
added versioning details and annotation coverage infromation

@@
 
-<!-- [[mention using XML childes representation]] -->
-
 The above tabular datastore is implemented as a MySQL database.
@@
 <!-- [[Details of extraction]] -->
+<!-- [[mention using XML childes representation]] -->
 <!-- [[Details of testing: Why trust our parse over your in-house library? CLAN TESTING DETAILS - also mention versioning as a reason to trust]] -->
@@
 
+The content of CHILDES changes as additional corpora are added or transcriptions are updated; currently, these changes are not systematically tracked.
+Here we introduce a simple versioning system by proucing a new complete parse of the current state of CHILDES every 6 months.
+Users interact with the most recent version of the database available by default. 
+To support reproduction of results with previous versions of the database, we continue to host recent versions (up to the last 6 versions) through our `childesr` API so that researchers can run analyses against specific versions of the database.
 <!-- Potentially put some R code to show easy it is to access previous versions? -->
+For versions more than three years old, users have the options of downloading compressed .sql files and serving them with a local installation of MySQL server.
+<!-- How easy is it to implement versioning in the Shiny apps?-->
 
-To reflect changes in the CHILDES database as new corpora are added or previous annotations revised, we produce a new versioned parse of the complete database every 6 months.
-By default, users connect to the most recent version.
-To support reproduction of results with previous versions of the database, we provide access to hosted versions through our `childesr` API so that researchers can refer to specific releases.
-Compressed .sql files are available for download.
+### Current Annotation Coverage
 
-### Current annotation coverage
+The current implementation of childes-db emphasizes the computation of lexical statistics, and consequently focuses on reproducing the words, utterances, and speaker information in CHILDES.
+We have yet to parse and make available the following information:
 
-<!-- [[ How are we handling the long tail of tiers? first priority is lexical statistics & morphology, but there's other stuff too]].  -->
+- Sparsely annotated tiers, e.g. phonology (`%pho`) and situation (`%sit`)
+- Media links, including timing information
+- compound word markers, indicated with underscores ("green_eggs_and_ham")
+- Tone direction and stress
+- Filled pauses 
+- Reformulations, word revision, and phrase revision, e.g. <what did you> [//] how can you see it ?
+- paralinguistic material, e.g. [=! cries]
 
-<!-- no media links currently, no timing information -->
-
-<!-- this stuff could be added at some point without changing format -->
+We will prioritize the addition of these additional information sources in response to community feedback;
+extending coverage to the above information sources will not require changes to the existing database schema.
 

### 2018-04-11 paper.Rmd
streamlined teaching section

@@
 library(directlabels)
-library(koRpus)
 library(here)
@@
 
-<!-- [[Details of extraction]] -->
-<!-- [[mention using XML childes representation]] -->
+In order to import the CHILDES corpora into the MySQL schema described above, it must first be accurately parsed and subsequently vetted to ensure the integrity of the data migration.
+Along with being made available in CHAT, the corpora are also downloadable in a format known as eXtensible Markup Language, or XML; this significantly mitigated any problems related to parsing CHAT as it allowed us to leverage the built-in XML parsing libraries available in numerous programming languages such as Python.
+Once the XML data were crawled and parsed into tokens and utterances, they were imported into a MySQL database.
+The meticulously constructed CHILDES XML Schema allowed us to make precise decisions as to what information we found relevant to our database and what was to be discarded or ignored; further information on what data were retained can be found below.
+
+The data imported into childes-db requires extensive data integrity checks to ensure that our parse of the corpora is accurate and preferable over the adhoc parsers developed by many individual researchers.
+In order to evaluate this, we compared unigram counts in our database with those outputted by CLAN, the command-line tool built specifically for analysis of transcripts coded in CHAT.
+<!-- technically this is all we do right now -->
+
 <!-- [[Details of testing: Why trust our parse over your in-house library? CLAN TESTING DETAILS - also mention versioning as a reason to trust]] -->
@@
 - Sparsely annotated tiers, e.g. phonology (`%pho`) and situation (`%sit`)
-- Media links, including timing information
-- compound word markers, indicated with underscores ("green_eggs_and_ham")
+- Media links, including timing information <!-- we have media timestamps -->
+- compound word markers, indicated with underscores ("green_eggs_and_ham") <!-- we have this -->
 - Tone direction and stress
@@
 Teachers of courses on early language acquisition often want to demonstrate interesting changes in children’s early language.
-One teaching method is to present static displays on slides or handouts that show (a) text from parent-child conversations extracted from CHILDES or (b) data visualizations of changes in various metrics related to language production and input (e.g., MLU or Frequency).
-The challenge with static displays or handouts is that they cannot be modified during a lecture and thus rely on the instructor selecting examples that they think will be compelling to their students.
+One method is to present static displays that show text from parent-child conversations extracted from CHILDES or data visualizations of various metrics of production and input (e.g., MLU or Frequency).
+The challenge with static displays is that they cannot be modified during a lecture and thus rely on the instructor selecting examples that will be compelling to students.
 
-Interactive, in-class demonstrations offer a solution and can be a powerful way to explain complex concepts while increasing student engagement with the course materials.
-One goal in creating the browser-based interactive visualization tools in childes-db is to provide instructors with a new tool for illustrating key concepts in children's language acquisition.
-In this section, we walk through one example using childes-db to show the developmental trajectories of different kinds of words that in children's early production.
-We also highlight ways that instructors could make the in-class demonstration interactive by eliciting hypotheses from students, which become operationalized as queries to the interactive visualization tools.
+In-class demonstrations offer a solution and can be a powerful way to explain complex concepts while increasing student engagement with the course materials.
+One goal of creating the web applications in childes-db is to give instructors a new tool for illustrating concepts in children's language acquisition.
 
-### Worked example
-
-What kinds of words do children first produce?
-This question is a good candidate for an interesting "hook" that gets students thinking about some key issues in child language acquisition.
+For example, a teacher could want to show the types of words that children first produce.
 Diary studies and large-scale studies using parent report show that children’s first words tend to fall into a fairly small number of categories: people, food, body parts, clothing, animals, vehicles, toys, household objects, routines, and activities or states [@clark2009first; @fenson1994variability].
@@
 A teacher could illustrate children's early vocabularies by showing a table with columns for the lexical category, examples of words from that category, and children's average age of production.
-To make this an interactive demonstration using childes-db, the teacher could:
+To make this interactive using the web apps, the teacher could:
 
-1. introduce the question: What kinds of words do children first produce?
+1. introduce the research question
 2. allow students to reflect or do a pair-and-share discussion with their neighbor
-3. show the trajectory of a single lexical item while explaining the key parts of the visualization (see Panel A of Fig. XX)
+3. show the trajectory of a single lexical item while explaining the key parts of the visualization (see Panel A of Figure \ref{fig:in-class-demo-plot})
 4. elicit hypotheses from students about the kinds of words that children are likely to produce 
-5. make real-time queries to the interactive visualization tool to add students' suggestions and talk through the updated plots 
-6. finish by entering a pre-selected set of words that communicate the critical takeaway (e.g., see Panel C of Fig. XX)
+5. make real-time queries to the web applications to add students' suggestions and talk through the updated plots (Panels B and C of Figure \ref{fig:in-class-demo-plot})
+6. finish by entering a pre-selected set of words that communicate the critical takeaway (Panel D of Figure \ref{fig:in-class-demo-plot})
 
-```{r in-class-demo-flow, fig.pos="h", out.width = "100%", fig.cap="Worked example of using the interactive visualization tools for in-class teaching.
+```{r in-class-demo-plot, fig.pos="h", out.width = "100%", fig.cap="Worked example of using the web applications for in-class teaching.
 Panels A-D show how an instructor could dynamically build a plot during a lecture to interactively demonstrate a key concept in language acquisition."}
@@
 
-Another benefit of this workflow is that the instructor could easily reuse it to demonstrate other concepts or questions that come up during the lecture, e.g., order of production for different types of words such as quantifiers vs.
-number words or function vs.
-content words.
-Moreover, future work on childes-db interactive visualization tools could add part of speech tags, phonological information, and n-grams.
-This would allow instructors to use childes-db to demonstrate concepts such as when children start producing certain phonological patterns (e.g., consonant clusters), the debate about the extent of the noun/verb bias in children's early vocabulary, and the development of particular syntactic constructions.
-
-In addition to the interactive data visualizations, instructors can use childes-db to download the raw data used to generate the figures.
+In addition to the web applications, instructors can use childes-db to download the raw data used to generate the figures.
 This allows teachers an opportunity to have students further explore and analyze the data they see in the plots.
 Moreover, this functionality provides an example of another way childes-db can facilitate language acquisition courses by making data easily-accessible for tutorials and student research projects.
-In the next section, we walk through one example of how an instructor could use childes-db to create a tutorial that scaffolds students' ability to ask interesting computational questions using child language data.
 
@@
 
-In the introduction, we outlined the challenge of getting students started with analyzing data from CHILDES.
-These barriers include learning the details of a special file format (CHAT) and a command-line tool (CLAN) built to specifically for the CHAT format.
-This workflow can be daunting to navigate for a beginner.
-Thus, an instructor would have to spend course time teaching students to use CHAT and CLAN, leaving less time for students to try analyses that might increase their interest in child language acquisition.
-Moreover,  while CLAN provides a useful tool for interactive exploration,  it does not connect to a pipeline of modern natural language processing tools [e.g., models of word embeddings; @mikolov2013distributed], making it difficult for students to explore the data using new NLP tools.
-
-We hope that childes-db combined with the `childesr` API will provide instructors with a solution to both of these challenges.
-The API gives students an easy-to-learn method for programmatic access to CHILDES data.
+One goal of childes-db and `childesr` API is to provide instructors with an easy-to-learn method for giving students programmatic access to CHILDES data.
 This will reduce the time spent getting data out of CHILDES and allow more time for thinking about interesting questions that can be answered with NLP and child language data.
@@
 
-For example, an instructor could create a programming assignment with the goal of reproducing the key findings in the case studies that we presented earlier -- color words or gender.
-Depending on the students' knowledge of R, the instructor could decide how much of the `childesr` code to provide before asking students to generate their own plots and writeups.
-
-Moreover, the instructor could take advantage of recent advances in tools for teaching literate programming such as R Markdown [@xie2015dynamic; @baumer2014r].
-Writing the assignment in an R Markdown notebook would allow the interleaving of prose with executable chunks of code.
-Instructors could then write prose that is directly related to the course content and the assignment while building in pre-written blocks of executable code that demonstrate key programming concepts for students.
-To push students' understanding, instructors could write incomplete or blank code blocks that students can use to write their own code.
-In short, `childesr` combined with R Markdown provides a powerful teaching tool, allowing students to start analyzing language data quickly and providing instructors with a way to build customized lessons for students at different levels of programming experience.
+For example, an instructor could create a programming assignment with the goal of reproducing the key findings in the case studies presented above -- color words or gender.
+Depending on the students' knowledge of R, the instructor could decide how much of the `childesr` starter code to provide before asking students to generate their own plots and writeups.
+Moreover, the instructor could take advantage of recent advances in tools for teaching literate programming such as R Markdown [@xie2015dynamic; @baumer2014r], building in pre-written blocks of executable code that demonstrate key programming concepts for students and including incomplete or blank code blocks that students can use to write their own code.
 

### 2018-04-11 paper.Rmd

--- 2018-04-11 paper.Rmd
@@
 library(directlabels)
+library(koRpus)
+library(png)
 library(here)
@@
 
-1. introduce the research question
+1. introduce the research question (e.g., What are the types of words that children first produce?)
 2. allow students to reflect or do a pair-and-share discussion with their neighbor
-3. show the trajectory of a single lexical item while explaining the key parts of the visualization (see Panel A of Figure \ref{fig:in-class-demo-plot})
+3. show the trajectory of a single lexical item while explaining key parts of the visualization (see Panel A of Figure \ref{fig:in-class-demo-plot})
 4. elicit hypotheses from students about the kinds of words that children are likely to produce 
-5. make real-time queries to the web applications to add students' suggestions and talk through the updated plots (Panels B and C of Figure \ref{fig:in-class-demo-plot})
-6. finish by entering a pre-selected set of words that communicate the critical takeaway (Panel D of Figure \ref{fig:in-class-demo-plot})
+5. make real-time queries to the web application to add students' suggestions and talk through the updated plots (Panels B and C of Figure \ref{fig:in-class-demo-plot})
+6. finish by entering a pre-selected set of words that communicate the important takeaway point (Panel D of Figure \ref{fig:in-class-demo-plot})
 
-```{r in-class-demo-plot, fig.pos="h", out.width = "100%", fig.cap="Worked example of using the web applications for in-class teaching.
-Panels A-D show how an instructor could dynamically build a plot during a lecture to interactively demonstrate a key concept in language acquisition."}
-library(here)
-library(png)
+```{r in-class-demo-plot, fig.pos="h", out.width = "95%", fig.cap="Worked example of using the web applications for in-class teaching.
+Panels A-D show how an instructor could dynamically build a plot during a lecture to demonstrate a key concept in language acquisition."}
+
 png::readPNG(here::here("imgs/in_class_teach_viz.png")) %>% grid::grid.raster()
@@
 In addition to the web applications, instructors can use childes-db to download the raw data used to generate the figures.
-This allows teachers an opportunity to have students further explore and analyze the data they see in the plots.
-Moreover, this functionality provides an example of another way childes-db can facilitate language acquisition courses by making data easily-accessible for tutorials and student research projects.
+This gives students the opportunity to further explore and analyze the data they see in the plots.
+Moreover, this functionality provides an example of a second way that childes-db can facilitate language acquisition courses by making data easily-accessible for tutorials and student research projects.
 
-### Tutorials and assignments
+### Tutorials and programming assignments
 
-One of the goals for courses on applied natural language processing (NLP) is for students to get hands-on experience using state of the art NLP tools to analyze real-world language data [@hearst2005teaching].
+One of the goals for courses on applied natural language processing (NLP) is for students to get hands-on experience using NLP tools to analyze real-world language data.
 A primary challenge for the instructor is to decide how much time should be spent teaching the requisite programming skills for accessing and formatting language data, which are typically unstructured.
 One pedagogical strategy is to abstract away these details and avoid having students deal with obtaining data and formatting text.
-This approach shifts students' effort away from data cleaning and towards programming analyses that encourage students to explore language data and test interesting hypotheses.
+This approach shifts students' effort away from data cleaning and towards programming analyses that encourage the exploration and testing of interesting hypotheses.
 
-One goal of childes-db and `childesr` API is to provide instructors with an easy-to-learn method for giving students programmatic access to CHILDES data.
-This will reduce the time spent getting data out of CHILDES and allow more time for thinking about interesting questions that can be answered with NLP and child language data.
-Second, by writing our API in the R programming language, students will gain access to a general workflow can be integrated into a pipeline that includes the newest NLP tools.
+One goal of childes-db and `childesr` API is to provide instructors with an easy-to-learn method for giving students programmatic access to child language data.
+The aim is to reduce time spent getting data out of CHILDES and allow more course time for thinking about interesting questions that could be answered with NLP and child language data.
+Moreover, by writing our API in the R programming language, students will gain access to a general workflow can be integrated into an analysis pipeline that includes state of the art NLP tools.
 
@@
 Depending on the students' knowledge of R, the instructor could decide how much of the `childesr` starter code to provide before asking students to generate their own plots and writeups.
-Moreover, the instructor could take advantage of recent advances in tools for teaching literate programming such as R Markdown [@xie2015dynamic; @baumer2014r], building in pre-written blocks of executable code that demonstrate key programming concepts for students and including incomplete or blank code blocks that students can use to write their own code.
+Moreover, the instructor could take advantage of recent advances in tools for teaching literate programming such as R Markdown [@xie2015dynamic], building in pre-written blocks of executable code that demonstrate key programming concepts for students and including incomplete or blank code blocks that students can use to write their own code.
 

### 2018-04-11 paper.Rmd
color example

@@
 do children hear "blue" as often at 12 months as they do at 24 months?
-We answer this question in two ways -- first using the childesb shiny apps, and then using the `childesr` package.
+We answer this question in two ways -- first using the web apps, and then using the `childesr` package.
 
@@
 In this example we select "Mother" because, across corpora in CHILDES, the speaker tagged with the Mother role tends to produce most of the words that children hear.
-Because children learn most of their basic color words by the age of 5, the age range of 1-5 years is a reaonable choice for `Ages to include`.
+Because children learn most of their basic color words by the age of 5, the age range of 1-5 years is a reasonable choice for `Ages to include`.
 The results of these selections are shown in Figure \ref{fig:color-shiny}.
@@
 
-```{r color-shiny, fig.width = 10, fig.height = 6, fig.cap = "An example of using the Frequency shiny app to explore how children's color input changes over development"}
+```{r color-shiny, out.width = "500px", fig.cap = "An example of using the Frequency shiny app to explore how children's color input changes over development"}
 knitr::include_graphics("images/color-shiny.png")
@@
 
-Figure \ref{fig:color-plot}.
+We can analyze these learning trajectories using childesr by braking the process into five steps: (1) define our words of interest, (2) find the frequencies with with which children hear these words, (3) Find the proportion of the *total words* children hear that these frequencies account for, (4) Aggregate across transcripts and children to determine the error in our estimates of these proportions, and (5) plot the results.
 
-```{r color-plot, fig.height = 5, fig.width = 10, fig.cap ="Color frequency as a function of age.
-Points represent means across transcripts, error bars represent 95% confidence intervals computed by nonparametric bootstrap", cache = TRUE, message = FALSE, warning = FALSE}
+For this analysis, we will define our words of interest as the basic color words of English (except for gray, which children hear very rarely).
+We store these in the `colors` variable, and then use the `get_types` function from `childesr` to get the type frequency of each of these words in all of the corpora in CHILDES.
+For demonstration, we look only at the types produced by the speakers in each corpus oftagged as Mother and Father.
+We also restrict ourselves to children from 1-5 years old (12 to 60 months), and look only the American English corpora.
+
+```{r define-colors, echo = TRUE}
 colors <- c("black", "white", "red", "green", "yellow", "blue", "brown", 
@@
 
+color_counts <- get_types(collection = "Eng-NA", role = c("Mother", "Father"),
+                          age = c(12,60), type = colors)
+```
+
+```{r graph-colors}
 graph_colors <- data_frame(color = colors,
@@
 
-color_counts <- get_types(collection = "Eng-NA", role = c("Mother", "Father"),
-                          age = c(12,60), type = colors)
+```
 
-parents <- color_counts %>%
-  distinct(collection_id, corpus_id, transcript_id, speaker_id) %>%
+To normalize correctly (i.e.
+to ask what proprotion of the input children hear consists of these color words), we need to know how many total words these children hear from their parents in these transcripts.
+To do this, we use the `get_speaker_statistics` function, which will return a total number of tokens (`num_tokens`) for each of these speakers.
+
+```{r get-token-frequencies, echo = TRUE}
+# Get the ids corresponding to all of the speakers we are interested in
+parent_ids <- color_counts %>%
+  distinct(collection_id, corpus_id, transcript_id, speaker_id)
+
+# Find in the total number of tokens produced by these speakers
+parents <- parent_ids %>%
   left_join(get_speaker_statistics(collection = "Eng-NA")) %>%
   select(collection_id, corpus_id, transcript_id, speaker_id, num_tokens)
+```
 
+We now join these two pieces of information together--how many times each speaker produced each color word, and how many total words they produced.
+We then group the data into 6-month age bins, and compute the proportion of tokens that comprise each color for each child in each 6-month bin.
+For comparability with the web app analysis, we convert the proportions to parts per million words
+```{r join-colors-and-tokens, echo = TRUE}
 count_estimates <- color_counts %>%
   left_join(parents) %>%
-  mutate(age = target_child_age / 30.5,
-         age_bin = as.integer(floor(age / 6) * 6),
+  mutate(age_months = target_child_age / 30.5,
+         age_bin = as.integer(floor(age_months / 6) * 6),
          color = tolower(gloss)) %>%
-  group_by(age_bin, color, transcript_id) %>%
-  summarise(count = sum(count), tokens = sum(num_tokens)) %>%
-  mutate(prop = count/tokens * 1000000) %>%
-  tidyboot::tidyboot_mean(prop) %>%
+  group_by(age_bin, color, target_child_id, transcript_id) %>%
+  summarise(count = sum(count), num_tokens = sum(num_tokens)) %>%
+  summarise(count = sum(count), num_tokens = sum(num_tokens)) %>%
+  mutate(parts = count / num_tokens * 1e6)
+```
+
+Finally, we use non-parametric bootstrapping to estimate 95% confidence intervals for our estimates of the parts per million words of each color term with the `tidyboot` package.
+
+```{r computer-color-error, echo = TRUE}
+count_estimates_with_error <- count_estimates %>%  
+  tidyboot::tidyboot_mean(parts) %>%
   left_join(graph_colors) %>%
   mutate(color = factor(color, levels = colors))
+```
 
-ggplot(count_estimates, aes(x = age_bin, y = empirical_stat, 
+Figure \ref{fig:color-plot} shows the results of these analyses: Input frequency varies significantly over the 1-5 year range for nearly every color word.
+
+```{r color-plot, fig.height = 5, fig.width = 10, fig.cap ="Color frequency as a function of age.
+Points represent means across transcripts, error bars represent 95% confidence intervals computed by nonparametric bootstrap", cache = TRUE, message = FALSE, warning = FALSE}
+ggplot(count_estimates_with_error, aes(x = age_bin, y = empirical_stat, 
                             ymin = ci_lower, ymax = ci_upper,
@@
 ```
-
-non-stationarity
 

### 2018-04-11 paper.Rmd
worked on technical section

@@
 
-<!-- Remind readers why they're reading this: understanding the structure helps you use and understand the tools -->
-
+<!-- Remind readers why they're reading this: understanding the structure helps you use and understand the tools. SM: Moved this up from database format because it addresses more general issues -->
+Unlike the *parsing* approach in CLAN which entails the sequential processing of strings, childes-db treats CHILDES as a set of linked tables, with records corresponding to intuitive abstractions such as words, utterances, and transcripts.
+Users of data analysis languages like R or Julia, libraries like Pandas, or those familiar with Structured Query Language (SQL) will be familiar with operations on tables such as filtering (subsetting), sorting, aggregation (grouping), and joins (merges).
+Although we expect most users to interact with the database through the `childesr` API for the `R` language described below, we begin with a top-level description and motivation for the design of the database schema, details of the database’s current technical implementation, and our versioning scheme.
 <!-- We first present the design and technical approach of childes-db, then describe the interactive analysis tools provided by our web-based frontend.  -->
@@
 
-Unlike the *parsing* approach in CLAN which entails the sequential processing of strings, childes-db treats CHILDES as a set of linked tables, with records corresponding to intuitive abstractions such as words, utterances, and transcripts.
-Users of data analysis languages like R or Julia, libraries like Pandas, or those familiar with Structured Query Language (SQL) will be familiar with operations on tables such as filtering (subsetting), sorting, aggregation (grouping), and joins (merges).
-Although we expect most users to interact with the database through the `childesr` API for the `R` language described below, we begin with a top-level description and motivation for the design of the database schema, details of the database’s current technical implementation, and our versioning scheme.
-
-At its core, childes-db is a database consisting of a set of linked tabular data stores where rows correspond to records.
-<!-- what did you mean here?
-don't rows = records--> The smallest unit of abstraction tracked by the database is a token, treated here as the citation orthographic form of a word.
+At its core, childes-db is a database consisting of a set of linked tabular data stores where rows correspond to linguistic entities like words, utterances, and sampling units like transcriptions and corpora.
+<!-- what did you mean here? don't rows = records; SM: that's a description of a database in general; we can make a specific assertion about childes-db-->
+The smallest unit of abstraction tracked by the database is a token, treated here as the citation orthographic form of a word.
 Using the standardized written form of the word facilitates the computation of lexical frequency statistics, for comparison or aggregation across children or time periods. 
@@
 Finally, a *collection* is a superordinate collection of corpora generally corresponding to a geographic region, following the convention in CHILDES.
-Because every entity can be linked to a top-level collection, childes-db places content from all languages in the same table for a given level of abstraction <!-- a little confused by this last sentence -->.
+Because every record can be linked to a top-level collection (generally corresponding to a language), each table includes data from all languages represented in CHILDES.
 
 We chose to treat participants separate from the above hierarchy because children can occur across transcripts (e.g., the Brown Corpus).
-A participant is associated with every word and utterance, including a name, role, 3-letter CHILDES identifier (CHI, MOT, FAT, etc.), and the range of ages (or age of corresponding child) for which they are observed.
-For non-child participants, the record contains an identifier for the target child to associate caregivers and children.
+A participant identifier is associated with every word and utterance, including a name, role, 3-letter CHILDES identifier (CHI, MOT, FAT, etc.), and the range of ages (or age of corresponding child) for which they are observed.
+For non-child participants (caregivers and others), the record additionally contains an identifier for the corresponding target child to associate caregivers and children.
 
@@
 
-In order to import the CHILDES corpora into the MySQL schema described above, it must first be accurately parsed and subsequently vetted to ensure the integrity of the data migration.
-Along with being made available in CHAT, the corpora are also downloadable in a format known as eXtensible Markup Language, or XML; this significantly mitigated any problems related to parsing CHAT as it allowed us to leverage the built-in XML parsing libraries available in numerous programming languages such as Python.
-Once the XML data were crawled and parsed into tokens and utterances, they were imported into a MySQL database.
-The meticulously constructed CHILDES XML Schema allowed us to make precise decisions as to what information we found relevant to our database and what was to be discarded or ignored; further information on what data were retained can be found below.
+In order to import the CHILDES corpora into the MySQL schema described above, it must first be accurately parsed and subsequently vetted to ensure its integrity.
+We parse the XML (eXtensible Markup Language) release of CHILDES hosted by \url{childes.talkbank.org} using the NLTK library in Python.
+<!-- citation for NLTK -->
+<!-- details of our contribution to the CHILDES corpus module in NLTK---didn't we change it? -->
+Logic implemented in Python converts the linear, multi-tier parse into a tabular format appropriate for childes-db.
+This logic includes decisions regarding what information sources are captured in the current release of the database, and which were left for future development (detailed below).
 
-The data imported into childes-db requires extensive data integrity checks to ensure that our parse of the corpora is accurate and preferable over the adhoc parsers developed by many individual researchers.
+<!-- testing details -->
+The data imported into childes-db requires extensive data integrity checks to ensure that our parse of the corpora is accurate and preferable over ad hoc parsers developed by many individual researchers.
 In order to evaluate this, we compared unigram counts in our database with those outputted by CLAN, the command-line tool built specifically for analysis of transcripts coded in CHAT.
-<!-- technically this is all we do right now -->
-
-<!-- [[Details of testing: Why trust our parse over your in-house library? CLAN TESTING DETAILS - also mention versioning as a reason to trust]] -->
+<!-- SM: what were the results of these comparisons?  -->
+<!-- AS: technically this is all we do right now SM: what more can we do for low technical investment-->
 
@@
 
-<!-- Continue to make the case that it's important to version! childes changes without warning, corpora get added, etc. -->
-
-The content of CHILDES changes as additional corpora are added or transcriptions are updated; currently, these changes are not systematically tracked.
-Here we introduce a simple versioning system by proucing a new complete parse of the current state of CHILDES every 6 months.
-Users interact with the most recent version of the database available by default. 
-To support reproduction of results with previous versions of the database, we continue to host recent versions (up to the last 6 versions) through our `childesr` API so that researchers can run analyses against specific versions of the database.
+The content of CHILDES changes as additional corpora are added or transcriptions are updated; as of the time of writing, these changes are not systematically tracked.
+Here we introduce a simple versioning system by adding a new complete parse of the current state of CHILDES every six months.
+By default, users interact with the most recent version of the database available. 
+To support reproduction of results with previous versions of the database, we continue to host recent versions (up to the last three years / six versions) through our `childesr` API so that researchers can run analyses against specific hsitorical versions of the database.
 <!-- Potentially put some R code to show easy it is to access previous versions? -->
-For versions more than three years old, users have the options of downloading compressed .sql files and serving them with a local installation of MySQL server.
-<!-- How easy is it to implement versioning in the Shiny apps?-->
+For versions more than three years old, we host compressed .sql files that users may download and serve using a local installation of MySQL server.
+<!-- SM: How easy is it to implement versioning in the Shiny apps? This would show anyone using the web app the availability of the versions-->
+<!-- SM: can people get to these sql files currently on the website?-->
 
@@
 - Sparsely annotated tiers, e.g. phonology (`%pho`) and situation (`%sit`)
-- Media links, including timing information <!-- we have media timestamps -->
-- compound word markers, indicated with underscores ("green_eggs_and_ham") <!-- we have this -->
+<!--- Media links, including timing information AS: we have media timestamps -->
+<!--- Compound word markers, indicated with underscores ("green_eggs_and_ham") AS: we have this -->
 - Tone direction and stress

### 2018-04-11 paper.Rmd
shiny app edits so far

--- 2018-04-11 paper.Rmd
@@
 library(directlabels)
-library(koRpus)
-library(png)
 library(here)
@@
 - Media links, including timing information <!-- we have media timestamps -->
-- compound word markers, indicated with underscores ("green_eggs_and_ham") <!-- we have this -->
+- compound word markers, indicated with underscores ("green_eggs_and_ham") <!-- we have this too -->
 - Tone direction and stress
@@
 
-childes-db offers a collection of browser-based interactive visualizations that enable data exploration through easy-to-use interfaces that do not require any programming knowledge.
-These tools are intended for casual users who seek to extract useful developmental indices from CHILDES without any technical overhead.
-The visualizations available through childes-db are the following: Frequency Counts (lexical statistics), Derived Measures (lexical diversity), and Population Viewer (corpora statistics).
-All three of these visualizations were built using Shiny, an R package that enables easy building of web applications in R.
-For each visualization, a user specifies a subset of the data (in particular the collection, corpus, child, and speaker) and is also able to restrict the age range of the child with slider bars.
-Any number of corpora, children, or speakers may be selected for analysis, allowing for aggregate counts across a whole range of data in the CHILDES system, a task that could not have been easily done using previous methods.
-Along with the visualizations, childes-db makes the raw data itself available for download under a separate “Table” tab; this allows researchers to further explore the underlying data from the plots.
+The ability to easily browse and explore the CHILDES corpora is a cornerstone of the childes-db project.
+To this end we have created powerful yet easy-to-use interactive web applications that enable users to visualize various dimensions of the CHILDES corpus: frequency counts, mean lengths of utterance, type-token ratios, and more.
+All of this is doable without the requirement of understanding command-line tools or any kind of programming knowledge as had been the case with CLAN.
+The only true requirements are a browser and internet connection.
+
+The interactive plots (web apps) are built using Shiny, an R package that enables easy building of web applications in R.
+Underneath the hood, each web app is making calls to our `childesr` API and subsequently plotting the data using the R plotting package `ggplot2`.
+A user's only task is to configure exactly what should be plotted through a series of buttons, sliders, and text boxes.
+The user may specify what collection, corpus, child, age range, caregiver, etc., should be included in a given analysis (e.g.
+mean length of utterance).
+The plot is displayed and updated in real-time, and if a user were to feel the need to explore the data independently then that data is also available for download alongside the plot.
+All of these analyses may also be reproduced only using the `childesr` package; the web apps were intended for the casual user who seeks to easy extract developmental indices without any technical overhead.
+
+In the following sections we will walk through the three web apps made available on our website and describe their functionality in more detail.
 
 ### Frequency Counts
+
+A child's lexical statistics has long been an object of study in child language acquisition research.
+Namely, the frequency of a word <!-- something important about frequency and development?-->.
+In this web app, we provide users the ability to search for any word or string
+
+The study of frequency counts of individual lexical items as long been the
 
@@
 Teachers of courses on early language acquisition often want to demonstrate interesting changes in children’s early language.
-One method is to present static displays that show text from parent-child conversations extracted from CHILDES or data visualizations of various metrics of production and input (e.g., MLU or Frequency).
-The challenge with static displays is that they cannot be modified during a lecture and thus rely on the instructor selecting examples that will be compelling to students.
+One teaching method is to present static displays on slides or handouts that show (a) text from parent-child conversations extracted from CHILDES or (b) data visualizations of changes in various metrics related to language production and input (e.g., MLU or Frequency).
+The challenge with static displays or handouts is that they cannot be modified during a lecture and thus rely on the instructor selecting examples that they think will be compelling to their students.
 
-In-class demonstrations offer a solution and can be a powerful way to explain complex concepts while increasing student engagement with the course materials.
-One goal of creating the web applications in childes-db is to give instructors a new tool for illustrating concepts in children's language acquisition.
+Interactive, in-class demonstrations offer a solution and can be a powerful way to explain complex concepts while increasing student engagement with the course materials.
+One goal in creating the browser-based interactive visualization tools in childes-db is to provide instructors with a new tool for illustrating key concepts in children's language acquisition.
+In this section, we walk through one example using childes-db to show the developmental trajectories of different kinds of words that in children's early production.
+We also highlight ways that instructors could make the in-class demonstration interactive by eliciting hypotheses from students, which become operationalized as queries to the interactive visualization tools.
 
-For example, a teacher could want to show the types of words that children first produce.
+### Worked example
+
+What kinds of words do children first produce?
+This question is a good candidate for an interesting "hook" that gets students thinking about some key issues in child language acquisition.
 Diary studies and large-scale studies using parent report show that children’s first words tend to fall into a fairly small number of categories: people, food, body parts, clothing, animals, vehicles, toys, household objects, routines, and activities or states [@clark2009first; @fenson1994variability].
@@
 A teacher could illustrate children's early vocabularies by showing a table with columns for the lexical category, examples of words from that category, and children's average age of production.
-To make this interactive using the web apps, the teacher could:
+To make this an interactive demonstration using childes-db, the teacher could:
 
-1. introduce the research question (e.g., What are the types of words that children first produce?)
+1. introduce the question: What kinds of words do children first produce?
 2. allow students to reflect or do a pair-and-share discussion with their neighbor
-3. show the trajectory of a single lexical item while explaining key parts of the visualization (see Panel A of Figure \ref{fig:in-class-demo-plot})
+3. show the trajectory of a single lexical item while explaining the key parts of the visualization (see Panel A of Fig. XX)
 4. elicit hypotheses from students about the kinds of words that children are likely to produce 
-5. make real-time queries to the web application to add students' suggestions and talk through the updated plots (Panels B and C of Figure \ref{fig:in-class-demo-plot})
-6. finish by entering a pre-selected set of words that communicate the important takeaway point (Panel D of Figure \ref{fig:in-class-demo-plot})
+5. make real-time queries to the interactive visualization tool to add students' suggestions and talk through the updated plots 
+6. finish by entering a pre-selected set of words that communicate the critical takeaway (e.g., see Panel C of Fig. XX)
 
-```{r in-class-demo-plot, fig.pos="h", out.width = "95%", fig.cap="Worked example of using the web applications for in-class teaching.
-Panels A-D show how an instructor could dynamically build a plot during a lecture to demonstrate a key concept in language acquisition."}
-
+```{r in-class-demo-flow, fig.pos="h", out.width = "100%", fig.cap="Worked example of using the interactive visualization tools for in-class teaching.
+Panels A-D show how an instructor could dynamically build a plot during a lecture to interactively demonstrate a key concept in language acquisition."}
+library(here)
+library(png)
 png::readPNG(here::here("imgs/in_class_teach_viz.png")) %>% grid::grid.raster()
@@
 
-In addition to the web applications, instructors can use childes-db to download the raw data used to generate the figures.
-This gives students the opportunity to further explore and analyze the data they see in the plots.
-Moreover, this functionality provides an example of a second way that childes-db can facilitate language acquisition courses by making data easily-accessible for tutorials and student research projects.
+Another benefit of this workflow is that the instructor could easily reuse it to demonstrate other concepts or questions that come up during the lecture, e.g., order of production for different types of words such as quantifiers vs.
+number words or function vs.
+content words.
+Moreover, future work on childes-db interactive visualization tools could add part of speech tags, phonological information, and n-grams.
+This would allow instructors to use childes-db to demonstrate concepts such as when children start producing certain phonological patterns (e.g., consonant clusters), the debate about the extent of the noun/verb bias in children's early vocabulary, and the development of particular syntactic constructions.
 
-### Tutorials and programming assignments
+In addition to the interactive data visualizations, instructors can use childes-db to download the raw data used to generate the figures.
+This allows teachers an opportunity to have students further explore and analyze the data they see in the plots.
+Moreover, this functionality provides an example of another way childes-db can facilitate language acquisition courses by making data easily-accessible for tutorials and student research projects.
+In the next section, we walk through one example of how an instructor could use childes-db to create a tutorial that scaffolds students' ability to ask interesting computational questions using child language data.
 
-One of the goals for courses on applied natural language processing (NLP) is for students to get hands-on experience using NLP tools to analyze real-world language data.
+### Tutorials and assignments
+
+One of the goals for courses on applied natural language processing (NLP) is for students to get hands-on experience using state of the art NLP tools to analyze real-world language data [@hearst2005teaching].
 A primary challenge for the instructor is to decide how much time should be spent teaching the requisite programming skills for accessing and formatting language data, which are typically unstructured.
 One pedagogical strategy is to abstract away these details and avoid having students deal with obtaining data and formatting text.
-This approach shifts students' effort away from data cleaning and towards programming analyses that encourage the exploration and testing of interesting hypotheses.
+This approach shifts students' effort away from data cleaning and towards programming analyses that encourage students to explore language data and test interesting hypotheses.
 
-One goal of childes-db and `childesr` API is to provide instructors with an easy-to-learn method for giving students programmatic access to child language data.
-The aim is to reduce time spent getting data out of CHILDES and allow more course time for thinking about interesting questions that could be answered with NLP and child language data.
-Moreover, by writing our API in the R programming language, students will gain access to a general workflow can be integrated into an analysis pipeline that includes state of the art NLP tools.
+In the introduction, we outlined the challenge of getting students started with analyzing data from CHILDES.
+These barriers include learning the details of a special file format (CHAT) and a command-line tool (CLAN) built to specifically for the CHAT format.
+This workflow can be daunting to navigate for a beginner.
+Thus, an instructor would have to spend course time teaching students to use CHAT and CLAN, leaving less time for students to try analyses that might increase their interest in child language acquisition.
+Moreover,  while CLAN provides a useful tool for interactive exploration,  it does not connect to a pipeline of modern natural language processing tools [e.g., models of word embeddings; @mikolov2013distributed], making it difficult for students to explore the data using new NLP tools.
 
-For example, an instructor could create a programming assignment with the goal of reproducing the key findings in the case studies presented above -- color words or gender.
-Depending on the students' knowledge of R, the instructor could decide how much of the `childesr` starter code to provide before asking students to generate their own plots and writeups.
-Moreover, the instructor could take advantage of recent advances in tools for teaching literate programming such as R Markdown [@xie2015dynamic], building in pre-written blocks of executable code that demonstrate key programming concepts for students and including incomplete or blank code blocks that students can use to write their own code.
+We hope that childes-db combined with the `childesr` API will provide instructors with a solution to both of these challenges.
+The API gives students an easy-to-learn method for programmatic access to CHILDES data.
+This will reduce the time spent getting data out of CHILDES and allow more time for thinking about interesting questions that can be answered with NLP and child language data.
+Second, by writing our API in the R programming language, students will gain access to a general workflow can be integrated into a pipeline that includes the newest NLP tools.
+
+For example, an instructor could create a programming assignment with the goal of reproducing the key findings in the case studies that we presented earlier -- color words or gender.
+Depending on the students' knowledge of R, the instructor could decide how much of the `childesr` code to provide before asking students to generate their own plots and writeups.
+
+Moreover, the instructor could take advantage of recent advances in tools for teaching literate programming such as R Markdown [@xie2015dynamic; @baumer2014r].
+Writing the assignment in an R Markdown notebook would allow the interleaving of prose with executable chunks of code.
+Instructors could then write prose that is directly related to the course content and the assignment while building in pre-written blocks of executable code that demonstrate key programming concepts for students.
+To push students' understanding, instructors could write incomplete or blank code blocks that students can use to write their own code.
+In short, `childesr` combined with R Markdown provides a powerful teaching tool, allowing students to start analyzing language data quickly and providing instructors with a way to build customized lessons for students at different levels of programming experience.
 

### 2018-04-11 paper.Rmd
builds, reorg images
spell check plus abstract

@@
   The Child Language Data Exchange System (CHILDES) has played a critical role in research on child language development.
-However, traditional methods of inquiry into the data have often relied on a command line interface to extract developmental indices or custom parsers to extract data for computational modeling.
-Now, advances in open-source software allow for novel approaches in working with this dataset.
+However, current methods of using CHILDES data rely on a command line interface to extract developmental indices or custom parsers to extract data for larger-scale analyses.
 We introduce childes-db, a database-formatted mirror of CHILDES that improves data accessibility and usability by offering novel interfaces, including browsable web applications and an R application programming interface (API).
-Along with a versioned infrastructure that facilitates reproduction of past analyses, these interfaces lower barriers to analyzing naturalistic parent-child language, allowing for a wider range of researchers in language and cognitive development to easily leverage CHILDES in their work.
+Along with a versioned infrastructure that facilitates reproducibility of past analyses, these interfaces lower barriers to analyzing naturalistic parent-child language, allowing for a wider range of researchers in language and cognitive development to easily leverage CHILDES in their work.
 
@@
 library(here)
+library(lme4)
 knitr::opts_chunk$set(cache = TRUE, echo = FALSE)
@@
 The first examines the frequency of color words in children's input; the second quantifies the growth of lexical diversity across genders.
-We end by presenting some suggestions for the use of the childes-db web frontend in classroom teaching.
+We end by presenting some suggestions for the use of the childes-db web front-end in classroom teaching.
 
@@
 By default, users interact with the most recent version of the database available. 
-To support reproduction of results with previous versions of the database, we continue to host recent versions (up to the last three years / six versions) through our `childesr` API so that researchers can run analyses against specific hsitorical versions of the database.
+To support reproduction of results with previous versions of the database, we continue to host recent versions (up to the last three years / six versions) through our `childesr` API so that researchers can run analyses against specific historical versions of the database.
 <!-- Potentially put some R code to show easy it is to access previous versions? -->
@@
 We store these in the `colors` variable, and then use the `get_types` function from `childesr` to get the type frequency of each of these words in all of the corpora in CHILDES.
-For demonstration, we look only at the types produced by the speakers in each corpus oftagged as Mother and Father.
+For demonstration, we look only at the types produced by the speakers in each corpus tagged as Mother and Father.
 We also restrict ourselves to children from 1-5 years old (12 to 60 months), and look only the American English corpora.
@@
 To normalize correctly (i.e.
-to ask what proprotion of the input children hear consists of these color words), we need to know how many total words these children hear from their parents in these transcripts.
+to ask what proportion of the input children hear consists of these color words), we need to know how many total words these children hear from their parents in these transcripts.
 To do this, we use the `get_speaker_statistics` function, which will return a total number of tokens (`num_tokens`) for each of these speakers.
@@
 
-mod <- lme4::lmer(measure ~ target_child_sex * poly(age_years, 2) - 1 + 
+mod <- lmer(measure ~ target_child_sex * poly(age_years, 2) - 1 + 
                     (1 | target_child_id) +
@@
 
-png::readPNG(here::here("imgs/in_class_teach_viz.png")) %>% grid::grid.raster()
+png::readPNG(here::here("images/in_class_teach_viz.png")) %>% grid::grid.raster()
 ```
@@
 For example, an instructor could create a programming assignment with the goal of reproducing the key findings in the case studies presented above -- color words or gender.
-Depending on the students' knowledge of R, the instructor could decide how much of the `childesr` starter code to provide before asking students to generate their own plots and writeups.
+Depending on the students' knowledge of R, the instructor could decide how much of the `childesr` starter code to provide before asking students to generate their own plots and write-ups.
 Moreover, the instructor could take advantage of recent advances in tools for teaching literate programming such as R Markdown [@xie2015dynamic], building in pre-written blocks of executable code that demonstrate key programming concepts for students and including incomplete or blank code blocks that students can use to write their own code.

### 2018-04-11 paper.Rmd
complete freq

--- 2018-04-11 paper.Rmd
@@
 Namely, the frequency of a word <!-- something important about frequency and development?-->.
-In this web app, we provide users the ability to search for any word or string
+In this web app, we provide users the ability to search for any word spoken by a participant in the CHILDES corpora and track the usage of that word by a child or caregiver over time.
+Because of the various toggles available to the user that can subset the data, a user may display frequency curves for a single child in the Brown corpus or all Spanish speaking children if desired.
+In addition, users may plot frequency curves belonging to caregivers alongside their child for convenient visual comparisons.
+A single word or multiple words may be entered in the input box, separated only by a comma.
 
-The study of frequency counts of individual lexical items as long been the
-
-<!-- sm: intro sentence here for the role of frequency in CLA research  --> 
-The Frequency Counts app allows users to enter a word (or words) in a text box and see the frequency (in parts per million) in which this word was uttered across a child's age for a particular subset of the data, whether it is an individual child in the Brown corpus or all Spanish speaking children and their mothers.
-<!-- sm: "whether..." is awk.
--->
-The ability to view the caregiver’s usage of a term alongside a child’s assists researchers in better understanding the relationship between input and uptake.
-Users can also plot word trajectories alongside each other (such as “mom” and “dad” or “front” and “back”) for convenient cross-word comparisons.
+<!-- figure of freq app -->
 
 ### Derived Measures
+
+Characterizing lexical diversity has been another highly utli
 

### 2018-04-11 paper.Rmd
intro

--- 2018-04-11 paper.Rmd
@@
   The Child Language Data Exchange System (CHILDES) has played a critical role in research on child language development.
-However, current methods of using CHILDES data rely on a command line interface to extract developmental indices or custom parsers to extract data for larger-scale analyses.
-We introduce childes-db, a database-formatted mirror of CHILDES that improves data accessibility and usability by offering novel interfaces, including browsable web applications and an R application programming interface (API).
+Access to these data can be complex for novices and difficult to automate for advanced users, however.
+To address these issues, we introduce childes-db, a database-formatted mirror of CHILDES that improves data accessibility and usability by offering novel interfaces, including browsable web applications and an R application programming interface (API).
 Along with a versioned infrastructure that facilitates reproducibility of past analyses, these interfaces lower barriers to analyzing naturalistic parent-child language, allowing for a wider range of researchers in language and cognitive development to easily leverage CHILDES in their work.
@@
 Children's language acquisition is a core puzzle in cognitive science.
-How are the representations underlying the generative use of language learned over children's first years?
+How are the representations underlying the productive use of language learned over children's first years?
 Because a language is comprised of many individual pieces of information -- the names of particular objects, the marker for the past tense, the position of determiners -- language development is fundamentally a question of learning from environmental input, which is the only source of information on these specifics.
@@
 But creating individual transcripts is extremely hard work.
-Beyond gathering the data, even orthographic transcriptions (without phonetic detail) can take ten times the length of the original recording to transcribe.
+Beyond gathering the data, even orthographic transcriptions (without phonetic detail) can take ten times the length of the original recording to transcribe [@macwhinney2000].
 Thus, researchers' desire for data from multiple ages or multiple children can very quickly outstrip their resources.
@@
 Yet there are some challenges working with CHILDES, both for students and for advanced users.
-The CHILDES ecosystem uses a special file format (CHAT), which is stored in a plain text format but includes structured annotations grouped into tiers (stored on separate lines) so that information about utterances can be stored with accompanying information such as the phonological or morphological structure of the utterance.
+The CHILDES ecosystem uses a special file format (CHAT), which is stored as plain text but includes structured annotations grouped into tiers (stored on separate lines) so that information about utterances can be stored with accompanying information such as the phonological or morphological structure of the utterance.
 These files are analyzed using a command-line tool (CLAN) with a host of specific arguments that allow users to count word frequencies, compute statistics (e.g., mean length of utterance, or MLU).
-While this system is flexible and powerful, mastering the CHAT codes and the CLAN tool can be daunting to navigate for a beginner or student.
+While this system is flexible and powerful, mastering the CHAT codes and the CLAN tool can be daunting for a beginner or student.
 These technical barriers decrease the ease of casual exploration by a novice researcher or in a classroom exercise.
@@
 On the opposite end of the spectrum, for data-oriented researchers who are interested in doing large-scale analyses of CHILDES, the current tools are also not ideal.
-CLAN software is an excellent tool for interactive exploration, but -- as a free-standing application -- can be tricky to build into a processing pipeline written in a modern analytic platform such as Python or R.
-Thus, researchers who would like to ingest the entire corpus (or some large subset) into a computational analysis typically write their own parsers of the CHAT format to extract the subset of the data they would like to use [e.g., @redington1998;@yang2013;@meylan2017].
+CLAN software is an excellent tool for interactive exploration, but -- as a free-standing application -- it can be tricky to build into a processing pipeline written in a modern analytic language such as Python or R.
+Thus, researchers who would like to ingest the entire corpus (or some large subset) into a computational analyses typically write their own parsers of the CHAT format to extract the subset of the data they would like to use [e.g., @redington1998;@yang2013;@meylan2017].
 
-This practice is problematic for a number of reasons.
+The practice of writing custom parsers is problematic for a number of reasons.
 Effort is wasted in rewriting the same parser again and again.
 In addition, this process introduces a major source of errors and inconsistencies in data handling due to difficulties dealing with the many special cases in the CHAT standard.
-Further, these scripts are rarely shared, leading to much greater difficulty in reproducing the exact numerical results from papers using CHILDES [see e.g., @meylan2017].
-This lack of computational reproducibility is widely recognized as problematic because it does not allow researchers to verify or build on published research [@donoho2010;@stodden2016].
-Finally, the CHILDES corpus itself is a moving target;  computational work using the entire corpus at one time point may include a different set of data than subsequent work (due to new contributions or revisions).
+Further, these parsing scripts are rarely shared, leading to much greater difficulty in reproducing the exact numerical results from papers using CHILDES [see e.g., @meylan2017 for an example].
+In addition, the CHILDES corpus itself is a moving target; computational work using the entire corpus at one time point may include a different set of data than subsequent work (due to new contributions or revisions) and there is no simple way for researchers to document exactly what version of the corpus has been used short of creating a full mirror of the data.
+These factors together lead to a lack of computational reproducibility, a major problem that keeps researchers from verifying or building on published research [@donoho2010;@stodden2016].
 

### 2018-04-11 paper.Rmd
finish shiny sections

--- 2018-04-11 paper.Rmd
@@
 To this end we have created powerful yet easy-to-use interactive web applications that enable users to visualize various dimensions of the CHILDES corpus: frequency counts, mean lengths of utterance, type-token ratios, and more.
-All of this is doable without the requirement of understanding command-line tools or any kind of programming knowledge as had been the case with CLAN.
+All of this is doable without the requirement of understanding command-line tools or any kind of programming knowledge, as had been the case with CLAN.
 The only true requirements are a browser and internet connection.
@@
 mean length of utterance).
-The plot is displayed and updated in real-time, and if a user were to feel the need to explore the data independently then that data is also available for download alongside the plot.
-All of these analyses may also be reproduced only using the `childesr` package; the web apps were intended for the casual user who seeks to easy extract developmental indices without any technical overhead.
+The plot is displayed and updated in real-time, and if a user were to want to explore the data independently then that data is also available for download alongside the plots.
+Every one of these analyses can also be produced using only the `childesr` package; the web apps were intended for the casual user who seeks to easily extract developmental indices without any technical overhead.
 
@@
 A child's lexical statistics has long been an object of study in child language acquisition research.
-Namely, the frequency of a word <!-- something important about frequency and development?-->.
+Namely, the frequency counts of various words may provide insight into the cognitive, conceptual, and linguistic experience of a young child.
 In this web app, we provide users the ability to search for any word spoken by a participant in the CHILDES corpora and track the usage of that word by a child or caregiver over time.
-Because of the various toggles available to the user that can subset the data, a user may display frequency curves for a single child in the Brown corpus or all Spanish speaking children if desired.
-In addition, users may plot frequency curves belonging to caregivers alongside their child for convenient visual comparisons.
-A single word or multiple words may be entered in the input box, separated only by a comma.
+Because of the various toggles available to the user that can subset the data, a user may display frequency curves for a single child in the Brown corpus or all Spanish speaking children, if desired.
+In addition, users may plot frequency curves belonging to caregivers alongside their child for convenient side-by-side comparisons.
+A single word or multiple words may be entered into the input box, separated only by a comma.
+<!-- cite original ChildFreq paper as inspiration?
+-->
 
@@
 
-Characterizing lexical diversity has been another highly utli
+Researchers in child language acquisition have often assessed a child's linguistic performance by extracting various measures from their speech.
+These developmental metrics have also had a major application in speech-language pathology (SLP), where subpar measures may be indicative of speech, language, or communication disorders.
+The Derived Measures web app is built to provide the most common measures typically extracted from children's speech.
+The most famous of these measures, mean length of utterance (MLU), is used to characterize the linguistic productivity of children and is available through this web app.
+However, only MLU-w (mean length of utterance in words) is provided which has been shown to correlate with MLU-m (mean length of utterance in morphemes) in the English language <!-- citation needed -->.
+The other measures provided in this app are measures of lexical diversity: TTR (type-token ratio), MTLD (measure of textual lexical diversity) and HD-D (hypergeometric distribution of diversity) <!-- cite hd-d?
+mccarthy jarvis 2007 -->.
+As with the Frequency Counts app, a user may subset the data as he or she chooses, compare measures between caregivers and children, and aggregate across children from different corpora.
 
-<!-- sm: intro sentence here for the role of developmental metrics in CLA research and in downstream applications like SLP--> 
-There are a number of ways to measure the lexical diversity of an individual's productive speech. <!-- sm: different intro; MLU isn't a lexical diversity measure -->
-Some of these measures, including MLU-w (mean length of utterance in words), TTR (type-token ratio), MTLD (measure of textual lexical diversity), and voc-d are available in the Derived Measures app, which plots these measures across age for a given subset of data, again specified by collection, corpora, children, and speakers.
-<!-- sm: MLU is not a measure of lexical diversity-->
-Because of the underlying gaps in available CHILDES morphological data, MLU-w is computed and not MLU-m (mean length of utterance in morphemes).
-<!-- sm: I think we should go ahead and compute this for transcript with MOR tiers?
-Given that this was one of the first uses, I don't think Brian will like the 'underlying gaps' explanation-->
-In the English language, these two measures have been shown to be highly correlated. <!-- sm: citation needed -->
-As with the Frequency Counts app, caregiver’s lexical diversity measures can be plotted alongside those of children.
+<!-- figure of derived measures app -->
 
@@
 
-The population viewer is a tool that helps users understand the relative sizes and temporal extents of various corpora by displaying the total number of utterances or tokens in the available corpora at particular ages.
-In many instances the size of a given corpus may inform researchers’ decisions as to what is the best target dataset for analysis.
+Many times a researcher will want to investigate the statistics of a given corpus (i.e the size, number of utterances, number of tokens, etc.) before choosing a target corpus for their analysis.
+This web app is intended to provide a basic overview of various corpora in the CHILDES ecosystem, facilitating familiarity with the different corpora and thus assisting with the overall analytic end-to-end process.
 <!-- note that the graphs don't currently include a grouping factor, so adding multiple corpora generates one line, so you can't actually compare two corpora in the same visualization -->
+
+<!-- figure of population viewer app -->
 

### 2018-04-11 paper.Rmd
technical and access tweaks

--- 2018-04-11 paper.Rmd
@@
 We next describe the `childesr` API, which allows for programmatic access to the data.
-We then present two worked examples of specific uses of the system using both the web app and the API.
-The first examines the frequency of color words in children's input; the second quantifies the growth of lexical diversity across genders.
-We end by presenting some suggestions for the use of the childes-db web front-end in classroom teaching.
+We then present several worked examples of specific uses of the system for research and teaching using both the web app and the API.
 
@@
 <!-- Remind readers why they're reading this: understanding the structure helps you use and understand the tools. SM: Moved this up from database format because it addresses more general issues -->
-Unlike the *parsing* approach in CLAN which entails the sequential processing of strings, childes-db treats CHILDES as a set of linked tables, with records corresponding to intuitive abstractions such as words, utterances, and transcripts.
+As described above, CHILDES is most often approached as a set of distinct CHAT files, which are then parsed by users (often using CLAN). 
+In contrast to this parsing approach, which entails the sequential processing of strings, childes-db treats CHILDES as a set of linked tables, with records corresponding to intuitive abstractions such as words, utterances, and transcripts.
 Users of data analysis languages like R or Julia, libraries like Pandas, or those familiar with Structured Query Language (SQL) will be familiar with operations on tables such as filtering (subsetting), sorting, aggregation (grouping), and joins (merges).
-Although we expect most users to interact with the database through the `childesr` API for the `R` language described below, we begin with a top-level description and motivation for the design of the database schema, details of the database’s current technical implementation, and our versioning scheme.
+These operations obviate the need for users to consider the specifics of the CHAT representation -- instead they simply request the entities they need for their research and allow the API to take care of the formatting details.
+We begin by orienting readers to the design of the system via a top-level description and motivation for the design of the database schema, details of the database’s current technical implementation, and our versioning scheme.
+Users primarily interested in accessing the database can skip these details and focus on access through the `childesr` API and the web apps. 
 <!-- We first present the design and technical approach of childes-db, then describe the interactive analysis tools provided by our web-based frontend.  -->
@@
 
-At its core, childes-db is a database consisting of a set of linked tabular data stores where rows correspond to linguistic entities like words, utterances, and sampling units like transcriptions and corpora.
+At its core, childes-db is a database consisting of a set of linked tabular data stores where records correspond to linguistic entities like words, utterances, and sampling units like transcriptions and corpora.
 <!-- what did you mean here? don't rows = records; SM: that's a description of a database in general; we can make a specific assertion about childes-db-->
-The smallest unit of abstraction tracked by the database is a token, treated here as the citation orthographic form of a word.
+The smallest unit of abstraction tracked by the database is a token, treated here as the standard (citation) orthographic form of a word.
 Using the standardized written form of the word facilitates the computation of lexical frequency statistics, for comparison or aggregation across children or time periods. 
@@
 
-Other tables in the database comprise a hierarchical system of collections built on top of tokens -- *utterance*, *transcription*, *corpus*, and *collection* -- which stores attributes appropriate for each level of description.
+Other tables in the database are hierarchical collections built out of tokens -- *utterance*, *transcription*, *corpus*, and *collection* -- which store attributes appropriate for each level of description.
 Every entity includes attributes that link it to all higher-order collections, e.g., an utterance lists the transcript, corpus, and collection to which it belongs.
 An *utterance* contains one or more words and includes fields such as the utterance type (declarative, question, etc.), total number of tokens, and the total number of morphemes (if the morphological structure is available in the original CHAT).
-
 A *transcription* consists of one or more utterances and includes the date collected, the name of the target child, and age in days (if defined), and the filename from CHILDES.
@@
 
-We chose to treat participants separate from the above hierarchy because children can occur across transcripts (e.g., the Brown Corpus).
-A participant identifier is associated with every word and utterance, including a name, role, 3-letter CHILDES identifier (CHI, MOT, FAT, etc.), and the range of ages (or age of corresponding child) for which they are observed.
-For non-child participants (caregivers and others), the record additionally contains an identifier for the corresponding target child to associate caregivers and children.
+Participants (e.g., children) are represented separately from the token hierarchy because it is common for the same children to appear in multiple transcripts.
+A participant identifier is associated with every word and utterance, including a name, role, 3-letter CHILDES identifier (CHI = child, MOT = mother, FAT = father, etc.), and the range of ages (or age of corresponding child) for which they are observed.
+For non-child participants (caregivers and others), the record additionally contains an identifier for the corresponding target child to facilitate associating caregivers with children.
 
@@
 
-The above tabular datastore is implemented as a MySQL database.
-As an industry-standard, open-source relational database server, the above database can be accessed directly from a wide range of programming languages.
+childes-db is implemented as a MySQL database, an industry-standard, open-source relational database server that can be accessed directly from a wide range of programming languages.
 The childes-db project provides hosted read-only databases for direct access and for `childesr` (described below) as well as compressed .sql exports for local installation.
@@
 In order to import the CHILDES corpora into the MySQL schema described above, it must first be accurately parsed and subsequently vetted to ensure its integrity.
-We parse the XML (eXtensible Markup Language) release of CHILDES hosted by \url{childes.talkbank.org} using the NLTK library in Python.
-<!-- citation for NLTK -->
+We parse the XML (eXtensible Markup Language) release of CHILDES hosted by \url{childes.talkbank.org} using the NLTK library in Python [@bird2004].
 <!-- details of our contribution to the CHILDES corpus module in NLTK---didn't we change it? -->
 Logic implemented in Python converts the linear, multi-tier parse into a tabular format appropriate for childes-db.
-This logic includes decisions regarding what information sources are captured in the current release of the database, and which were left for future development (detailed below).
+This logic includes decisions regarding what information sources are captured in the current release of the database, and which are left for future development (detailed below).
 
-<!-- testing details -->
-The data imported into childes-db requires extensive data integrity checks to ensure that our parse of the corpora is accurate and preferable over ad hoc parsers developed by many individual researchers.
-In order to evaluate this, we compared unigram counts in our database with those outputted by CLAN, the command-line tool built specifically for analysis of transcripts coded in CHAT.
+The data imported into childes-db is subject to data integrity checks to ensure that our import of the corpora is accurate and preferable over ad-hoc parsers developed by many individual researchers.
+<!-- MCF: more testing details would be useful -->
+In order to evaluate our success in replicating CLAN parses, we compared unigram counts in our database with those outputted by CLAN, the command-line tool built specifically for analysis of transcripts coded in CHAT.
 <!-- SM: what were the results of these comparisons?  -->
@@
 The content of CHILDES changes as additional corpora are added or transcriptions are updated; as of the time of writing, these changes are not systematically tracked.
-Here we introduce a simple versioning system by adding a new complete parse of the current state of CHILDES every six months.
+We introduce a simple versioning system by adding a new complete parse of the current state of CHILDES every six months or as warranted by changes in CHILDES.
 By default, users interact with the most recent version of the database available. 
-To support reproduction of results with previous versions of the database, we continue to host recent versions (up to the last three years / six versions) through our `childesr` API so that researchers can run analyses against specific historical versions of the database.
-<!-- Potentially put some R code to show easy it is to access previous versions? -->
+But to support reproduction of results with previous versions of the database, we continue to host recent versions (up to the last three years / six versions) through our `childesr` API so that researchers can run analyses against specific historical versions of the database.
+<!-- Potentially put some R code to show easy it is to access previous versions? MCF: Nah. -->
 For versions more than three years old, we host compressed .sql files that users may download and serve using a local installation of MySQL server.
@@
 - Filled pauses 
-- Reformulations, word revision, and phrase revision, e.g. <what did you> [//] how can you see it ?
+- Reformulations, word revision, and phrase revision, e.g. \textlt what did you\textgt [//] how can you see it ?
 - paralinguistic material, e.g. [=! cries]
 
-We will prioritize the addition of these additional information sources in response to community feedback;
-extending coverage to the above information sources will not require changes to the existing database schema.
+We will prioritize the addition of these additional information sources in response to community feedback; extending coverage to the above information sources will not require changes to the existing database schema.
 
@@
 
-We first discuss the childes-db web apps and then introduce the `childesr` package.
+We first discuss the childes-db web apps and then introduce the `childesr` `R` package.
 
-## Interactive Analysis Tools
+## Interactive Web Apps
 
@@
 To this end we have created powerful yet easy-to-use interactive web applications that enable users to visualize various dimensions of the CHILDES corpus: frequency counts, mean lengths of utterance, type-token ratios, and more.
-All of this is doable without the requirement of understanding command-line tools or any kind of programming knowledge as had been the case with CLAN.
-The only true requirements are a browser and internet connection.
+All of this is doable without the requirement of understanding command-line tools or any kind of programming knowledge as had been the case with CLAN.^[The LuCiD toolkit [@chang2017] provides related functionality for a number of common analyses.
+In contrast to those tools, which focus on filling gaps not covered by CLAN -- e.g., the use of n-gram models, incremental sentence generation, and distributional word classification -- our web apps focus on covering the same common tasks as CLAN, but making the outputs into browsable visualizations.]
 
-The interactive plots (web apps) are built using Shiny, an R package that enables easy building of web applications in R.
-Underneath the hood, each web app is making calls to our `childesr` API and subsequently plotting the data using the R plotting package `ggplot2`.
+Our web apps are built using Shiny, an R package that enables easy app construction using `R`.
+Underneath the hood, each web app is making calls to our `childesr` API and subsequently plotting the data using the popular R plotting package `ggplot2`.
 A user's only task is to configure exactly what should be plotted through a series of buttons, sliders, and text boxes.
-The user may specify what collection, corpus, child, age range, caregiver, etc., should be included in a given analysis (e.g.
-mean length of utterance).
-The plot is displayed and updated in real-time, and if a user were to feel the need to explore the data independently then that data is also available for download alongside the plot.
-All of these analyses may also be reproduced only using the `childesr` package; the web apps were intended for the casual user who seeks to easy extract developmental indices without any technical overhead.
+The user may specify what collection, corpus, child, age range, caregiver, etc., should be included in a given analysis.
+The plot is displayed and updated in real-time, and the underlying data are also available for download alongside the plot to facilitate customization.
+All of these analyses may also be reproduced using the `childesr` package.
 
-In the following sections we will walk through the three web apps made available on our website and describe their functionality in more detail.
+The web apps are intended for the casual user who seeks to easy extract developmental indices quickly and without any technical overhead.
+We next walk through the three web apps provided on our website and describe their functionality in more detail.
 
@@
 
-A child's lexical statistics has long been an object of study in child language acquisition research.
-Namely, the frequency of a word <!-- something important about frequency and development?-->.
-In this web app, we provide users the ability to search for any word or string
-
-The study of frequency counts of individual lexical items as long been the
-
-<!-- sm: intro sentence here for the role of frequency in CLA research  --> 
-The Frequency Counts app allows users to enter a word (or words) in a text box and see the frequency (in parts per million) in which this word was uttered across a child's age for a particular subset of the data, whether it is an individual child in the Brown corpus or all Spanish speaking children and their mothers.
-<!-- sm: "whether..." is awk.
--->
+The lexical statistics of languaeg input to children have long been an object of study in child language acquisition research, with ocurrence frequency specifically posited to have critical importance [see e.g., @ambridge2015 for review].
+The Frequency Counts app allows users to enter a word (or words) in a text box and see the frequency (in parts per million) in which this word was uttered across a child's age for a particular subset of the data, whether for an individual child in the Brown corpus or for all Spanish speaking children and their mothers.
 The ability to view the caregiver’s usage of a term alongside a child’s assists researchers in better understanding the relationship between input and uptake.
@@
 
-<!-- sm: intro sentence here for the role of developmental metrics in CLA research and in downstream applications like SLP--> 
-There are a number of ways to measure the lexical diversity of an individual's productive speech. <!-- sm: different intro; MLU isn't a lexical diversity measure -->
-Some of these measures, including MLU-w (mean length of utterance in words), TTR (type-token ratio), MTLD (measure of textual lexical diversity), and voc-d are available in the Derived Measures app, which plots these measures across age for a given subset of data, again specified by collection, corpora, children, and speakers.
-<!-- sm: MLU is not a measure of lexical diversity-->
-Because of the underlying gaps in available CHILDES morphological data, MLU-w is computed and not MLU-m (mean length of utterance in morphemes).
+The syntactic complexity and lexial diversity of children's speech are similarly critical metrics for acquisition researchers [@miller1981,@watkins1995].
+There are a number of well-established measures of children's speech that operationalize complexity and diversity. 
+Several of the most common of these measures are available in the Derived Measures app, which plots these measures across age for a given subset of data, again specified by collection, corpora, children, and speakers.
+As with the Frequency Counts app, caregivers' lexical diversity measures can be plotted alongside children's.
+
+We have currently implemented:
+
+- MLU-w (mean length of utterance in words, a common measure of complexity),^[Because of gaps in available CHILDES morphological data, MLU-w is computed and not MLU-m (mean length of utterance in morphemes).
 <!-- sm: I think we should go ahead and compute this for transcript with MOR tiers?
-Given that this was one of the first uses, I don't think Brian will like the 'underlying gaps' explanation-->
-In the English language, these two measures have been shown to be highly correlated. <!-- sm: citation needed -->
-As with the Frequency Counts app, caregiver’s lexical diversity measures can be plotted alongside those of children.
+Given that this was one of the first uses, I don't think Brian will like the 'underlying gaps' explanation.
+MCF: I agree.
+-->
+In English, these two measures are highly correlated. <!-- sm: citation needed -->
+alongside those of children.] 
+- TTR (type-token ratio, a measure of lexical diversity), 
+- MTLD (measure of textual lexical diversity), 
+- and voc-d (a further measure of lexical diversity).  <!-- sm: MLU is not a measure of lexical diversity-->
+<!-- MCF: need citations for voc-d and MTLD -->
 
@@
 Although the interactive analysis tools described above cover some of the most common use cases of CHILDES data, researchers interested in more detailed and flexible analyses will want to interface directly with the data in childes-db.
-Making use of the `R` programming language [@R-base], we provide the `childesr` package to help researchers accomplish this task.
+Making use of the `R` programming language [@R-base], we provide the `childesr` package.
 `R` is an open-source, extensible statistical computing environment that is rapidly growing in popularity across fields and is increasing in use in child language research [e.g.
 @song2015; @norrman2015].
-The `childesr` package abstracts away the details of connecting to <!-- sm: and querying-->  the database.
+The `childesr` package abstracts away the details of connecting to and querying  the database.
 Users can take advantage of the tools developed in the popular `dplyr` package [@R-dplyr], which make manipulating large datasets quick and easy.
@@
 To install, simply type: `install.packages("childesr")`.
-After installation, users have access to data loading functions that retrieve a variety of subsets <!-- sm: entities?
-variety of subsets sounds non-hierarchical-->  of the data:
+After installation, users have access to data loading functions that can be used to access different parts of the database:
 
@@
 
-Each of these functions can be supplied with arguments that restrict the query to a particular subset of the data (e.g.
-by collection, by corpus, by speaker role, by target child age, etc.) For more detailed documentation, see the package repository (http://github.com/langcog/childesr).
+\noindent Each of these functions can be supplied with arguments that restrict the query to a particular subset of the data (e.g.
+by collection, by corpus, by speaker role, by target child age, etc.) and for a particular version of the data.
+For more detailed documentation, see the package repository (http://github.com/langcog/childesr).
 

### 2018-04-11 paper.Rmd

@@
 
-<!-- need a good shorthand for the interactive analysis tools: web apps -->
-
 The ability to easily browse and explore the CHILDES corpora is a cornerstone of the childes-db project. 
@@
 
-The lexical statistics of languaeg input to children have long been an object of study in child language acquisition research, with ocurrence frequency specifically posited to have critical importance [see e.g., @ambridge2015 for review].
-The Frequency Counts app allows users to enter a word (or words) in a text box and see the frequency (in parts per million) in which this word was uttered across a child's age for a particular subset of the data, whether for an individual child in the Brown corpus or for all Spanish speaking children and their mothers.
-The ability to view the caregiver’s usage of a term alongside a child’s assists researchers in better understanding the relationship between input and uptake.
-Users can also plot word trajectories alongside each other (such as “mom” and “dad” or “front” and “back”) for convenient cross-word comparisons.
+The lexical statistics of languaeg input to children have long been an object of study in child language acquisition research.
+Frequency counts of words in particular may provide insight into the cognitive, conceptual, and linguistic experience of a young child [see e.g., @ambridge2015 for review].
+In this web app, inspired by ChildFreq [@baath2015], we provide users the ability to search for any word spoken by a participant in the CHILDES corpora and track the usage of that word by a child or caregiver over time.
+Because of the various toggles available to the user that can subset the data, a user may display frequency curves for a single child in the Brown corpus or all Spanish speaking children, if desired.
+In addition, users can plot frequency curves belonging to caregivers alongside their child for convenient side-by-side comparisons.
+A single word or multiple words may be entered into the input box.
 
@@
 There are a number of well-established measures of children's speech that operationalize complexity and diversity.
+These developmental metrics have many applications in speech-language pathology (SLP), where measures outside of the normal range may be indicative of speech, language, or communication disorders.
+
 Several of the most common of these measures are available in the Derived Measures app, which plots these measures across age for a given subset of data, again specified by collection, corpora, children, and speakers.
@@
 - MTLD (measure of textual lexical diversity), 
+- HD-D (lexical diversity via the hypergeometric distribution), 
 - and voc-d (a further measure of lexical diversity).  <!-- sm: MLU is not a measure of lexical diversity-->
-<!-- MCF: need citations for voc-d and MTLD -->
+<!-- MCF: need citations for voc-d HD-dand MTLD -->
+
+\noindent As with the Frequency Counts app, a user may subset the data as he or she chooses, compare measures between caregivers and children, and aggregate across children from different corpora.
 
@@
 
-The population viewer is a tool that helps users understand the relative sizes and temporal extents of various corpora by displaying the total number of utterances or tokens in the available corpora at particular ages.
-In many instances the size of a given corpus may inform researchers’ decisions as to what is the best target dataset for analysis.
+Many times a researcher will want to investigate the statistics of a given corpus (i.e the size, number of utterances, number of tokens, etc.) before choosing a target corpus for their analysis.
+This web app is intended to provide a basic overview of various corpora in the CHILDES ecosystem, facilitating familiarity with the different corpora and thus assisting with the overall analytic end-to-end process.
 <!-- note that the graphs don't currently include a grouping factor, so adding multiple corpora generates one line, so you can't actually compare two corpora in the same visualization -->
+ <!-- cite original ChildFreq paper as inspiration? -->
+
+<!-- figure of freq app -->
 

### 2018-04-11 paper.Rmd
up to conclusion

@@
 One common use of the CHILDES corpora is estimating the frequency with which children hear different words.
-These frequency estimates are used both in the development of theory [e.g.
-frequent words are learned earlier; @goodman2008], and in the construction of age-appropriate experimental stimuli.
+These frequency estimates are used both in the development of theory [e.g., frequent words are learned earlier; @goodman2008], and in the construction of age-appropriate experimental stimuli.
 One benefit of the childes-db interface is that it allows for easy analysis of how the frequencies of words change over development.
@@
 
-To demonstrate how one might discover such a non-stationary, we take as a case study the frequency with which children hear the color words of English (e.g.
+To demonstrate how one might discover such non-stationary, we take as a case study the frequency with which children hear the color words of English (e.g.
 "blue", "green").
@@
 
-To investigate whether the frequency of color words is stationary over development, a user can navigate to the Frequency app, and enter a set of color words into the `Word` selector separated by a comma: here "blue, red, green." Because the question of interest is about the frequency of words in the input (rather than produced by children), the `Speaker` field can be set to reflect this.
-In this example we select "Mother" because, across corpora in CHILDES, the speaker tagged with the Mother role tends to produce most of the words that children hear.
-Because children learn most of their basic color words by the age of 5, the age range of 1-5 years is a reasonable choice for `Ages to include`.
+To investigate whether the frequency of color words is stationary over development, a user can navigate to the Frequency app, and enter a set of color words into the `Word` selector separated by a comma: here "blue, red, green." Because the question of interest is about the frequency of words in the input (rather than produced by children), the `Speaker` field can be set to reflect this choice.
+In this example we select "Mother." Because children learn most of their basic color words by the age of 5, the age range 1--5 years is a reasonable choice for `Ages to include`.
 The results of these selections are shown in Figure \ref{fig:color-shiny}.
-We can also create a hyperlink to store these set of choices so that we can share these results with others, or with ourselves in the future, by clicking on the `Bookmark` button in the bottom left corner.
+We can also create a hyperlink to store these set of choices so that we can share these results with others (or with ourselves in the future) by clicking on the `Bookmark` button in the bottom left corner.
 
@@
 We also do not have a good sense of the errors of these measurements, are limited to just a few colors at a time before the plot becomes too busy, and cannot combine frequencies across speakers.
-To perform this analysis in a more compelling and complete way, a user can use the `childesr` interface to get more direct access to the CHILDES corpora.
+To perform this analysis in a more compelling and complete way, a user can use the `childesr` interface.
 
@@
 
-We can analyze these learning trajectories using childesr by braking the process into five steps: (1) define our words of interest, (2) find the frequencies with with which children hear these words, (3) Find the proportion of the *total words* children hear that these frequencies account for, (4) Aggregate across transcripts and children to determine the error in our estimates of these proportions, and (5) plot the results.
+We can analyze these learning trajectories using childesr by breaking the process into five steps: (1) define our words of interest, (2) find the frequencies with with which children hear these words, (3) find the proportion of the *total words* children hear that these frequencies account for, (4) aggregate across transcripts and children to determine the error in our estimates of these proportions, and (5) plot the results.
 
@@
 For demonstration, we look only at the types produced by the speakers in each corpus tagged as Mother and Father.
-We also restrict ourselves to children from 1-5 years old (12 to 60 months), and look only the American English corpora.
+We also restrict ourselves to children from 1-5 years old (12 to 60 months), and look only at the American English corpora.
 
@@
 
-To normalize correctly (i.e.
-to ask what proportion of the input children hear consists of these color words), we need to know how many total words these children hear from their parents in these transcripts.
+To normalize correctly (i.e., to ask what proportion of the input children hear consists of these color words), we need to know how many total words these children hear from their parents in these transcripts.
 To do this, we use the `get_speaker_statistics` function, which will return a total number of tokens (`num_tokens`) for each of these speakers.
@@
 We then group the data into 6-month age bins, and compute the proportion of tokens that comprise each color for each child in each 6-month bin.
-For comparability with the web app analysis, we convert the proportions to parts per million words
+For comparability with the web app analysis, we convert the proportions to parts per million words.
+
 ```{r join-colors-and-tokens, echo = TRUE}
@@
 ### Gender
+
 ```{r}
@@
 
-### In class demonstrations
+### In-class demonstrations
 
-Teachers of courses on early language acquisition often want to demonstrate interesting changes in children’s early language.
-One method is to present static displays that show text from parent-child conversations extracted from CHILDES or data visualizations of various metrics of production and input (e.g., MLU or Frequency).
-The challenge with static displays is that they cannot be modified during a lecture and thus rely on the instructor selecting examples that will be compelling to students.
+Teachers of courses on early language acquisition often want to illustrate the striking developmental changes in children’s early language.
+One method is to present static displays that show text from parent-child conversations extracted from CHILDES or data visualizations of various metrics of production and input (e.g., MLU or Frequency), but one challenge of such graphics is that they cannot be modified during a lecture and thus rely on the instructor selecting examples that will be compelling to students.
+In contrast, in-class demonstrations can be a powerful way to explain complex concepts while increasing student engagement with the course materials.
 
-In-class demonstrations offer a solution and can be a powerful way to explain complex concepts while increasing student engagement with the course materials.
-One goal of creating the web applications in childes-db is to give instructors a new tool for illustrating concepts in children's language acquisition.
+Consider the following demonstration about children's first words.
+Diary studies and large-scale studies using parent report show that children’s first words tend to fall into a fairly small number of categories: people, food, body parts, clothing, animals, vehicles, toys, household objects, routines, and activities or states [@clark2009first;@fenson1994variability;@tardif2008].
+The key insight is that young children talk about what is going on around them: people they see every day, e.g., toys and small household objects they can manipulate or food they can control.
+To illustrate this point, an instructor could:
 
-For example, a teacher could want to show the types of words that children first produce.
-Diary studies and large-scale studies using parent report show that children’s first words tend to fall into a fairly small number of categories: people, food, body parts, clothing, animals, vehicles, toys, household objects, routines, and activities or states [@clark2009first; @fenson1994variability].
-The key insight is that young children talk about what is going on around them: people they see every day, e.g., toys and small household objects they can manipulate or food they can control.
-
-A teacher could illustrate children's early vocabularies by showing a table with columns for the lexical category, examples of words from that category, and children's average age of production.
-To make this interactive using the web apps, the teacher could:
-
-1. introduce the research question (e.g., What are the types of words that children first produce?)
-2. allow students to reflect or do a pair-and-share discussion with their neighbor
-3. show the trajectory of a single lexical item while explaining key parts of the visualization (see Panel A of Figure \ref{fig:in-class-demo-plot})
-4. elicit hypotheses from students about the kinds of words that children are likely to produce 
-5. make real-time queries to the web application to add students' suggestions and talk through the updated plots (Panels B and C of Figure \ref{fig:in-class-demo-plot})
-6. finish by entering a pre-selected set of words that communicate the important takeaway point (Panel D of Figure \ref{fig:in-class-demo-plot})
+1. introduce the research question (e.g., What are the types of words that children first produce?),
+2. allow students to reflect or do a pair-and-share discussion with their neighbor,
+3. show the trajectory of a single lexical item while explaining key parts of the visualization (see Panel A of Figure \ref{fig:in-class-demo-plot}),
+4. elicit hypotheses from students about the kinds of words that children are likely to produce,
+5. make real-time queries to the web application to add students' suggestions and talk through the updated plots (Panels B and C of Figure \ref{fig:in-class-demo-plot}), and 
+6. finish by entering a pre-selected set of words that communicate the important takeaway point (Panel D of Figure \ref{fig:in-class-demo-plot}) .
 
@@
 
-In addition to the web applications, instructors can use childes-db to download the raw data used to generate the figures.
+<!-- In addition to the web applications, instructors can use childes-db to download the raw data used to generate the figures.
 This gives students the opportunity to further explore and analyze the data they see in the plots.
 Moreover, this functionality provides an example of a second way that childes-db can facilitate language acquisition courses by making data easily-accessible for tutorials and student research projects.
+-->
 
@@
 
-One of the goals for courses on applied natural language processing (NLP) is for students to get hands-on experience using NLP tools to analyze real-world language data.
+One goal for courses on applied natural language processing (NLP) is for students to get hands-on experience using NLP tools to analyze real-world language data.
 A primary challenge for the instructor is to decide how much time should be spent teaching the requisite programming skills for accessing and formatting language data, which are typically unstructured.
@@
 
-One goal of childes-db and `childesr` API is to provide instructors with an easy-to-learn method for giving students programmatic access to child language data.
+The `childesr` API provides instructors with an easy-to-learn method for giving students programmatic access to child language data.
 The aim is to reduce time spent getting data out of CHILDES and allow more course time for thinking about interesting questions that could be answered with NLP and child language data.
@@
 # Conclusion
+
+We have presented childes-db, a database formatted mirror of the CHILDES dataset.
+This database -- together with the `R` API and web apps -- facilitates the use of child language data.
+For teachers, students, and casual explorers, the web apps allow browsing and demonstration.
+For researchers interested in scripting more complex analyses, the API allows them to abstract away from the details of the CHAT format and easily create reproducible analyses of the data.
+We hope that these functionalities broaden the set of users who can easily interact with CHILDES data, leading to future insights into the process of language acquisition.
+
+Of course,
 

### 2018-04-11 paper.Rmd
finished pass

@@
 
-The lexical statistics of languaeg input to children have long been an object of study in child language acquisition research.
+The lexical statistics of language input to children have long been an object of study in child language acquisition research.
 Frequency counts of words in particular may provide insight into the cognitive, conceptual, and linguistic experience of a young child [see e.g., @ambridge2015 for review].
@@
 
-The syntactic complexity and lexial diversity of children's speech are similarly critical metrics for acquisition researchers [@miller1981,@watkins1995].
+The syntactic complexity and lexical diversity of children's speech are similarly critical metrics for acquisition researchers [@miller1981,@watkins1995].
 There are a number of well-established measures of children's speech that operationalize complexity and diversity.
@@
 
-Of course,
-
-<!-- reiterate importance -->
-
-<!-- adding more tiers is a goal for the future -->
+We have created childes-db to address a number of needs in our own teaching and research, but there are still a number of limitations that point the way to future improvements.
+For example, currently childes-db operates only on transcript data, without links to the underlying media files; these linkages facilitate deeper analyses of phonology, prosody, social interaction, and many other aspects of acquisition.
+Further, we have focused on including the most common and widely-used tiers of CHAT annotation into the database first, but our plan is eventually to include the full range of tiers.
+Finally, a wide range of further interactive analyses could easily be added to the current suite of web apps.
+We invite other researchers to join us in both suggesting and contributing new functionality as our system grows and adapts to researchers' needs.
 

### 2018-04-11 paper.Rmd
transcription -> transcript

--- 2018-04-11 paper.Rmd
@@
   The Child Language Data Exchange System (CHILDES) has played a critical role in research on child language development.
-Access to these data can be complex for novices and difficult to automate for advanced users, however.
-To address these issues, we introduce childes-db, a database-formatted mirror of CHILDES that improves data accessibility and usability by offering novel interfaces, including browsable web applications and an R application programming interface (API).
+However, current methods of using CHILDES data rely on a command line interface to extract developmental indices or custom parsers to extract data for larger-scale analyses.
+We introduce childes-db, a database-formatted mirror of CHILDES that improves data accessibility and usability by offering novel interfaces, including browsable web applications and an R application programming interface (API).
 Along with a versioned infrastructure that facilitates reproducibility of past analyses, these interfaces lower barriers to analyzing naturalistic parent-child language, allowing for a wider range of researchers in language and cognitive development to easily leverage CHILDES in their work.
@@
 Children's language acquisition is a core puzzle in cognitive science.
-How are the representations underlying the productive use of language learned over children's first years?
+How are the representations underlying the generative use of language learned over children's first years?
 Because a language is comprised of many individual pieces of information -- the names of particular objects, the marker for the past tense, the position of determiners -- language development is fundamentally a question of learning from environmental input, which is the only source of information on these specifics.
@@
 But creating individual transcripts is extremely hard work.
-Beyond gathering the data, even orthographic transcriptions (without phonetic detail) can take ten times the length of the original recording to transcribe [@macwhinney2000].
+Beyond gathering the data, even orthographic transcriptions (without phonetic detail) can take ten times the length of the original recording to transcribe.
 Thus, researchers' desire for data from multiple ages or multiple children can very quickly outstrip their resources.
@@
 Yet there are some challenges working with CHILDES, both for students and for advanced users.
-The CHILDES ecosystem uses a special file format (CHAT), which is stored as plain text but includes structured annotations grouped into tiers (stored on separate lines) so that information about utterances can be stored with accompanying information such as the phonological or morphological structure of the utterance.
+The CHILDES ecosystem uses a special file format (CHAT), which is stored in a plain text format but includes structured annotations grouped into tiers (stored on separate lines) so that information about utterances can be stored with accompanying information such as the phonological or morphological structure of the utterance.
 These files are analyzed using a command-line tool (CLAN) with a host of specific arguments that allow users to count word frequencies, compute statistics (e.g., mean length of utterance, or MLU).
-While this system is flexible and powerful, mastering the CHAT codes and the CLAN tool can be daunting for a beginner or student.
+While this system is flexible and powerful, mastering the CHAT codes and the CLAN tool can be daunting to navigate for a beginner or student.
 These technical barriers decrease the ease of casual exploration by a novice researcher or in a classroom exercise.
@@
 On the opposite end of the spectrum, for data-oriented researchers who are interested in doing large-scale analyses of CHILDES, the current tools are also not ideal.
-CLAN software is an excellent tool for interactive exploration, but -- as a free-standing application -- it can be tricky to build into a processing pipeline written in a modern analytic language such as Python or R.
-Thus, researchers who would like to ingest the entire corpus (or some large subset) into a computational analyses typically write their own parsers of the CHAT format to extract the subset of the data they would like to use [e.g., @redington1998;@yang2013;@meylan2017].
+CLAN software is an excellent tool for interactive exploration, but -- as a free-standing application -- can be tricky to build into a processing pipeline written in a modern analytic platform such as Python or R.
+Thus, researchers who would like to ingest the entire corpus (or some large subset) into a computational analysis typically write their own parsers of the CHAT format to extract the subset of the data they would like to use [e.g., @redington1998;@yang2013;@meylan2017].
 
-The practice of writing custom parsers is problematic for a number of reasons.
+This practice is problematic for a number of reasons.
 Effort is wasted in rewriting the same parser again and again.
 In addition, this process introduces a major source of errors and inconsistencies in data handling due to difficulties dealing with the many special cases in the CHAT standard.
-Further, these parsing scripts are rarely shared, leading to much greater difficulty in reproducing the exact numerical results from papers using CHILDES [see e.g., @meylan2017 for an example].
-In addition, the CHILDES corpus itself is a moving target; computational work using the entire corpus at one time point may include a different set of data than subsequent work (due to new contributions or revisions) and there is no simple way for researchers to document exactly what version of the corpus has been used short of creating a full mirror of the data.
-These factors together lead to a lack of computational reproducibility, a major problem that keeps researchers from verifying or building on published research [@donoho2010;@stodden2016].
+Further, these scripts are rarely shared, leading to much greater difficulty in reproducing the exact numerical results from papers using CHILDES [see e.g., @meylan2017].
+This lack of computational reproducibility is widely recognized as problematic because it does not allow researchers to verify or build on published research [@donoho2010;@stodden2016].
+Finally, the CHILDES corpus itself is a moving target;  computational work using the entire corpus at one time point may include a different set of data than subsequent work (due to new contributions or revisions).
 
@@
 We next describe the `childesr` API, which allows for programmatic access to the data.
-We then present several worked examples of specific uses of the system for research and teaching using both the web app and the API.
+We then present two worked examples of specific uses of the system using both the web app and the API.
+The first examines the frequency of color words in children's input; the second quantifies the growth of lexical diversity across genders.
+We end by presenting some suggestions for the use of the childes-db web front-end in classroom teaching.
 
@@
 <!-- Remind readers why they're reading this: understanding the structure helps you use and understand the tools. SM: Moved this up from database format because it addresses more general issues -->
-As described above, CHILDES is most often approached as a set of distinct CHAT files, which are then parsed by users (often using CLAN). 
-In contrast to this parsing approach, which entails the sequential processing of strings, childes-db treats CHILDES as a set of linked tables, with records corresponding to intuitive abstractions such as words, utterances, and transcripts.
+Unlike the *parsing* approach in CLAN which entails the sequential processing of strings, childes-db treats CHILDES as a set of linked tables, with records corresponding to intuitive abstractions such as words, utterances, and transcripts.
 Users of data analysis languages like R or Julia, libraries like Pandas, or those familiar with Structured Query Language (SQL) will be familiar with operations on tables such as filtering (subsetting), sorting, aggregation (grouping), and joins (merges).
-These operations obviate the need for users to consider the specifics of the CHAT representation -- instead they simply request the entities they need for their research and allow the API to take care of the formatting details.
-We begin by orienting readers to the design of the system via a top-level description and motivation for the design of the database schema, details of the database’s current technical implementation, and our versioning scheme.
-Users primarily interested in accessing the database can skip these details and focus on access through the `childesr` API and the web apps. 
+Although we expect most users to interact with the database through the `childesr` API for the `R` language described below, we begin with a top-level description and motivation for the design of the database schema, details of the database’s current technical implementation, and our versioning scheme.
 <!-- We first present the design and technical approach of childes-db, then describe the interactive analysis tools provided by our web-based frontend.  -->
@@
 
-At its core, childes-db is a database consisting of a set of linked tabular data stores where records correspond to linguistic entities like words, utterances, and sampling units like transcriptions and corpora.
+At its core, childes-db is a database consisting of a set of linked tabular data stores where rows correspond to linguistic entities like words, utterances, and sampling units like transcripts and corpora.
 <!-- what did you mean here? don't rows = records; SM: that's a description of a database in general; we can make a specific assertion about childes-db-->
-The smallest unit of abstraction tracked by the database is a token, treated here as the standard (citation) orthographic form of a word.
+The smallest unit of abstraction tracked by the database is a token, treated here as the citation orthographic form of a word.
 Using the standardized written form of the word facilitates the computation of lexical frequency statistics, for comparison or aggregation across children or time periods. 
@@
 
-Other tables in the database are hierarchical collections built out of tokens -- *utterance*, *transcription*, *corpus*, and *collection* -- which store attributes appropriate for each level of description.
+Other tables in the database comprise a hierarchical system of collections built on top of tokens -- *utterance*, *transcript*, *corpus*, and *collection* -- which stores attributes appropriate for each level of description.
 Every entity includes attributes that link it to all higher-order collections, e.g., an utterance lists the transcript, corpus, and collection to which it belongs.
 An *utterance* contains one or more words and includes fields such as the utterance type (declarative, question, etc.), total number of tokens, and the total number of morphemes (if the morphological structure is available in the original CHAT).
-A *transcription* consists of one or more utterances and includes the date collected, the name of the target child, and age in days (if defined), and the filename from CHILDES.
+
+A *transcript* consists of one or more utterances and includes the date collected, the name of the target child, and age in days (if defined), and the filename from CHILDES.
 A *corpus* consists of one or more transcripts, corresponding to well-known collections like the Brown or Providence corpus.
@@
 
-Participants (e.g., children) are represented separately from the token hierarchy because it is common for the same children to appear in multiple transcripts.
-A participant identifier is associated with every word and utterance, including a name, role, 3-letter CHILDES identifier (CHI = child, MOT = mother, FAT = father, etc.), and the range of ages (or age of corresponding child) for which they are observed.
-For non-child participants (caregivers and others), the record additionally contains an identifier for the corresponding target child to facilitate associating caregivers with children.
+We chose to treat participants separate from the above hierarchy because children can occur across transcripts (e.g., the Brown Corpus).
+A participant identifier is associated with every word and utterance, including a name, role, 3-letter CHILDES identifier (CHI, MOT, FAT, etc.), and the range of ages (or age of corresponding child) for which they are observed.
+For non-child participants (caregivers and others), the record additionally contains an identifier for the corresponding target child to associate caregivers and children.
 
@@
 
-childes-db is implemented as a MySQL database, an industry-standard, open-source relational database server that can be accessed directly from a wide range of programming languages.
+The above tabular datastore is implemented as a MySQL database.
+As an industry-standard, open-source relational database server, the above database can be accessed directly from a wide range of programming languages.
 The childes-db project provides hosted read-only databases for direct access and for `childesr` (described below) as well as compressed .sql exports for local installation.
@@
 In order to import the CHILDES corpora into the MySQL schema described above, it must first be accurately parsed and subsequently vetted to ensure its integrity.
-We parse the XML (eXtensible Markup Language) release of CHILDES hosted by \url{childes.talkbank.org} using the NLTK library in Python [@bird2004].
+We parse the XML (eXtensible Markup Language) release of CHILDES hosted by \url{childes.talkbank.org} using the NLTK library in Python.
+<!-- citation for NLTK -->
 <!-- details of our contribution to the CHILDES corpus module in NLTK---didn't we change it? -->
 Logic implemented in Python converts the linear, multi-tier parse into a tabular format appropriate for childes-db.
-This logic includes decisions regarding what information sources are captured in the current release of the database, and which are left for future development (detailed below).
+This logic includes decisions regarding what information sources are captured in the current release of the database, and which were left for future development (detailed below).
 
-The data imported into childes-db is subject to data integrity checks to ensure that our import of the corpora is accurate and preferable over ad-hoc parsers developed by many individual researchers.
-<!-- MCF: more testing details would be useful -->
-In order to evaluate our success in replicating CLAN parses, we compared unigram counts in our database with those outputted by CLAN, the command-line tool built specifically for analysis of transcripts coded in CHAT.
+<!-- testing details -->
+The data imported into childes-db requires extensive data integrity checks to ensure that our parse of the corpora is accurate and preferable over ad hoc parsers developed by many individual researchers.
+In order to evaluate this, we compared unigram counts in our database with those outputted by CLAN, the command-line tool built specifically for analysis of transcripts coded in CHAT.
 <!-- SM: what were the results of these comparisons?  -->
@@
 The content of CHILDES changes as additional corpora are added or transcriptions are updated; as of the time of writing, these changes are not systematically tracked.
-We introduce a simple versioning system by adding a new complete parse of the current state of CHILDES every six months or as warranted by changes in CHILDES.
+Here we introduce a simple versioning system by adding a new complete parse of the current state of CHILDES every six months.
 By default, users interact with the most recent version of the database available. 
-But to support reproduction of results with previous versions of the database, we continue to host recent versions (up to the last three years / six versions) through our `childesr` API so that researchers can run analyses against specific historical versions of the database.
-<!-- Potentially put some R code to show easy it is to access previous versions? MCF: Nah. -->
+To support reproduction of results with previous versions of the database, we continue to host recent versions (up to the last three years / six versions) through our `childesr` API so that researchers can run analyses against specific historical versions of the database.
+<!-- Potentially put some R code to show easy it is to access previous versions? -->
 For versions more than three years old, we host compressed .sql files that users may download and serve using a local installation of MySQL server.
@@
 - Filled pauses 
-- Reformulations, word revision, and phrase revision, e.g. \textlt what did you\textgt [//] how can you see it ?
+- Reformulations, word revision, and phrase revision, e.g. <what did you> [//] how can you see it ?
 - paralinguistic material, e.g. [=! cries]
 
-We will prioritize the addition of these additional information sources in response to community feedback; extending coverage to the above information sources will not require changes to the existing database schema.
+We will prioritize the addition of these additional information sources in response to community feedback;
+extending coverage to the above information sources will not require changes to the existing database schema.
 
@@
 
-We first discuss the childes-db web apps and then introduce the `childesr` `R` package.
+We first discuss the childes-db web apps and then introduce the `childesr` package.
 
-## Interactive Web Apps
+## Interactive Analysis Tools
+
+<!-- need a good shorthand for the interactive analysis tools: web apps -->
 
@@
 To this end we have created powerful yet easy-to-use interactive web applications that enable users to visualize various dimensions of the CHILDES corpus: frequency counts, mean lengths of utterance, type-token ratios, and more.
-All of this is doable without the requirement of understanding command-line tools or any kind of programming knowledge as had been the case with CLAN.^[The LuCiD toolkit [@chang2017] provides related functionality for a number of common analyses.
-In contrast to those tools, which focus on filling gaps not covered by CLAN -- e.g., the use of n-gram models, incremental sentence generation, and distributional word classification -- our web apps focus on covering the same common tasks as CLAN, but making the outputs into browsable visualizations.]
+All of this is doable without the requirement of understanding command-line tools or any kind of programming knowledge as had been the case with CLAN.
+The only true requirements are a browser and internet connection.
 
-Our web apps are built using Shiny, an R package that enables easy app construction using `R`.
-Underneath the hood, each web app is making calls to our `childesr` API and subsequently plotting the data using the popular R plotting package `ggplot2`.
+The interactive plots (web apps) are built using Shiny, an R package that enables easy building of web applications in R.
+Underneath the hood, each web app is making calls to our `childesr` API and subsequently plotting the data using the R plotting package `ggplot2`.
 A user's only task is to configure exactly what should be plotted through a series of buttons, sliders, and text boxes.
-The user may specify what collection, corpus, child, age range, caregiver, etc., should be included in a given analysis.
-The plot is displayed and updated in real-time, and the underlying data are also available for download alongside the plot to facilitate customization.
-All of these analyses may also be reproduced using the `childesr` package.
+The user may specify what collection, corpus, child, age range, caregiver, etc., should be included in a given analysis (e.g.
+mean length of utterance).
+The plot is displayed and updated in real-time, and if a user were to feel the need to explore the data independently then that data is also available for download alongside the plot.
+All of these analyses may also be reproduced only using the `childesr` package; the web apps were intended for the casual user who seeks to easy extract developmental indices without any technical overhead.
 
-The web apps are intended for the casual user who seeks to easy extract developmental indices quickly and without any technical overhead.
-We next walk through the three web apps provided on our website and describe their functionality in more detail.
+In the following sections we will walk through the three web apps made available on our website and describe their functionality in more detail.
 
@@
 
-The lexical statistics of languaeg input to children have long been an object of study in child language acquisition research.
-Frequency counts of words in particular may provide insight into the cognitive, conceptual, and linguistic experience of a young child [see e.g., @ambridge2015 for review].
-In this web app, inspired by ChildFreq [@baath2015], we provide users the ability to search for any word spoken by a participant in the CHILDES corpora and track the usage of that word by a child or caregiver over time.
-Because of the various toggles available to the user that can subset the data, a user may display frequency curves for a single child in the Brown corpus or all Spanish speaking children, if desired.
-In addition, users can plot frequency curves belonging to caregivers alongside their child for convenient side-by-side comparisons.
-A single word or multiple words may be entered into the input box.
+A child's lexical statistics has long been an object of study in child language acquisition research.
+Namely, the frequency of a word <!-- something important about frequency and development?-->.
+In this web app, we provide users the ability to search for any word or string
+
+The study of frequency counts of individual lexical items as long been the
+
+<!-- sm: intro sentence here for the role of frequency in CLA research  --> 
+The Frequency Counts app allows users to enter a word (or words) in a text box and see the frequency (in parts per million) in which this word was uttered across a child's age for a particular subset of the data, whether it is an individual child in the Brown corpus or all Spanish speaking children and their mothers.
+<!-- sm: "whether..." is awk.
+-->
+The ability to view the caregiver’s usage of a term alongside a child’s assists researchers in better understanding the relationship between input and uptake.
+Users can also plot word trajectories alongside each other (such as “mom” and “dad” or “front” and “back”) for convenient cross-word comparisons.
 
@@
 
-The syntactic complexity and lexial diversity of children's speech are similarly critical metrics for acquisition researchers [@miller1981,@watkins1995].
-There are a number of well-established measures of children's speech that operationalize complexity and diversity.
-These developmental metrics have many applications in speech-language pathology (SLP), where measures outside of the normal range may be indicative of speech, language, or communication disorders.
-
-Several of the most common of these measures are available in the Derived Measures app, which plots these measures across age for a given subset of data, again specified by collection, corpora, children, and speakers.
-As with the Frequency Counts app, caregivers' lexical diversity measures can be plotted alongside children's.
-
-We have currently implemented:
-
-- MLU-w (mean length of utterance in words, a common measure of complexity),^[Because of gaps in available CHILDES morphological data, MLU-w is computed and not MLU-m (mean length of utterance in morphemes).
+<!-- sm: intro sentence here for the role of developmental metrics in CLA research and in downstream applications like SLP--> 
+There are a number of ways to measure the lexical diversity of an individual's productive speech. <!-- sm: different intro; MLU isn't a lexical diversity measure -->
+Some of these measures, including MLU-w (mean length of utterance in words), TTR (type-token ratio), MTLD (measure of textual lexical diversity), and voc-d are available in the Derived Measures app, which plots these measures across age for a given subset of data, again specified by collection, corpora, children, and speakers.
+<!-- sm: MLU is not a measure of lexical diversity-->
+Because of the underlying gaps in available CHILDES morphological data, MLU-w is computed and not MLU-m (mean length of utterance in morphemes).
 <!-- sm: I think we should go ahead and compute this for transcript with MOR tiers?
-Given that this was one of the first uses, I don't think Brian will like the 'underlying gaps' explanation.
-MCF: I agree.
--->
-In English, these two measures are highly correlated. <!-- sm: citation needed -->
-alongside those of children.] 
-- TTR (type-token ratio, a measure of lexical diversity), 
-- MTLD (measure of textual lexical diversity), 
-- HD-D (lexical diversity via the hypergeometric distribution), 
-- and voc-d (a further measure of lexical diversity).  <!-- sm: MLU is not a measure of lexical diversity-->
-<!-- MCF: need citations for voc-d HD-dand MTLD -->
-
-\noindent As with the Frequency Counts app, a user may subset the data as he or she chooses, compare measures between caregivers and children, and aggregate across children from different corpora.
+Given that this was one of the first uses, I don't think Brian will like the 'underlying gaps' explanation-->
+In the English language, these two measures have been shown to be highly correlated. <!-- sm: citation needed -->
+As with the Frequency Counts app, caregiver’s lexical diversity measures can be plotted alongside those of children.
 
@@
 
-Many times a researcher will want to investigate the statistics of a given corpus (i.e the size, number of utterances, number of tokens, etc.) before choosing a target corpus for their analysis.
-This web app is intended to provide a basic overview of various corpora in the CHILDES ecosystem, facilitating familiarity with the different corpora and thus assisting with the overall analytic end-to-end process.
+The population viewer is a tool that helps users understand the relative sizes and temporal extents of various corpora by displaying the total number of utterances or tokens in the available corpora at particular ages.
+In many instances the size of a given corpus may inform researchers’ decisions as to what is the best target dataset for analysis.
 <!-- note that the graphs don't currently include a grouping factor, so adding multiple corpora generates one line, so you can't actually compare two corpora in the same visualization -->
- <!-- cite original ChildFreq paper as inspiration? -->
-
-<!-- figure of freq app -->
 
@@
 Although the interactive analysis tools described above cover some of the most common use cases of CHILDES data, researchers interested in more detailed and flexible analyses will want to interface directly with the data in childes-db.
-Making use of the `R` programming language [@R-base], we provide the `childesr` package.
+Making use of the `R` programming language [@R-base], we provide the `childesr` package to help researchers accomplish this task.
 `R` is an open-source, extensible statistical computing environment that is rapidly growing in popularity across fields and is increasing in use in child language research [e.g.
 @song2015; @norrman2015].
-The `childesr` package abstracts away the details of connecting to and querying  the database.
+The `childesr` package abstracts away the details of connecting to <!-- sm: and querying-->  the database.
 Users can take advantage of the tools developed in the popular `dplyr` package [@R-dplyr], which make manipulating large datasets quick and easy.
@@
 To install, simply type: `install.packages("childesr")`.
-After installation, users have access to data loading functions that can be used to access different parts of the database:
+After installation, users have access to data loading functions that retrieve a variety of subsets <!-- sm: entities?
+variety of subsets sounds non-hierarchical-->  of the data:
 
@@
 
-\noindent Each of these functions can be supplied with arguments that restrict the query to a particular subset of the data (e.g.
-by collection, by corpus, by speaker role, by target child age, etc.) and for a particular version of the data.
-For more detailed documentation, see the package repository (http://github.com/langcog/childesr).
+Each of these functions can be supplied with arguments that restrict the query to a particular subset of the data (e.g.
+by collection, by corpus, by speaker role, by target child age, etc.) For more detailed documentation, see the package repository (http://github.com/langcog/childesr).
 

### 2018-04-11 paper.Rmd
builds, tweaks to refs

--- 2018-04-11 paper.Rmd
@@
 
-keywords          : "keywords"
-wordcount         : "X"
+keywords          : "child language; corpus linguistics; reproducibility; R packages; research software"
+wordcount         : "2925"
 
@@
 library(lme4)
+library(ggthemes)
 knitr::opts_chunk$set(cache = TRUE, echo = FALSE)
@@
 - Filled pauses 
-- Reformulations, word revision, and phrase revision, e.g. \textlt what did you\textgt [//] how can you see it ?
+- Reformulations, word revision, and phrase revision, e.g. \textless what did you\textgreater [//] how can you see it ?
 - paralinguistic material, e.g. [=! cries]
@@
 Frequency counts of words in particular may provide insight into the cognitive, conceptual, and linguistic experience of a young child [see e.g., @ambridge2015 for review].
-In this web app, inspired by ChildFreq [@baath2015], we provide users the ability to search for any word spoken by a participant in the CHILDES corpora and track the usage of that word by a child or caregiver over time.
+In this web app, inspired by ChildFreq [@baath2010], we provide users the ability to search for any word spoken by a participant in the CHILDES corpora and track the usage of that word by a child or caregiver over time.
 Because of the various toggles available to the user that can subset the data, a user may display frequency curves for a single child in the Brown corpus or all Spanish speaking children, if desired.
@@
 
-The syntactic complexity and lexical diversity of children's speech are similarly critical metrics for acquisition researchers [@miller1981,@watkins1995].
+The syntactic complexity and lexical diversity of children's speech are similarly critical metrics for acquisition researchers [@miller1981;@watkins1995].
 There are a number of well-established measures of children's speech that operationalize complexity and diversity.
@@
   facet_wrap(~ color, scales = "free_y") + 
-  theme_classic(base_size = 14) +
+  theme_few() +
   scale_color_manual(values = graph_colors$graph_color) +

### 2018-04-11 paper.Rmd

--- 2018-04-11 paper.Rmd
@@
 At its core, childes-db is a database consisting of a set of linked tabular data stores where records correspond to linguistic entities like words, utterances, and sampling units like transcriptions and corpora.
-<!-- what did you mean here? don't rows = records; SM: that's a description of a database in general; we can make a specific assertion about childes-db-->
 The smallest unit of abstraction tracked by the database is a token, treated here as the standard (citation) orthographic form of a word.
@@
 
-Other tables in the database are hierarchical collections built out of tokens -- *utterance*, *transcription*, *corpus*, and *collection* -- which store attributes appropriate for each level of description.
+Other tables in the database are hierarchical collections built out of tokens -- *utterance*, *transcript*, *corpus*, and *collection* -- which store attributes appropriate for each level of description.
 Every entity includes attributes that link it to all higher-order collections, e.g., an utterance lists the transcript, corpus, and collection to which it belongs.
 An *utterance* contains one or more words and includes fields such as the utterance type (declarative, question, etc.), total number of tokens, and the total number of morphemes (if the morphological structure is available in the original CHAT).
-A *transcription* consists of one or more utterances and includes the date collected, the name of the target child, and age in days (if defined), and the filename from CHILDES.
+A *transcript* consists of one or more utterances and includes the date collected, the name of the target child, and age in days (if defined), and the filename from CHILDES.
 A *corpus* consists of one or more transcripts, corresponding to well-known collections like the Brown or Providence corpus.

### 2018-04-12 paper.Rmd
gender text

@@
 library(directlabels)
-library(koRpus)
+#library(koRpus)
 library(png)
@@
 
-```{r}
+Gender differences in language production have been found in multiple studies, notably in @eriksson2012differences, a large-scale study that spanned ten languages.
+A researcher or student in child language acquisition may now investigate this difference with relative ease using `childesr`.
+By pulling data from the `transcript_by_speaker` table, a user has access to the the multitude of derived linguistic measures that are often used to evaluate a child's linguistic performance.
+In this worked example we walk through a sample analysis that probes the gender differences in language production.
+
+First, we utilize the `childesr` function call `get_speaker_statistics` to pull data relating to the aforementioned derived measures for children and their transcripts.
+We exclusively select children and not their caregivers.
+
+```{r echo=TRUE}
 stats <- get_speaker_statistics(role = "Target_Child")
@@
 
-```{r warning=FALSE}
-lang_threshold <- 575
-age_max <- 4
+This call pulls data from all collections and corpora, including those languages for which there are very sparse data.
+In order to make any substantial inferences from our analysis we begin by filterering languages for which there are a significant number of children.
+We also filter for children under the age of four years.
 
-lang_counts <- stats %>% 
-  filter(target_child_age < age_max * 365) %>%
+```{r echo=TRUE}
+number_of_transcripts_threshold <- 575
+max_age <- 4
+
+included_languages <- stats %>%
+  filter(target_child_age < max_age * 365) %>%
   group_by(language) %>%
+  summarise(n = n()) %>%
+  filter(n > number_of_transcripts_threshold) %>%
+  .$language
+```
 
-  summarise(n = n())
+Our `transcript_by_speaker` table contains multiple derived measures of language productivity and lexical diversity -- here we used MTLD [@mccarthy2005assessment], a measure that is robust to transcript length, as a way of evaluating children's linguistic productivity.
+We start by filtering those children for which a sex was defined, who speak a language in our subset, and are in the appropriate age range.
+We then average their MTLD score across transcripts after having grouped the data by child, age, sex, and language.
 
-included_languages <- filter(lang_counts, n > lang_threshold)$language
-
+```{r echo=TRUE}
 data <- stats %>%
@@
          target_child_sex = factor(target_child_sex, levels = c("male","female"))) %>%
-  filter(age_years < age_max)
+  filter(age_years < max_age)
+```
 
+The data contained in CHILDES is populated from a diverse array of studies that adhered to varying conditions and paradigms.
+This point is particularly salient in our gender analysis due to potential non-independence issues that result from longitudinal studies.
+We chose to fit a model (`mod_data`) of this data that takes these facts into account, and for brevity, is not included here.^[All code and analyses are available at https://github.com/langcog/childes-db-paper] The resultant plot displays the average MTLD scores for various children at different ages, split by gender, with a line corresponding to our model.
+
+<!-- AS: not sure whether I should have included this; I didn't want to overwhelm the reader-->
+
+```{r warning=FALSE}
 mod <- lmer(measure ~ target_child_sex * poly(age_years, 2) - 1 + 
@@
 
-mod_data <- expand.grid(age_years = seq(0,age_max, .1), 
+mod_data <- expand.grid(age_years = seq(0, max_age, .1), 
                                           target_child_sex = c("male","female"),
@@
                             re.form = ~(target_child_sex + age_years | language)) 
+```
 
+```{r echo=TRUE, warning=FALSE}
 ggplot(data, aes(x = age_years, y = measure, 
@@
             aes(x = age_years, y = measure, color = target_child_sex)) +
-  # geom_smooth(span = 1, method = "loess") + 
   facet_wrap(~language) + 
@@
   ggthemes::scale_color_ptol(name = "Sex") + 
-  xlim(0,age_max) + 
+  xlim(0,max_age) + 
   ylim(0,15) + 
@@
 ```
+
+This plot reveals a slight gender difference in linguistic productivity in young children, replicating the results of @eriksson2012differences.
+The goal of this analysis was to showcase an example of the the steps involved in using `childesr` for exploring the CHILDES dataset, as well as some of the potential hazards in working with a diverse set of corpora.
 

### 2018-04-13 paper.Rmd
comments

@@
 library(directlabels)
-#library(koRpus)
 library(png)
@@
 We parse the XML (eXtensible Markup Language) release of CHILDES hosted by \url{childes.talkbank.org} using the NLTK library in Python [@bird2004].
-<!-- details of our contribution to the CHILDES corpus module in NLTK---didn't we change it? -->
+<!-- details of our contribution to the CHILDES corpus module in NLTK---didn't we change it?
+AS: I don't know if I would call it a "contribution" -- we just refactored it to make it more convenient for our purposes -->
 Logic implemented in Python converts the linear, multi-tier parse into a tabular format appropriate for childes-db.
@@
 But to support reproduction of results with previous versions of the database, we continue to host recent versions (up to the last three years / six versions) through our `childesr` API so that researchers can run analyses against specific historical versions of the database.
-<!-- Potentially put some R code to show easy it is to access previous versions? MCF: Nah. -->
 For versions more than three years old, we host compressed .sql files that users may download and serve using a local installation of MySQL server.
-<!-- SM: How easy is it to implement versioning in the Shiny apps? This would show anyone using the web app the availability of the versions-->
-<!-- SM: can people get to these sql files currently on the website?-->
+<!-- SM: How easy is it to implement versioning in the Shiny apps?
+This would show anyone using the web app the availability of the versions AS: app functionality depends on db version in some cases, so nontrivial -->
+<!-- SM: can people get to these sql files currently on the website? AS: yes-->
 

### 2018-04-16 paper.Rmd
added a schema figure

@@
 Deviations from the citation form -- reflecting the deviations from citation forms that are particularly common in the course of language development -- are kept as a separate (possibly null) field associated with each token.
+
+```{r database-schema, fig.pos="h", out.width = "95%", fig.cap="Database schema for childes-db.
+Tokens are linked to superordinate groupings of utterances, transcripts, corpora, and collections (red arrows).
+All tokens and utterances are additionally associated with a participant (blue arrows)."}
+
+png::readPNG(here::here("images/schema.png")) %>% grid::grid.raster()
+```
 

### 2018-04-16 paper.Rmd
comments on several sections

@@
   Access to these data can be complex for novices and difficult to automate for advanced users, however.
-To address these issues, we introduce childes-db, a database-formatted mirror of CHILDES that improves data accessibility and usability by offering novel interfaces, including browsable web applications and an R application programming interface (API).
+  To address these issues, we introduce *childes-db*, a database-formatted mirror of CHILDES that improves data accessibility and usability by offering novel interfaces, including browsable web applications and an R application programming interface (API).
   Along with a versioned infrastructure that facilitates reproducibility of past analyses, these interfaces lower barriers to analyzing naturalistic parent-child language, allowing for a wider range of researchers in language and cognitive development to easily leverage CHILDES in their work.
@@
 Because a language is comprised of many individual pieces of information -- the names of particular objects, the marker for the past tense, the position of determiners -- language development is fundamentally a question of learning from environmental input, which is the only source of information on these specifics.
+<!-- SM: innate knowledge?
+generalzation?
+-->
 Further, children are often reluctant experimental participants; often their most revealing and most sophisticated uses of language emerge in naturalistic play.
+<!-- SM: we could motivate this differently-- it's not just that experiments are hard with young children; it's that many phenomena can only be seen in naturalistic data -->
 Thus, language acquisition research must be grounded in the ecological behavior of children, both their language input and their language production in context.
@@
 The foundational resource for studies of language acquisition is recordings and transcripts of children.
+<!-- SM: this needs to be qualified--- Piaget, Berko, etc.
+experimental and diary studies predate naturalistic trancscripts -->
 Starting with Roger Brown's [-@brown1973] work on Adam, Eve, and Sarah, longitudinal recordings of children's talk (and the talk they hear from their caregivers) have formed the basis for this field.
+<!-- SM: what field?
+child language?
+see above comment -->
 But creating individual transcripts is extremely hard work.
-Beyond gathering the data, even orthographic transcriptions (without phonetic detail) can take ten times the length of the original recording to transcribe [@macwhinney2000].
+Beyond gathering the data, even orthographic transcriptions (without phonetic detail) can take ten times the length of the original recording to transcribe [@macwhinney2000].<!--SM: qualify with "relatively simple" orthographic transcriptions; may need to define `phonetic` for general psych or (non-language) child dev audience -->
 Thus, researchers' desire for data from multiple ages or multiple children can very quickly outstrip their resources.
 
-Established in 1984 to address this issue, the Child Language Data Exchange System (CHILDES) aims to make transcripts and recording of children's language available for researchers via a free resource [@macwhinney2000].
-CHILDES archives tens of thousands of transcripts of children’s speech, making it a critical resource for characterizing children’s early productive language use as well as their language environment.
+Established in 1984 to address this issue, the Child Language Data Exchange System (CHILDES) aims to make transcripts and recording of children's language available for researchers via a free resource [@macwhinney2000].<!--SM: we might want to cite the whole set of CHILDES releases, starting with the earliest one in 1984 -->
+CHILDES archives tens of thousands of transcripts and related media, making it a critical resource for characterizing children’s early productive language use as well as their language environment.
+<!-- SM: added a point about related media b/c Brian MacWhinney thinks this is a central point of CHILDES; tens of thousands of transcripts of children's speech +[across n languages] -->
 As the first major effort to consolidate and share transcripts of child language, CHILDES has been a pioneer in the move to curate and disseminate large-scale behavioral datasets publicly.
 
-Since its inception, a tremendous corpus of research has made use of CHILDES data.
+Since its inception, a tremendous body of research has made use of CHILDES data. <!-- SM: switched corpus -> body to avoid confusion -->
 Individual studies are too numerous to list, but classics include studies of morphological over-regularization [@marcus1992], distributional learning [@redington1998], word segmentation [@goldwater2009], the role of frequency in word learning [@goodman2008] and many others.
-Some studies analyze individual examples in depth [e.g., @snyder2007] while others work with the corpus as a whole [e.g., @meylan2017].
+<!-- SM: to convey its importance, we could say how many citations CHILDES main papers have received and how many citations papers that use CHILDES have received -->
+Some studies analyze individual examples in depth [e.g., @snyder2007], other track multiple child-caregiver dyads [e.g., @meylan2017], and still others use the aggregate properties of child or caregiver speech.
+<!--SM: need a citation for using the aggregate speech -->
 These are just a few of the thousands of studies that take CHILDES corpora -- transcripts, annotations, and associated media -- as starting points.
@@
 Yet there are some challenges working with CHILDES, both for students and for advanced users.
-The CHILDES ecosystem uses a special file format (CHAT), which is stored as plain text but includes structured annotations grouped into tiers (stored on separate lines) so that information about utterances can be stored with accompanying information such as the phonological or morphological structure of the utterance.
-These files are analyzed using a command-line tool (CLAN) with a host of specific arguments that allow users to count word frequencies, compute statistics (e.g., mean length of utterance, or MLU).
-While this system is flexible and powerful, mastering the CHAT codes and the CLAN tool can be daunting for a beginner or student.
+The CHILDES ecosystem typically uses a special file format (CHAT), which is stored as plain text but includes structured annotations grouped into tiers (stored on separate lines) so that information about utterances can be stored with accompanying information such as the phonological, morphological, or syntactic structure of the utterance.
+<!-- added "typically" b/c BMW will note the availability of XML -->
+These files are typically analyzed using a command-line tool (CLAN) with a host of specific arguments that allow users to count word frequencies, compute statistics (e.g., mean length of utterance, or MLU), and execture complex searches against the data.
+While this system is flexible and powerful, mastering the CHAT codes and especially the CLAN tool can be daunting for a beginner or student.
 These technical barriers decrease the ease of casual exploration by a novice researcher or in a classroom exercise.
@@
 On the opposite end of the spectrum, for data-oriented researchers who are interested in doing large-scale analyses of CHILDES, the current tools are also not ideal.
-CLAN software is an excellent tool for interactive exploration, but -- as a free-standing application -- it can be tricky to build into a processing pipeline written in a modern analytic language such as Python or R.
+CLAN software is an excellent tool for interactive exploration, but -- as a free-standing application -- it can be tricky to build into a processing pipeline written in a data analysis platform such as the Python Pandas pacakage or R.
+<!--analytic -> data analysis platform -->
 Thus, researchers who would like to ingest the entire corpus (or some large subset) into a computational analyses typically write their own parsers of the CHAT format to extract the subset of the data they would like to use [e.g., @redington1998;@yang2013;@meylan2017].
@@
 The practice of writing custom parsers is problematic for a number of reasons.
-Effort is wasted in rewriting the same parser again and again.
-In addition, this process introduces a major source of errors and inconsistencies in data handling due to difficulties dealing with the many special cases in the CHAT standard.
-Further, these parsing scripts are rarely shared, leading to much greater difficulty in reproducing the exact numerical results from papers using CHILDES [see e.g., @meylan2017 for an example].
-In addition, the CHILDES corpus itself is a moving target; computational work using the entire corpus at one time point may include a different set of data than subsequent work (due to new contributions or revisions) and there is no simple way for researchers to document exactly what version of the corpus has been used short of creating a full mirror of the data.
+First, effort is wasted in implementing the same features again and again.
+Second, this process introduces a major source of errors and inconsistencies in data handling due to difficulties dealing with the many special cases in the CHAT standard.
+Third, these parsing scripts are rarely shared, leading to much greater difficulty in reproducing the exact numerical results from papers using CHILDES [see e.g., @meylan2017 for an example].
+Fourth, the CHILDES corpus itself is a moving target; computational work using the entire corpus at one time point may include a different set of data than subsequent work (due to newly-contributed corpora or revised annotations of existing ones) and there is no simple way for researchers to document exactly what version of the corpus has been used short of creating a full mirror of the data.
 These factors together lead to a lack of computational reproducibility, a major problem that keeps researchers from verifying or building on published research [@donoho2010;@stodden2016].
@@
 Further, it can be accessed programmatically by advanced researchers, obviating the need to write one-off parsers of the CHAT format.
-The database is versioned for access to previous images, allowing computational reproducibility of particular analyses.
+The database is versioned for access to previous releases, allowing computational reproducibility of particular analyses.
 
@@
 We then present several worked examples of specific uses of the system for research and teaching using both the web app and the API. 
+<!-- a little too much "we" in this final paragraph; needs a final hopefil statement -->
 
@@
 # Using childes-db: Worked Examples
-
+<!-- SM: we should note somewhere that examples all use Tidyverse, so that R users who don't know what %>% don't trip -->
 In this section we give a number of examples of how childes-db can be used in both research and teaching, using both the web apps and the R API.
@@
 
-To demonstrate how one might discover such non-stationary, we take as a case study the frequency with which children hear the color words of English (e.g.
-"blue", "green").
+To demonstrate how one might discover such non-stationary, we take as a case study the frequency with which children hear the color words of English (e.g. "blue", "green").
 Color words tend to be learned relatively late by children, potentially in part due to the abstractness of the meanings to which they refer [see @wagner2013].
@@
 
-To investigate whether the frequency of color words is stationary over development, a user can navigate to the Frequency app, and enter a set of color words into the `Word` selector separated by a comma: here "blue, red, green." Because the question of interest is about the frequency of words in the input (rather than produced by children), the `Speaker` field can be set to reflect this choice.
-In this example we select "Mother." Because children learn most of their basic color words by the age of 5, the age range 1--5 years is a reasonable choice for `Ages to include`.
+To investigate whether the frequency of color words is stationary over development, a user can navigate to the Frequency app, and enter a set of color words into the `Word` selector separated by a comma: here "blue, red, green."
+Because the question of interest is about the frequency of words in the input (rather than produced by children), the `Speaker` field can be set to reflect this choice.
+In this example we select "Mother."
+Because children learn most of their basic color words by the age of 5, the age range 1--5 years is a reasonable choice for `Ages to include`.
 The results of these selections are shown in Figure \ref{fig:color-shiny}.
 We can also create a hyperlink to store these set of choices so that we can share these results with others (or with ourselves in the future) by clicking on the `Bookmark` button in the bottom left corner.
+<!--SM: do we need to update "Bookmark"" to "Share this analysis" ?-->
 
@@
 We also do not have a good sense of the errors of these measurements, are limited to just a few colors at a time before the plot becomes too busy, and cannot combine frequencies across speakers.
+<!-- SM:"busy" seems a bit informal -->
 To perform this analysis in a more compelling and complete way, a user can use the `childesr` interface.
@@
 We then group the data into 6-month age bins, and compute the proportion of tokens that comprise each color for each child in each 6-month bin.
-For comparability with the web app analysis, we convert the proportions to parts per million words.
+For comparability with the web app analysis, these proportions are converted to parts per million words.
 
@@
 Gender differences in language production have been found in multiple studies, notably in @eriksson2012differences, a large-scale study that spanned ten languages.
-A researcher or student in child language acquisition may now investigate this difference with relative ease using `childesr`.
+<!-- SM what research questions specifically?
+many aspects of gender in language...
+probably better to note the range of gender-related effects in early language -->
+A researcher or student in child language acquisition may now investigate this difference with relative ease using `childesr`. <!-- SM: unclear what "this difference" refers to -->
 By pulling data from the `transcript_by_speaker` table, a user has access to the the multitude of derived linguistic measures that are often used to evaluate a child's linguistic performance.
+<!-- SM: linguistic performance -> grammatical development?
+-->
 In this worked example we walk through a sample analysis that probes the gender differences in language production.
+<!-- SM: language production > gender differences in early grammatical development -->
 
 First, we utilize the `childesr` function call `get_speaker_statistics` to pull data relating to the aforementioned derived measures for children and their transcripts.
-We exclusively select children and not their caregivers.
+Notee that we exclusively select the data of children, and exclude their caregivers.
 
@@
 This call pulls data from all collections and corpora, including those languages for which there are very sparse data.
-In order to make any substantial inferences from our analysis we begin by filterering languages for which there are a significant number of children.
+In order to make any substantial inferences from our analysis, we begin by filterering languages for which there are a significant number of children. 
 We also filter for children under the age of four years.
@@
 
-Our `transcript_by_speaker` table contains multiple derived measures of language productivity and lexical diversity -- here we used MTLD [@mccarthy2005assessment], a measure that is robust to transcript length, as a way of evaluating children's linguistic productivity.
+Our `transcript_by_speaker` table contains multiple derived measures of language productivity and lexical diversity -- here we used MTLD [@mccarthy2005assessment], a measure that is more robust to transcript length, as a way of evaluating children's linguistic productivity.
 We start by filtering those children for which a sex was defined, who speak a language in our subset, and are in the appropriate age range.
@@
 
-<!-- AS: not sure whether I should have included this; I didn't want to overwhelm the reader-->
+<!-- AS: not sure whether I should have included this; I didn't want to overwhelm the reader -->
+<!-- SM: need a few more details: what kind of model was it basically? loess, logistic growth curve, etc. -->
 
@@
 The goal of this analysis was to showcase an example of the the steps involved in using `childesr` for exploring the CHILDES dataset, as well as some of the potential hazards in working with a diverse set of corpora.
+
+<!-- SM: not terribly clear what the hazards were: sparsity and autocorrelation? -->
 
@@
 The aim is to reduce time spent getting data out of CHILDES and allow more course time for thinking about interesting questions that could be answered with NLP and child language data.
+<!-- SM: somewhat redundant with the above -->
 Moreover, by writing our API in the R programming language, students will gain access to a general workflow can be integrated into an analysis pipeline that includes state of the art NLP tools.
+<!-- SM: sorry, what tools?
+there's very little NLP support in R b/c it's string handling is slow -->
 
 For example, an instructor could create a programming assignment with the goal of reproducing the key findings in the case studies presented above -- color words or gender. 
-Depending on the students' knowledge of R, the instructor could decide how much of the `childesr` starter code to provide before asking students to generate their own plots and write-ups.
+epending on the students' knowledge of R, the instructor could decide how much of the `childesr` starter code to provide before asking students to generate their own plots and write-ups.
 Moreover, the instructor could take advantage of recent advances in tools for teaching literate programming such as R Markdown [@xie2015dynamic], building in pre-written blocks of executable code that demonstrate key programming concepts for students and including incomplete or blank code blocks that students can use to write their own code.
+<!-- SM: cool, but scope creep?
+-->
 
@@
 We hope that these functionalities broaden the set of users who can easily interact with CHILDES data, leading to future insights into the process of language acquisition. 
+<!-- need a sentence on how it addresses complex use cases -->
 
@@
 For example, currently childes-db operates only on transcript data, without links to the underlying media files; these linkages facilitate deeper analyses of phonology, prosody, social interaction, and many other aspects of acquisition.
+<!-- SM: I thought we did have media links?
+-->
 Further, we have focused on including the most common and widely-used tiers of CHAT annotation into the database first, but our plan is eventually to include the full range of tiers.

### 2018-04-16 paper.Rmd
minor changes in the comments

@@
 <!-- SM: How easy is it to implement versioning in the Shiny apps?
-This would show anyone using the web app the availability of the versions AS: app functionality depends on db version in some cases, so nontrivial -->
-<!-- SM: can people get to these sql files currently on the website? AS: yes-->
+This would show anyone using the web app the availability of the versions AS: app functionality depends on db version in some cases, so nontrivial SM: let's punt-->
 
@@
 - Sparsely annotated tiers, e.g. phonology (`%pho`) and situation (`%sit`)
-<!--- Media links, including timing information AS: we have media timestamps -->
+- Media links
 <!--- Compound word markers, indicated with underscores ("green_eggs_and_ham") AS: we have this -->
@@
 
-We will prioritize the addition of these additional information sources in response to community feedback; extending coverage to the above information sources will not require changes to the existing database schema.
+We will prioritize the addition of these information sources in response to community feedback; extending coverage to the above information sources will not require changes to the existing database schema.
 
@@
 We hope that these functionalities broaden the set of users who can easily interact with CHILDES data, leading to future insights into the process of language acquisition.
-<!-- need a sentence on how it addresses complex use cases -->
 
@@
 For example, currently childes-db operates only on transcript data, without links to the underlying media files; these linkages facilitate deeper analyses of phonology, prosody, social interaction, and many other aspects of acquisition.
-<!-- SM: I thought we did have media links?
--->
 Further, we have focused on including the most common and widely-used tiers of CHAT annotation into the database first, but our plan is eventually to include the full range of tiers.

### 2018-04-17 paper.Rmd

@@
 
-Gender differences in language production have been found in multiple studies, notably in @eriksson2012differences, a large-scale study that spanned ten languages.
-<!-- SM what research questions specifically?
-many aspects of gender in language...
-probably better to note the range of gender-related effects in early language -->
-A researcher or student in child language acquisition may now investigate this difference with relative ease using `childesr`. <!-- SM: unclear what "this difference" refers to -->
-By pulling data from the `transcript_by_speaker` table, a user has access to the the multitude of derived linguistic measures that are often used to evaluate a child's linguistic performance.
-<!-- SM: linguistic performance -> grammatical development?
--->
-In this worked example we walk through a sample analysis that probes the gender differences in language production.
-<!-- SM: language production > gender differences in early grammatical development -->
+Gender differences in language production have been found in multiple studies, notably in @eriksson2012differences, a large-scale study that spanned ten languages and found that females have a higher vocabulary than males in early childhood.
+A researcher or student in child language acquisition may now investigate this difference with relative ease using `childesr`.
+By pulling data from the `transcript_by_speaker` table, a user has access to the the multitude of derived linguistic measures that are often used to evaluate a child's grammatical development.
+In this worked example we walk through a sample analysis that probes the gender differences in early grammatical development.
 
 First, we utilize the `childesr` function call `get_speaker_statistics` to pull data relating to the aforementioned derived measures for children and their transcripts.
-Notee that we exclusively select the data of children, and exclude their caregivers.
+Note that we exclusively select the data of children, and exclude their caregivers.
 
@@
 This call pulls data from all collections and corpora, including those languages for which there are very sparse data.
-In order to make any substantial inferences from our analysis, we begin by filterering languages for which there are a significant number of children. 
+In order to make any substantial inferences from our analysis, we begin by filtering languages for which there are a significant number of children. 
 We also filter for children under the age of four years.
@@
 This point is particularly salient in our gender analysis due to potential non-independence issues that result from longitudinal studies.
-We chose to fit a model (`mod_data`) of this data that takes these facts into account, and for brevity, is not included here.^[All code and analyses are available at https://github.com/langcog/childes-db-paper] The resultant plot displays the average MTLD scores for various children at different ages, split by gender, with a line corresponding to our model.
+For this reason, we fit a linear fixed effects model (`mod_data`, shown in the plot below) to remedy these concerns.^[All code and analyses are available at https://github.com/langcog/childes-db-paper]
+The plot below displays the average MTLD scores for various children at different ages, split by gender, with a line corresponding to our mixed effects model.
 
@@
 
-This plot reveals a slight gender difference in linguistic productivity in young children, replicating the results of @eriksson2012differences.
-The goal of this analysis was to showcase an example of the the steps involved in using `childesr` for exploring the CHILDES dataset, as well as some of the potential hazards in working with a diverse set of corpora.
+This plot reveals a slight gender difference in linguistic productivity in young children (displaying a moderate female advantage), replicating the results of @eriksson2012differences.
+The goal of this analysis was to showcase an example of the the steps involved in using `childesr` for exploring the CHILDES dataset, as well as some of the potential hazards in working with a diverse set of corpora, many of which are longitudinal.
 
-<!-- SM: not terribly clear what the hazards were: sparsity and autocorrelation? -->
+<!-- SM: not terribly clear what the hazards were: sparsity and autocorrelation? AS: mostly the fact that there are longitudinal studies in here-->
 

### 2018-04-18 paper.Rmd
made comments
addressed writing comments through line 246

@@
     email         : "sanchez7@stanford.edu"
-  - name          : "Stephan Meylan*"
+  - name          : "Stephan C. Meylan*"
     affiliation   : "2"
@@
 abstract: |
-  The Child Language Data Exchange System (CHILDES) has played a critical role in research on child language development.
+  The Child Language Data Exchange System (CHILDES) has played a critical role in research on child language development, particularly in characterizing the early language learning environment.
   Access to these data can be complex for novices and difficult to automate for advanced users, however.
@@
 
-Children's language acquisition is a core puzzle in cognitive science.
-How are the representations underlying the productive use of language learned over children's first years?
-Because a language is comprised of many individual pieces of information -- the names of particular objects, the marker for the past tense, the position of determiners -- language development is fundamentally a question of learning from environmental input, which is the only source of information on these specifics.
-<!-- SM: innate knowledge?
-generalzation?
--->
-Further, children are often reluctant experimental participants; often their most revealing and most sophisticated uses of language emerge in naturalistic play.
-<!-- SM: we could motivate this differently-- it's not just that experiments are hard with young children; it's that many phenomena can only be seen in naturalistic data -->
-Thus, language acquisition research must be grounded in the ecological behavior of children, both their language input and their language production in context.
+How children acquire language is one of the core puzzles in cognitive science, developmental psychology, and linguistics. 
+What are the representation that children learn about language, and how to they emerge from the interaction of learning mechanisms and environmental input?
+Developing facility with language requires learning a great many interlocking components -- meaningful distinctions between sounds (phonology), names of particular objects and actions (word learning), meaningful sub-word structure (morphology), rules for how to organize words together (syntax), and context-dependent and context-independent aspects of meaning (semantics and pragmatics).
+Key to learning all of these systems is the contribution of the child's input -- exposure to linguistic and non-linguistic data -- in the early environment.
+While in-lab experiments can shed light on linguistic knowledge and some of the implicated learning mechanisms, characterizing this early environment requires additional research methods and resources.
 
-The foundational resource for studies of language acquisition is recordings and transcripts of children.
-<!-- SM: this needs to be qualified--- Piaget, Berko, etc.
-experimental and diary studies predate naturalistic trancscripts -->
-Starting with Roger Brown's [-@brown1973] work on Adam, Eve, and Sarah, longitudinal recordings of children's talk (and the talk they hear from their caregivers) have formed the basis for this field.
-<!-- SM: what field?
-child language?
-see above comment -->
-But creating individual transcripts is extremely hard work.
-Beyond gathering the data, even orthographic transcriptions (without phonetic detail) can take ten times the length of the original recording to transcribe [@macwhinney2000].<!--SM: qualify with "relatively simple" orthographic transcriptions; may need to define `phonetic` for general psych or (non-language) child dev audience -->
-Thus, researchers' desire for data from multiple ages or multiple children can very quickly outstrip their resources.
+One of the key methods that has emerged to address this gap is the collection and annotation of speech to and by children, often in the context of the home. 
+Starting with Roger Brown's [-@brown1973] work on Adam, Eve, and Sarah, audio recordings -- and more recently augmented with video recordings -- have been augmented with rich, searchable annotations that have allowed researchers to address a number of questions regarding the language learning environment.
+Focusing on language learning in naturalistic contexts also reveals that children have in many cases productive and receptive abilities exceeding those revealed in experimental contexts; owing to challenges in lab-based experiments with young children, often their most revealing and sophisticated uses of language emerge in naturalistic play.
 
-Established in 1984 to address this issue, the Child Language Data Exchange System (CHILDES) aims to make transcripts and recording of children's language available for researchers via a free resource [@macwhinney2000].<!--SM: we might want to cite the whole set of CHILDES releases, starting with the earliest one in 1984 -->
-CHILDES archives tens of thousands of transcripts and related media, making it a critical resource for characterizing children’s early productive language use as well as their language environment.
-<!-- SM: added a point about related media b/c Brian MacWhinney thinks this is a central point of CHILDES; tens of thousands of transcripts of children's speech +[across n languages] -->
+While corpora of early language acquisition are extremely useful, creating such corpora requires significant resources. 
+Collecting and transcribing audio and video is costly and extremely time consuming  -- even orthographic transcription (i.e., transcriptions with minimal phonetic detail) can take ten times the length of the original recording to transcribe [@macwhinney2000].
+Automated, machine learning -based methods like automatic speech recognition (ASR) have provided only modest gains in efficiency, in that they are limited by the less-than-ideal acoustic properties of home recordings and the poor fit of language models built on adult-directed, adult-produced language samples for characterizing child-directed or child-produced speech.
+Thus, researchers' desire for sufficient sample sizes and coverage in analyses of child language corpora can very quickly outstrip their resources. 
+<!-- data pooling is necessary) -->
+
+Established in 1984 to address this issue, the Child Language Data Exchange System (CHILDES) aims to make transcripts and recording of children's language available for researchers as free, public datasets [@macwhinney1985; @macwhinney2000; @macwhinney2014].
+CHILDES archives tens of thousands of transcripts and associated media across 20+ languages, making it a critical resource for characterizing children’s early productive language use as well as their language environment.
 As the first major effort to consolidate and share transcripts of child language, CHILDES has been a pioneer in the move to curate and disseminate large-scale behavioral datasets publicly.
 
-Since its inception, a tremendous body of research has made use of CHILDES data. <!-- SM: switched corpus -> body to avoid confusion -->
-Individual studies are too numerous to list, but classics include studies of morphological over-regularization [@marcus1992], distributional learning [@redington1998], word segmentation [@goldwater2009], the role of frequency in word learning [@goodman2008] and many others.
-<!-- SM: to convey its importance, we could say how many citations CHILDES main papers have received and how many citations papers that use CHILDES have received -->
-Some studies analyze individual examples in depth [e.g., @snyder2007], other track multiple child-caregiver dyads [e.g., @meylan2017], and still others use the aggregate properties of child or caregiver speech.
-<!--SM: need a citation for using the aggregate speech -->
-These are just a few of the thousands of studies that take CHILDES corpora -- transcripts, annotations, and associated media -- as starting points.
+Since its inception, a tremendous body of research has made use of CHILDES data.
+Individual studies are too numerous to list, but classics include studies of morphological over-regularization [@marcus1992], distributional learning [@redington1998], word segmentation [@goldwater2009], the role of frequency in word learning [@goodman2008], and many others.
+Some studies analyze individual examples in depth [e.g., @snyder2007], other track multiple child-caregiver dyads [e.g., @meylan2017], and still others use the aggregate properties of child or caregiver speech [e.g., @redingtonEtAl1998; @montagEtAl2015].
+These are just a few of the thousands of studies that make use of CHILDES corpora.
 
-Yet there are some challenges working with CHILDES, both for students and for advanced users.
-The CHILDES ecosystem typically uses a special file format (CHAT), which is stored as plain text but includes structured annotations grouped into tiers (stored on separate lines) so that information about utterances can be stored with accompanying information such as the phonological, morphological, or syntactic structure of the utterance.
-<!-- added "typically" b/c BMW will note the availability of XML -->
-These files are typically analyzed using a command-line tool (CLAN) with a host of specific arguments that allow users to count word frequencies, compute statistics (e.g., mean length of utterance, or MLU), and execture complex searches against the data.
-While this system is flexible and powerful, mastering the CHAT codes and especially the CLAN tool can be daunting for a beginner or student.
-These technical barriers decrease the ease of casual exploration by a novice researcher or in a classroom exercise.
+Nonetheless, there are some outstanding challenges working with CHILDES, both for students and for advanced users.
+The CHILDES ecosystem typically uses a specialized file format (CHAT), which is stored as plain text but includes structured annotations grouped into tiers (stored on separate lines) so that information about utterances can be stored with accompanying information such as the phonological, morphological, or syntactic structure of the utterance.
+These files are usually analyzed using a command-line tool (CLAN), which has a number of specific tools that allow users to count word frequencies, compute statistics (e.g., mean length of utterance, or MLU), and execute complex searches against the data.
+While this system is flexible and powerful, mastering the CHAT codes and especially the CLAN tool with its many tools and flags can be daunting for a beginner or student.
+These technical barriers decrease the ease of exploration by a novice researcher or in a classroom exercise.
 
 On the opposite end of the spectrum, for data-oriented researchers who are interested in doing large-scale analyses of CHILDES, the current tools are also not ideal.
-CLAN software is an excellent tool for interactive exploration, but -- as a free-standing application -- it can be tricky to build into a processing pipeline written in a data analysis platform such as the Python Pandas pacakage or R.
-<!--analytic -> data analysis platform -->
+CLAN software is an excellent tool for interactive exploration, but -- as a free-standing application -- it can be tricky to build into a processing pipeline written in Python or R.
 Thus, researchers who would like to ingest the entire corpus (or some large subset) into a computational analyses typically write their own parsers of the CHAT format to extract the subset of the data they would like to use [e.g., @redington1998;@yang2013;@meylan2017].
@@
 Second, this process introduces a major source of errors and inconsistencies in data handling due to difficulties dealing with the many special cases in the CHAT standard.
-Third, these parsing scripts are rarely shared, leading to much greater difficulty in reproducing the exact numerical results from papers using CHILDES [see e.g., @meylan2017 for an example].
-Fourth, the CHILDES corpus itself is a moving target; computational work using the entire corpus at one time point may include a different set of data than subsequent work (due to newly-contributed corpora or revised annotations of existing ones) and there is no simple way for researchers to document exactly what version of the corpus has been used short of creating a full mirror of the data.
+Third, these parsing scripts are rarely shared or break with subsequent revisions to the dataset, leading to much greater difficulty in reproducing the exact numerical results from papers using CHILDES [see e.g., @meylan2017 for an example].
+Fourth, the CHILDES corpus itself is a moving target; computational work using the entire corpus at one time point may include a different set of data than subsequent work due to newly-contributed corpora or revised annotations of existing ones; currently, there is no simple way for researchers to document exactly what version of the corpus has been used short of creating a full mirror of the data.
 These factors together lead to a lack of computational reproducibility, a major problem that keeps researchers from verifying or building on published research [@donoho2010;@stodden2016].
@@
 Our system, childes-db, is a database-formatted mirror of CHILDES that allows programmatic access through an application programming interface (API). 
-This infrastructure allows the creation of web applications for browsing and visualizing the data easily, facilitating classroom use of the dataset. 
+This infrastructure allows the creation of web applications for browsing and easily visualizing the data, facilitating classroom use of the dataset. 
 Further, it can be accessed programmatically by advanced researchers, obviating the need to write one-off parsers of the CHAT format.
@@
 We begin by describing the architecture of childes-db and the web applications that we provide.
-We next describe the `childesr` API, which allows for programmatic access to the data.
-We then present several worked examples of specific uses of the system for research and teaching using both the web app and the API. 
-<!-- a little too much "we" in this final paragraph; needs a final hopefil statement -->
+Next, we describe the `childesr` API, which provides a set of R functions for programmatic access to the data.
+We conclude by presenting several worked examples of specific uses of the system--both web apps and the R API--for both research and teaching.
 
@@
 
-<!-- Remind readers why they're reading this: understanding the structure helps you use and understand the tools. SM: Moved this up from database format because it addresses more general issues -->
 As described above, CHILDES is most often approached as a set of distinct CHAT files, which are then parsed by users (often using CLAN). 
@@
 Users primarily interested in accessing the database can skip these details and focus on access through the `childesr` API and the web apps. 
+<!-- SM: not sure if we want to have readers skip this; there may be some conceptual dependencies here -->
 <!-- We first present the design and technical approach of childes-db, then describe the interactive analysis tools provided by our web-based frontend.  -->
@@
 
-Other tables in the database are hierarchical collections built out of tokens -- *utterance*, *transcript*, *corpus*, and *collection* -- which store attributes appropriate for each level of description.
+Manty of the other tables in the database are hierarchical collections built out of tokens -- *utterance*, *transcript*, *corpus*, and *collection* -- which store attributes appropriate for each level of description.
 Every entity includes attributes that link it to all higher-order collections, e.g., an utterance lists the transcript, corpus, and collection to which it belongs.
@@
 
-Participants (e.g., children) are represented separately from the token hierarchy because it is common for the same children to appear in multiple transcripts.
+Participants (e.g., children and caregivers) are represented separately from the token hierarchy because it is common for the same children to appear in multiple transcripts.
 A participant identifier is associated with every word and utterance, including a name, role, 3-letter CHILDES identifier (CHI = child, MOT = mother, FAT = father, etc.), and the range of ages (or age of corresponding child) for which they are observed.
-For non-child participants (caregivers and others), the record additionally contains an identifier for the corresponding target child to facilitate associating caregivers with children.
+For non-child participants (caregivers and others), the record additionally contains an identifier for the corresponding target child, such that data correpsonding to children and their caregivers can be easily associated.
 
@@
 childes-db is implemented as a MySQL database, an industry-standard, open-source relational database server that can be accessed directly from a wide range of programming languages.
-The childes-db project provides hosted read-only databases for direct access and for `childesr` (described below) as well as compressed .sql exports for local installation.
-While the former is appropriate for most users, local installation can provide performance gains (allowing a user to access the database on their local machine or on their network) as well as allowing users to store derived information in the same database.
+The childes-db project provides hosted, read-only databases for direct access and for `childesr` (described below) as well as compressed .sql exports for local installation.
+While the former is appropriate for most users, local installation can provide performance gains (allowing a user to access the database on their machine or on their local network) as well as allowing users to store derived information in the same database.
 
@@
 By default, users interact with the most recent version of the database available. 
-But to support reproduction of results with previous versions of the database, we continue to host recent versions (up to the last three years / six versions) through our `childesr` API so that researchers can run analyses against specific historical versions of the database.
+To support reproduction of results with previous versions of the database, we continue to host recent versions (up to the last three years / six versions) through our `childesr` API so that researchers can run analyses against specific historical versions of the database.
 For versions more than three years old, we host compressed .sql files that users may download and serve using a local installation of MySQL server.
@@
 
-The current implementation of childes-db emphasizes the computation of lexical statistics, and consequently focuses on reproducing the words, utterances, and speaker information in CHILDES.
+The current implementation of childes-db emphasizes the computation of lexical statistics, and consequently focuses on reproducing the words, utterances, and speaker information in CHILDES transcripts.
 We have yet to parse and make available the following information:
@@
 Our web apps are built using Shiny, an R package that enables easy app construction using `R`.
-Underneath the hood, each web app is making calls to our `childesr` API and subsequently plotting the data using the popular R plotting package `ggplot2`.
+Underneath the hood, each web app is making calls to our `childesr` API and subsequently plots the data using the popular R plotting package `ggplot2`.
 A user's only task is to configure exactly what should be plotted through a series of buttons, sliders, and text boxes.
@@
 
-We have currently implemented:
+We have currently implemented the following measures:
 
@@
 alongside those of children.] 
-- TTR (type-token ratio, a measure of lexical diversity), 
+- TTR [type-token ratio, a measure of lexical diversity; @templin1957], 
 - MTLD (measure of textual lexical diversity), 
@@
 For example, an instructor could create a programming assignment with the goal of reproducing the key findings in the case studies presented above -- color words or gender. 
-epending on the students' knowledge of R, the instructor could decide how much of the `childesr` starter code to provide before asking students to generate their own plots and write-ups.
+Depending on the students' knowledge of R, the instructor could decide how much of the `childesr` starter code to provide before asking students to generate their own plots and write-ups.
 Moreover, the instructor could take advantage of recent advances in tools for teaching literate programming such as R Markdown [@xie2015dynamic], building in pre-written blocks of executable code that demonstrate key programming concepts for students and including incomplete or blank code blocks that students can use to write their own code.

### 2018-04-18 paper.Rmd
test details

@@
 
-Manty of the other tables in the database are hierarchical collections built out of tokens -- *utterance*, *transcript*, *corpus*, and *collection* -- which store attributes appropriate for each level of description.
+Many of the other tables in the database are hierarchical collections built out of tokens -- *utterance*, *transcript*, *corpus*, and *collection* -- which store attributes appropriate for each level of description.
 Every entity includes attributes that link it to all higher-order collections, e.g., an utterance lists the transcript, corpus, and collection to which it belongs.
@@
 In order to evaluate our success in replicating CLAN parses, we compared unigram counts in our database with those outputted by CLAN, the command-line tool built specifically for analysis of transcripts coded in CHAT.
+In particular, the CLAN commands FREQ and MLU were used to compare total token counts and mean lengths of utterance for every speaker in every transcript; these numbers were compared to our own using the Pearson correlation coefficient.
+The results of the comparison were .99 and .98 for the unigram count and mlu data, respectively, indicating a reliable parse on our end.
 <!-- SM: what were the results of these comparisons?  -->
@@
 MCF: I agree.
--->
+AS: MLU-m still in development, but we should add it here anyway-->
 In English, these two measures are highly correlated. <!-- sm: citation needed -->
@@
 - HD-D (lexical diversity via the hypergeometric distribution), 
-- and voc-d (a further measure of lexical diversity).  <!-- sm: MLU is not a measure of lexical diversity-->
+- and voc-d (a further measure of lexical diversity).  <!-- AS: let's exclude voc-d? and say HDD is idealized version of voc-d -->
+
+<!-- sm: MLU is not a measure of lexical diversity-->
 <!-- MCF: need citations for voc-d HD-dand MTLD -->
@@
 \noindent As with the Frequency Counts app, a user may subset the data as he or she chooses, compare measures between caregivers and children, and aggregate across children from different corpora.
+
+<!-- AS: screenshots for individual apps? -->
 

### 2018-04-18 paper.Rmd
shiny screenshots

@@
 
+```{r frequency-app2, fig.pos="h", out.width = "95%", fig.cap="Frequency Counts."}
+
+png::readPNG(here::here("images/freq_app.png")) %>% grid::grid.raster()
+```
+
 ### Derived Measures
@@
 
-<!-- AS: screenshots for individual apps? -->
+```{r derived-measures-app2, fig.pos="h", out.width = "95%", fig.cap="Derived Measures."}
+
+png::readPNG(here::here("images/derived_measures_app.png")) %>% grid::grid.raster()
+```
 
@@
 
-<!-- figure of freq app -->
+```{r population-viewer-app, fig.pos="h", out.width = "95%", fig.cap="Population Viewer."}
+
+png::readPNG(here::here("images/population_viewer_app.png")) %>% grid::grid.raster()
+```
 

### 2018-04-18 paper.Rmd
mlu-m

@@
 How children acquire language is one of the core puzzles in cognitive science, developmental psychology, and linguistics. 
-What are the representation that children learn about language, and how to they emerge from the interaction of learning mechanisms and environmental input?
+What are the representations that children learn about language, and how do they emerge from the interaction of learning mechanisms and environmental input?
 Developing facility with language requires learning a great many interlocking components -- meaningful distinctions between sounds (phonology), names of particular objects and actions (word learning), meaningful sub-word structure (morphology), rules for how to organize words together (syntax), and context-dependent and context-independent aspects of meaning (semantics and pragmatics).
@@
 One of the key methods that has emerged to address this gap is the collection and annotation of speech to and by children, often in the context of the home. 
-Starting with Roger Brown's [-@brown1973] work on Adam, Eve, and Sarah, audio recordings -- and more recently augmented with video recordings -- have been augmented with rich, searchable annotations that have allowed researchers to address a number of questions regarding the language learning environment.
+Starting with Roger Brown's [-@brown1973] work on Adam, Eve, and Sarah, audio recordings -- and more recently video recordings -- have been augmented with rich, searchable annotations that have allowed researchers to address a number of questions regarding the language learning environment.
 Focusing on language learning in naturalistic contexts also reveals that children have in many cases productive and receptive abilities exceeding those revealed in experimental contexts; owing to challenges in lab-based experiments with young children, often their most revealing and sophisticated uses of language emerge in naturalistic play.
@@
 
-- MLU-w (mean length of utterance in words, a common measure of complexity),^[Because of gaps in available CHILDES morphological data, MLU-w is computed and not MLU-m (mean length of utterance in morphemes).
-<!-- sm: I think we should go ahead and compute this for transcript with MOR tiers?
-Given that this was one of the first uses, I don't think Brian will like the 'underlying gaps' explanation.
-MCF: I agree.
-AS: MLU-m still in development, but we should add it here anyway-->
-In English, these two measures are highly correlated. <!-- sm: citation needed -->
-alongside those of children.] 
+- MLU-w (mean length of utterance in words),
+- MLU-m (mean length of utterance in morphemes),
 - TTR [type-token ratio, a measure of lexical diversity; @templin1957], 
 - MTLD (measure of textual lexical diversity), 
-- HD-D (lexical diversity via the hypergeometric distribution), 
-- and voc-d (a further measure of lexical diversity).  <!-- AS: let's exclude voc-d? and say HDD is idealized version of voc-d -->
-
-<!-- sm: MLU is not a measure of lexical diversity-->
+- HD-D (lexical diversity via the hypergeometric distribution)
+<!-- AS: removed voc-d as HDD is idealized version of voc-d -->
 <!-- MCF: need citations for voc-d HD-dand MTLD -->
@@
 <!-- note that the graphs don't currently include a grouping factor, so adding multiple corpora generates one line, so you can't actually compare two corpora in the same visualization -->
- <!-- cite original ChildFreq paper as inspiration? -->
 
@@
 The results of these selections are shown in Figure \ref{fig:color-shiny}.
-We can also create a hyperlink to store these set of choices so that we can share these results with others (or with ourselves in the future) by clicking on the `Bookmark` button in the bottom left corner.
-<!--SM: do we need to update "Bookmark"" to "Share this analysis" ?-->
+We can also create a hyperlink to store these set of choices so that we can share these results with others (or with ourselves in the future) by clicking on the `Share Analysis` button in the bottom left corner.
 

### 2018-04-18 paper.Rmd
other edits

@@
 This point is particularly salient in our gender analysis due to potential non-independence issues that result from longitudinal studies.
-For this reason, we fit a linear fixed effects model (`mod_data`, shown in the plot below) to remedy these concerns.^[All code and analyses are available at https://github.com/langcog/childes-db-paper]
+For this reason, we fit a linear mixed effects model (`mod_data`, shown in the plot below) to remedy these concerns.^[All code and analyses are available at https://github.com/langcog/childes-db-paper]
 The plot below displays the average MTLD scores for various children at different ages, split by gender, with a line corresponding to our mixed effects model.
 
-<!-- AS: not sure whether I should have included this; I didn't want to overwhelm the reader -->
-<!-- SM: need a few more details: what kind of model was it basically? loess, logistic growth curve, etc. -->
+<!-- AS: is this sufficient for mixed effects model explanation? -->
 
@@
 This plot reveals a slight gender difference in linguistic productivity in young children (displaying a moderate female advantage), replicating the results of @eriksson2012differences.
-The goal of this analysis was to showcase an example of the the steps involved in using `childesr` for exploring the CHILDES dataset, as well as some of the potential hazards in working with a diverse set of corpora, many of which are longitudinal.
+The goal of this analysis was to showcase an example of the the steps involved in using `childesr` for exploring the CHILDES dataset, as well as some of the potential hazards in working with a diverse set of corpora that were collectd through various experimental paradigms.
 

### 2018-04-18 paper.Rmd
continued text revision

--- 2018-04-18 paper.Rmd
@@
 - TTR [type-token ratio, a measure of lexical diversity; @templin1957], 
-- MTLD (measure of textual lexical diversity), 
-- HD-D (lexical diversity via the hypergeometric distribution)
+- MTLD [measure of textual lexical diversity; @malvernRichards1997], 
+- HD-D [lexical diversity via the hypergeometric distribution; @mccarthy2010])
 <!-- AS: removed voc-d as HDD is idealized version of voc-d -->
@@
 
-\noindent As with the Frequency Counts app, a user may subset the data as he or she chooses, compare measures between caregivers and children, and aggregate across children from different corpora.
+\noindent As with the Frequency Counts app, a user may subset the data as they choose, compare measures between caregivers and children, and aggregate across children from different corpora.
 
@@
 Many times a researcher will want to investigate the statistics of a given corpus (i.e the size, number of utterances, number of tokens, etc.) before choosing a target corpus for their analysis.
-This web app is intended to provide a basic overview of various corpora in the CHILDES ecosystem, facilitating familiarity with the different corpora and thus assisting with the overall analytic end-to-end process.
-<!-- note that the graphs don't currently include a grouping factor, so adding multiple corpora generates one line, so you can't actually compare two corpora in the same visualization -->
+This web app is intended to provide a basic overview regarding the scale and temporal extent of various corpora in CHILDES, as well as giving researchers insight into the aggregate characteristics of CHILDES.
+For example, examining the aggregate statistics reveals that coverage peaks at around 30 months.
+<!-- SM: note that the graphs don't currently include a grouping factor, so adding multiple corpora generates one line, so you can't actually compare two corpora in the same visualization -->
 
@@
 The `childesr` package is easily installed via CRAN, the comprehensive R archive network.
-To install, simply type: `install.packages("childesr")`.
-After installation, users have access to data loading functions that can be used to access different parts of the database:
+To install, simply type: `install.packages("childesr")`. After installation, users have access to functions that can be used to retrieve tabular data from the database:
 
@@
 
-\noindent Each of these functions can be supplied with arguments that restrict the query to a particular subset of the data (e.g.
+\noindent Each of these functions take arguments that restrict the query to a particular subset of the data (e.g.
 by collection, by corpus, by speaker role, by target child age, etc.) and for a particular version of the data.
@@
 # Using childes-db: Worked Examples
-<!-- SM: we should note somewhere that examples all use Tidyverse, so that R users who don't know what %>% don't trip -->
+
 In this section we give a number of examples of how childes-db can be used in both research and teaching, using both the web apps and the R API.
@@
 
-One common use of the CHILDES corpora is estimating the frequency with which children hear different words.
+One common use of CHILDES is to estimate the frequency with which children hear different words.
 These frequency estimates are used both in the development of theory [e.g., frequent words are learned earlier; @goodman2008], and in the construction of age-appropriate experimental stimuli.
@@
 However, within the set of color words, the frequency with which these words are heard predicts a significant fraction of the variance in their order of acquisition [@yurovsky2015].
-But are these frequencies stationary -- e.g.
-do children hear "blue" as often at 12 months as they do at 24 months?
+But are these frequencies stationary -- e.g. do children hear "blue" as often at 12 months as they do at 24 months?
 We answer this question in two ways -- first using the web apps, and then using the `childesr` package.
@@
 
-From this figure, it seems likely that children hear "blue" more frequently early in development, but the trajectories of the "red" and "green" are less clear.
-We also do not have a good sense of the errors of these measurements, are limited to just a few colors at a time before the plot becomes too busy, and cannot combine frequencies across speakers.
-<!-- SM:"busy" seems a bit informal -->
+From this figure, it seems likely that children hear "blue" more frequently early in development, but the trajectories of "red" and "green" are less clear.
+We also do not have a good sense of the errors of these measurements, are limited to just a few colors at a time before the plot becomes too crowded, and cannot combine frequencies across speakers.
 To perform this analysis in a more compelling and complete way, a user can use the `childesr` interface.
@@
 
-We can analyze these learning trajectories using childesr by breaking the process into five steps: (1) define our words of interest, (2) find the frequencies with with which children hear these words, (3) find the proportion of the *total words* children hear that these frequencies account for, (4) aggregate across transcripts and children to determine the error in our estimates of these proportions, and (5) plot the results.
+We can analyze these learning trajectories using `childesr` by breaking the process into five steps: (1) define our words of interest, (2) find the frequencies with with which children hear these words, (3) find the proportion of the *total words* children hear that these frequencies account for, (4) aggregate across transcripts and children to determine the error in our estimates of these proportions, and (5) plot the results.
 
@@
 
-Gender differences in language production have been found in multiple studies, notably in @eriksson2012differences, a large-scale study that spanned ten languages and found that females have a higher vocabulary than males in early childhood.
-A researcher or student in child language acquisition may now investigate this difference with relative ease using `childesr`.
-By pulling data from the `transcript_by_speaker` table, a user has access to the the multitude of derived linguistic measures that are often used to evaluate a child's grammatical development.
+Gender has long been known to be an important factor for early vocabulary growth, with girls learning more words earlier than boys [@huttenlocher1991].
+More recently, this finding has been shown to be robust in ten languages [@eriksson2012differences].
+Important questions remain regarding the relative contribution of differences in the language input to boys and girls vs. maturational differences in neural development.
+A researcher or student in child language acquisition may now further investigate these differences with relative ease using `childesr`.
+By pulling data from the `transcript_by_speaker` table, a user has access to the multitude of derived linguistic measures that are often used to evaluate a child's grammatical development.
 In this worked example we walk through a sample analysis that probes the gender differences in early grammatical development.
 
-First, we utilize the `childesr` function call `get_speaker_statistics` to pull data relating to the aforementioned derived measures for children and their transcripts.
+First, we use the `childesr` function call `get_speaker_statistics` to pull data relating to the aforementioned derived measures for children and their transcripts.
 Note that we exclusively select the data of children, and exclude their caregivers.
@@
 
-This call pulls data from all collections and corpora, including those languages for which there are very sparse data.
-In order to make any substantial inferences from our analysis, we begin by filtering languages for which there are a significant number of children. 
-We also filter for children under the age of four years.
+This call pulls data from all collections and corpora, including several languages for which there are very sparse data.
+In order to make any substantial inferences from our analysis, we begin by filtering out languages with an insufficient number of children. 
+We also constrain our analysis to children under the age of four years.
 
@@
 Our `transcript_by_speaker` table contains multiple derived measures of language productivity and lexical diversity -- here we used MTLD [@mccarthy2005assessment], a measure that is more robust to transcript length, as a way of evaluating children's linguistic productivity.
-We start by filtering those children for which a sex was defined, who speak a language in our subset, and are in the appropriate age range.
+We start by filtering to those children for which a sex was defined, who speak a language in our subset, and are in the appropriate age range.
 We then average their MTLD score across transcripts after having grouped the data by child, age, sex, and language.
@@
 
-The data contained in CHILDES is populated from a diverse array of studies that adhered to varying conditions and paradigms.
-This point is particularly salient in our gender analysis due to potential non-independence issues that result from longitudinal studies.
-For this reason, we fit a linear fixed effects model (`mod_data`, shown in the plot below) to remedy these concerns.^[All code and analyses are available at https://github.com/langcog/childes-db-paper]
-The plot below displays the average MTLD scores for various children at different ages, split by gender, with a line corresponding to our mixed effects model.
+The data contained in CHILDES is populated from a diverse array of studies refelcting varying circumstances of data collection.
+This point is particularly salient in our gender analysis due to potential non-independence issues that may emerge from the inclusion of many transcripts from longitudinal studies.
+To rememdy these concerns, we fit a linear mixed effects model with gender * age (treated as a quadratic predictor) as fixed effects, child identity as a random intercept, and age + gender by language as a random slope.^[All code and analyses are available at https://github.com/langcog/childes-db-paper]
+<!--  mod <- lmer(measure ~ target_child_sex * poly(age_years, 2) - 1 + 
+                    (1 | target_child_id) +
+                    (target_child_sex + age_years | language), 
+  data = data) -->
+<!-- SM: not sure I got the random effects structure right -->
+The plot below displays the average MTLD scores for various children at different ages, split by gender, with a line corresponding to the prediction of our fit mixed effects model.
 
@@
 
-This plot reveals a slight gender difference in linguistic productivity in young children (displaying a moderate female advantage), replicating the results of @eriksson2012differences.
-The goal of this analysis was to showcase an example of the the steps involved in using `childesr` for exploring the CHILDES dataset, as well as some of the potential hazards in working with a diverse set of corpora, many of which are longitudinal.
-
-<!-- SM: not terribly clear what the hazards were: sparsity and autocorrelation? AS: mostly the fact that there are longitudinal studies in here-->
+This plot reveals a slight gender difference in linguistic productivity in young children, replicating the moderate female advantage found by @eriksson2012differences.
+The goal of this analysis was to showcase an example of the the steps involved in using `childesr` for exploring the CHILDES dataset, as well as some of the potential hazards -- sparsity and non-independance-- that emerge in working with a diverse set of corpora, many of which are longitudinal.
 

### 2018-04-18 paper.Rmd
testing density plots

@@
 The results of the comparison were .99 and .98 for the unigram count and mlu data, respectively, indicating a reliable parse on our end.
+
+```{r}
+# plot error between clan counts and our counts for speaker unigrams
+freq_data <- read_csv("data/freq_testing.csv")
+
+ggplot(freq_data %>%
+         mutate(error = abs(clan - db) / clan), 
+       aes(error)) +
+  geom_density() +
+  xlim(0,.25)
+```
+
+```{r}
+# plot error between clan counts and our counts for mlu-w
+mlu_data <- read_csv("data/mlu_testing.csv")
+
+ggplot(freq_data %>%
+         mutate(error = abs(clan - db) / clan), 
+       aes(error)) +
+  geom_density() +
+  xlim(0,.25)
+```
+
 <!-- SM: what were the results of these comparisons?  -->
@@
 
-<!-- AS: not sure whether I should have included this; I didn't want to overwhelm the reader -->
-<!-- SM: need a few more details: what kind of model was it basically? loess, logistic growth curve, etc. -->
-
 ```{r warning=FALSE}
@@
 ```
-
 This plot reveals a slight gender difference in linguistic productivity in young children, replicating the moderate female advantage found by @eriksson2012differences.
-The goal of this analysis was to showcase an example of the the steps involved in using `childesr` for exploring the CHILDES dataset, as well as some of the potential hazards -- sparsity and non-independance-- that emerge in working with a diverse set of corpora, many of which are longitudinal.
+The goal of this analysis was to showcase an example of the the steps involved in using `childesr` for exploring the CHILDES dataset, as well as some of the potential hazards -- sparsity and non-independance-- that emerge in working with a diverse set of corpora, many of which were collected in longitudinal studies.
 

### 2018-04-18 paper.Rmd
edit pass, tweaks throughout

--- 2018-04-18 paper.Rmd
@@
   The Child Language Data Exchange System (CHILDES) has played a critical role in research on child language development, particularly in characterizing the early language learning environment.
-  Access to these data can be complex for novices and difficult to automate for advanced users, however.
+  Access to these data can be both complex for novices and also difficult to automate for advanced users, however.
   To address these issues, we introduce *childes-db*, a database-formatted mirror of CHILDES that improves data accessibility and usability by offering novel interfaces, including browsable web applications and an R application programming interface (API).
@@
 Starting with Roger Brown's [-@brown1973] work on Adam, Eve, and Sarah, audio recordings -- and more recently video recordings -- have been augmented with rich, searchable annotations that have allowed researchers to address a number of questions regarding the language learning environment.
-Focusing on language learning in naturalistic contexts also reveals that children have in many cases productive and receptive abilities exceeding those revealed in experimental contexts; owing to challenges in lab-based experiments with young children, often their most revealing and sophisticated uses of language emerge in naturalistic play.
+Focusing on language learning in naturalistic contexts also reveals that children have in many cases productive and receptive abilities exceeding those demonstrated in experimental contexts; owing to challenges in lab-based experiments with young children, often their most revealing and sophisticated uses of language emerge in naturalistic play.
 
@@
 Collecting and transcribing audio and video is costly and extremely time consuming -- even orthographic transcription (i.e., transcriptions with minimal phonetic detail) can take ten times the length of the original recording to transcribe [@macwhinney2000].
-Automated, machine learning -based methods like automatic speech recognition (ASR) have provided only modest gains in efficiency, in that they are limited by the less-than-ideal acoustic properties of home recordings and the poor fit of language models built on adult-directed, adult-produced language samples for characterizing child-directed or child-produced speech.
-Thus, researchers' desire for sufficient sample sizes and coverage in analyses of child language corpora can very quickly outstrip their resources. 
+Automated, machine learning-based methods like automatic speech recognition (ASR) have provided only modest gains in efficiency. 
+Such systems are limited both by the less-than-ideal acoustic properties of home recordings and also by the poor fit of language models built on adult-directed, adult-produced language samples to child-directed and child-produced speech.
+Thus, researchers' desire for data in analyses of child language corpora can very quickly outstrip their resources. 
 <!-- data pooling is necessary) -->
@@
 Some studies analyze individual examples in depth [e.g., @snyder2007], other track multiple child-caregiver dyads [e.g., @meylan2017], and still others use the aggregate properties of child or caregiver speech [e.g., @redingtonEtAl1998; @montagEtAl2015].
-These are just a few of the thousands of studies that make use of CHILDES corpora.
+<!-- These are just a few of the thousands of studies that make use of CHILDES corpora. -->
 
 Nonetheless, there are some outstanding challenges working with CHILDES, both for students and for advanced users.
-The CHILDES ecosystem typically uses a specialized file format (CHAT), which is stored as plain text but includes structured annotations grouped into tiers (stored on separate lines) so that information about utterances can be stored with accompanying information such as the phonological, morphological, or syntactic structure of the utterance.
-These files are usually analyzed using a command-line tool (CLAN), which has a number of specific tools that allow users to count word frequencies, compute statistics (e.g., mean length of utterance, or MLU), and execute complex searches against the data.
-While this system is flexible and powerful, mastering the CHAT codes and especially the CLAN tool with its many tools and flags can be daunting for a beginner or student.
+The CHILDES ecosystem uses a specialized file format (CHAT), which is stored as plain text but includes structured annotations grouped into tiers (stored on separate lines) so that information about utterances can be stored with accompanying information such as the phonological, morphological, or syntactic structure of the utterance.
+These files are usually analyzed using a command-line program (CLAN) that allows users to count word frequencies, compute statistics (e.g., mean length of utterance, or MLU), and execute complex searches against the data.
+While this system is flexible and powerful, mastering the CHAT codes and especially the CLAN tool with its many functions and flags can be daunting.
 These technical barriers decrease the ease of exploration by a novice researcher or in a classroom exercise.
@@
 CLAN software is an excellent tool for interactive exploration, but -- as a free-standing application -- it can be tricky to build into a processing pipeline written in Python or R.
-Thus, researchers who would like to ingest the entire corpus (or some large subset) into a computational analyses typically write their own parsers of the CHAT format to extract the subset of the data they would like to use [e.g., @redington1998;@yang2013;@meylan2017].
+Thus, researchers who would like to ingest the entire corpus (or some large subset) into a computational analysis typically write their own parsers of the CHAT format to extract the subset of the data they would like to use [e.g., @redington1998;@yang2013;@meylan2017].
 
@@
 Second, this process introduces a major source of errors and inconsistencies in data handling due to difficulties dealing with the many special cases in the CHAT standard.
-Third, these parsing scripts are rarely shared or break with subsequent revisions to the dataset, leading to much greater difficulty in reproducing the exact numerical results from papers using CHILDES [see e.g., @meylan2017 for an example].
-Fourth, the CHILDES corpus itself is a moving target; computational work using the entire corpus at one time point may include a different set of data than subsequent work due to newly-contributed corpora or revised annotations of existing ones; currently, there is no simple way for researchers to document exactly what version of the corpus has been used short of creating a full mirror of the data.
-These factors together lead to a lack of computational reproducibility, a major problem that keeps researchers from verifying or building on published research [@donoho2010;@stodden2016].
+Third, these parsing scripts are rarely shared -- and when when they are, they typically break with subsequent revisions to the dataset -- leading to much greater difficulty in reproducing the exact numerical results from papers using CHILDES [see e.g., @meylan2017 for an example].
+Fourth, the CHILDES corpus itself is a moving target; computational work using the entire corpus at one time point may include a different set of data than subsequent work due to newly-contributed corpora or revised annotations of existing ones.
+Currently, there is no simple way for researchers to document exactly what version of the corpus has been used short of creating a full mirror of the data.
+These factors together lead to a lack of *computational reproducibility*, a major problem that keeps researchers from verifying or building on published research [@donoho2010;@stodden2016].
 
 In the current manuscript, we describe a system for extending the functionality of CHILDES to address these issues.
-Our system, childes-db, is a database-formatted mirror of CHILDES that allows programmatic access through an application programming interface (API). 
+Our system, childes-db, is a database-formatted mirror of CHILDES that allows access through an application programming interface (API). 
 This infrastructure allows the creation of web applications for browsing and easily visualizing the data, facilitating classroom use of the dataset. 
-Further, it can be accessed programmatically by advanced researchers, obviating the need to write one-off parsers of the CHAT format.
+Further, the database can be accessed programmatically by advanced researchers, obviating the need to write one-off parsers of the CHAT format.
 The database is versioned for access to previous releases, allowing computational reproducibility of particular analyses.
@@
 Next, we describe the `childesr` API, which provides a set of R functions for programmatic access to the data.
-We conclude by presenting several worked examples of specific uses of the system--both web apps and the R API--for both research and teaching.
+We conclude by presenting several worked examples of specific uses of the system--both web apps and the R API--for research and teaching.
 
@@
 
-```{r database-schema, fig.pos="h", out.width = "95%", fig.cap="Database schema for childes-db.
+```{r database-schema, fig.pos="t", out.width = "95%", fig.cap="Database schema for childes-db.
 Tokens are linked to superordinate groupings of utterances, transcripts, corpora, and collections (red arrows).
@@
 A *transcript* consists of one or more utterances and includes the date collected, the name of the target child, and age in days (if defined), and the filename from CHILDES.
-A *corpus* consists of one or more transcripts, corresponding to well-known collections like the Brown or Providence corpus.
+A *corpus* consists of one or more transcripts, corresponding to well-known collections like the Brown or Providence corpora.
 Finally, a *collection* is a superordinate collection of corpora generally corresponding to a geographic region, following the convention in CHILDES.
@@
 A participant identifier is associated with every word and utterance, including a name, role, 3-letter CHILDES identifier (CHI = child, MOT = mother, FAT = father, etc.), and the range of ages (or age of corresponding child) for which they are observed.
-For non-child participants (caregivers and others), the record additionally contains an identifier for the corresponding target child, such that data correpsonding to children and their caregivers can be easily associated.
+For non-child participants (caregivers and others), the record additionally contains an identifier for the corresponding target child, such that data corresponding to children and their caregivers can be easily associated.
 
@@
 Logic implemented in Python converts the linear, multi-tier parse into a tabular format appropriate for childes-db.
-This logic includes decisions regarding what information sources are captured in the current release of the database, and which are left for future development (detailed below).
+This logic includes decisions that we review below regarding what information sources are captured in the current release of the database and which are left for future development.
 
@@
 In order to evaluate our success in replicating CLAN parses, we compared unigram counts in our database with those outputted by CLAN, the command-line tool built specifically for analysis of transcripts coded in CHAT.
-In particular, the CLAN commands FREQ and MLU were used to compare total token counts and mean lengths of utterance for every speaker in every transcript; these numbers were compared to our own using the Pearson correlation coefficient.
-The results of the comparison were .99 and .98 for the unigram count and mlu data, respectively, indicating a reliable parse on our end.
+We used the CLAN commands FREQ and MLU to compare total token counts and mean lengths of utterance for every speaker in every transcript and compared these these values to our own using the Pearson correlation coefficient.
+The results of the comparison were .99 and .98 for the unigram count and MLU data, respectively, indicating reliable parsing.
 <!-- SM: what were the results of these comparisons?  -->
@@
 
-The content of CHILDES changes as additional corpora are added or transcriptions are updated; as of the time of writing, these changes are not systematically tracked.
-We introduce a simple versioning system by adding a new complete parse of the current state of CHILDES every six months or as warranted by changes in CHILDES.
+The content of CHILDES changes as additional corpora are added or transcriptions are updated; as of time of writing, these changes are not systematically tracked.
+To facilitate reproducibility of past analyses, we introduce a simple versioning system by adding a new complete parse of the current state of CHILDES every six months or as warranted by changes in CHILDES.
 By default, users interact with the most recent version of the database available. 
@@
 The plot is displayed and updated in real-time, and the underlying data are also available for download alongside the plot to facilitate customization. 
-All of these analyses may also be reproduced using the `childesr` package.
-
-The web apps are intended for the casual user who seeks to easy extract developmental indices quickly and without any technical overhead.
-We next walk through the three web apps provided on our website and describe their functionality in more detail.
+All of these analyses may also be reproduced using the `childesr` package, but the web apps are intended for the casual user who seeks to easy extract developmental indices quickly and without any technical overhead.
+<!-- We next walk through the three web apps provided on our website and describe their functionality in more detail.  -->
 
@@
 
-Many times a researcher will want to investigate the statistics of a given corpus (i.e the size, number of utterances, number of tokens, etc.) before choosing a target corpus for their analysis.
+Many times a researcher will want to investigate the statistics of corpora (e.g., their size, number of utterances, number of tokens) before choosing a target corpus or set of corpora for a project.
 This web app is intended to provide a basic overview of various corpora in the CHILDES ecosystem, facilitating familiarity with the different corpora and thus assisting with the overall analytic end-to-end process.
@@
 # Using childes-db: Worked Examples
-<!-- SM: we should note somewhere that examples all use Tidyverse, so that R users who don't know what %>% don't trip -->
+
 In this section we give a number of examples of how childes-db can be used in both research and teaching, using both the web apps and the R API.
+Note that all of these examples use `dplyr` syntax [@R-dplyr]; several accessible introductions to this framework are available online [e.g., @wickham2016].
 
@@
 We also do not have a good sense of the errors of these measurements, are limited to just a few colors at a time before the plot becomes too busy, and cannot combine frequencies across speakers.
-<!-- SM:"busy" seems a bit informal -->
+<!-- SM:"busy" seems a bit informal; MCF: I'm ok with it.
+:) -->
 To perform this analysis in a more compelling and complete way, a user can use the `childesr` interface.
@@
 For demonstration, we look only at the types produced by the speakers in each corpus tagged as Mother and Father.
-We also restrict ourselves to children from 1-5 years old (12 to 60 months), and look only at the American English corpora.
+We also restrict ourselves to children from 1--5 years old (12--60 months), and look only at the American English corpora.
 
@@
 
-Figure \ref{fig:color-plot} shows the results of these analyses: Input frequency varies significantly over the 1-5 year range for nearly every color word.
+Figure \ref{fig:color-plot} shows the results of these analyses: Input frequency varies substantially over the 1--5 year range for nearly every color word.
 
@@
 
-Gender differences in language production have been found in multiple studies, notably in @eriksson2012differences, a large-scale study that spanned ten languages and found that females have a higher vocabulary than males in early childhood.
-A researcher or student in child language acquisition may now investigate this difference with relative ease using `childesr`.
-By pulling data from the `transcript_by_speaker` table, a user has access to the the multitude of derived linguistic measures that are often used to evaluate a child's grammatical development.
-In this worked example we walk through a sample analysis that probes the gender differences in early grammatical development.
+Gender differences in language production are well-documented.
+For example, @eriksson2012differences used parent-report data from ten languages and found that female children had larger vocabularies on average than male children in nearly every language.
+Comparable cross-linguistic analysis of naturalistic production data has not been conducted, however, and these differences are easy to explore using `childesr`.
+By pulling data from the `transcript_by_speaker` table, users can access derived linguistic measures.
+In this worked example we walk through a sample analysis that probes the gender differences in early lexical diversity.
 
@@
 
-This call pulls data from all collections and corpora, including those languages for which there are very sparse data.
-In order to make any substantial inferences from our analysis, we begin by filtering languages for which there are a significant number of children. 
-We also filter for children under the age of four years.
+This command retrieves data from all collections and corpora, including those languages for which there are very sparse data.
+In order to make any substantial inferences from our analysis, we begin by filtering languages for which there are a large number of transcripts ($>500$). 
+We also restrict our analysis to children under the age of four years.
 
 ```{r echo=TRUE}
-number_of_transcripts_threshold <- 575
+number_of_transcripts_threshold <- 500
 max_age <- 4
@@
   filter(n > number_of_transcripts_threshold) %>%
-  .$language
+  pull(language)
 ```
 
-Our `transcript_by_speaker` table contains multiple derived measures of language productivity and lexical diversity -- here we used MTLD [@mccarthy2005assessment], a measure that is more robust to transcript length, as a way of evaluating children's linguistic productivity.
+Our `transcript_by_speaker` table contains multiple derived measures of lexical diversity -- here we used MTLD [@mccarthy2005assessment], a measure that is relatively robust to transcript length.
 We start by filtering those children for which a sex was defined, who speak a language in our subset, and are in the appropriate age range.
@@
 
-```{r echo=TRUE, warning=FALSE}
+```{r echo=FALSE, warning=FALSE}
 ggplot(data, aes(x = age_years, y = measure, 
@@
 This plot reveals a slight gender difference in linguistic productivity in young children (displaying a moderate female advantage), replicating the results of @eriksson2012differences.
-The goal of this analysis was to showcase an example of the the steps involved in using `childesr` for exploring the CHILDES dataset, as well as some of the potential hazards in working with a diverse set of corpora that were collectd through various experimental paradigms.
+The goal of this analysis was to showcase an example of the the steps involved in using `childesr` for exploring the CHILDES dataset, as well as some of the potential hazards in working with a diverse set of corpora that were collected through various experimental paradigms.
 
@@
 This approach shifts students' effort away from data cleaning and towards programming analyses that encourage the exploration and testing of interesting hypotheses.
+In particular, the `childesr` API provides instructors with an easy-to-learn method for giving students programmatic access to child language data.
 
-The `childesr` API provides instructors with an easy-to-learn method for giving students programmatic access to child language data.
-The aim is to reduce time spent getting data out of CHILDES and allow more course time for thinking about interesting questions that could be answered with NLP and child language data.
-<!-- SM: somewhat redundant with the above -->
-Moreover, by writing our API in the R programming language, students will gain access to a general workflow can be integrated into an analysis pipeline that includes state of the art NLP tools.
+<!-- The aim is to reduce time spent getting data out of CHILDES and allow more course time for thinking about interesting questions that could be answered with NLP and child language data.
+<!-- SM: somewhat redundant with the above --> -->
+<!-- Moreover, by writing our API in the R programming language, students will gain access to a general workflow can be integrated into an analysis pipeline that includes state of the art NLP tools.
 <!-- SM: sorry, what tools?
-there's very little NLP support in R b/c it's string handling is slow -->
+there's very little NLP support in R b/c it's string handling is slow.
+--> -->
 

### 2018-04-18 paper.Rmd

@@
 The results of the comparison were .99 and .98 for the unigram count and MLU data, respectively, indicating reliable parsing.
+
+```{r}
+# plot error between clan counts and our counts for speaker unigrams
+freq_data <- read_csv("data/freq_testing.csv")
+
+ggplot(freq_data %>%
+         mutate(error = abs(clan - db) / clan),
+       aes(error)) +
+  geom_density() +
+  xlim(0,.25)
+```
+
+```{r}
+# plot error between clan counts and our counts for mlu-w
+mlu_data <- read_csv("data/mlu_testing.csv")
+
+ggplot(freq_data %>%
+         mutate(error = abs(clan - db) / clan),
+       aes(error)) +
+  geom_density() +
+  xlim(0,.25)
+```
+
 <!-- SM: what were the results of these comparisons?  -->
@@
 - TTR [type-token ratio, a measure of lexical diversity; @templin1957],
-- MTLD (measure of textual lexical diversity), 
-- HD-D (lexical diversity via the hypergeometric distribution)
+- MTLD [measure of textual lexical diversity; @malvernRichards1997],
+- HD-D [lexical diversity via the hypergeometric distribution; @mccarthy2010])
 <!-- AS: removed voc-d as HDD is idealized version of voc-d -->
@@
 Many times a researcher will want to investigate the statistics of corpora (e.g., their size, number of utterances, number of tokens) before choosing a target corpus or set of corpora for a project.
-This web app is intended to provide a basic overview of various corpora in the CHILDES ecosystem, facilitating familiarity with the different corpora and thus assisting with the overall analytic end-to-end process.
-<!-- note that the graphs don't currently include a grouping factor, so adding multiple corpora generates one line, so you can't actually compare two corpora in the same visualization -->
+This web app is intended to provide a basic overview regarding the scale and temporal extent of various corpora in CHILDES, as well as giving researchers insight into the aggregate characteristics of CHILDES.
+For example, examining the aggregate statistics reveals that coverage peaks at around 30 months.
+<!-- SM: note that the graphs don't currently include a grouping factor, so adding multiple corpora generates one line, so you can't actually compare two corpora in the same visualization -->
 
@@
 
-From this figure, it seems likely that children hear "blue" more frequently early in development, but the trajectories of the "red" and "green" are less clear.
-We also do not have a good sense of the errors of these measurements, are limited to just a few colors at a time before the plot becomes too busy, and cannot combine frequencies across speakers.
-<!-- SM:"busy" seems a bit informal; MCF: I'm ok with it.
-:) -->
+From this figure, it seems likely that children hear "blue" more frequently early in development, but the trajectories of "red" and "green" are less clear.
+We also do not have a good sense of the errors of these measurements, are limited to just a few colors at a time before the plot becomes too crowded, and cannot combine frequencies across speakers.
 To perform this analysis in a more compelling and complete way, a user can use the `childesr` interface.
@@
 
-Gender differences in language production are well-documented.
-For example, @eriksson2012differences used parent-report data from ten languages and found that female children had larger vocabularies on average than male children in nearly every language.
+Gender has long been known to be an important factor for early vocabulary growth, with girls learning more words earlier than boys [@huttenlocher1991].
+Parent-report data from ten languages suggest that female children have larger vocabularies on average than male children in nearly every language [@eriksson2012differences].
 Comparable cross-linguistic analysis of naturalistic production data has not been conducted, however, and these differences are easy to explore using `childesr`.
-By pulling data from the `transcript_by_speaker` table, users can access derived linguistic measures.
-In this worked example we walk through a sample analysis that probes the gender differences in early lexical diversity.
+By pulling data from the `transcript_by_speaker` table, a user has access to the multitude of derived linguistic measures that are often used to evaluate a child's grammatical development.
+In this worked example we walk through a sample analysis that probes gender differences in early lexical diversity.
 
-First, we utilize the `childesr` function call `get_speaker_statistics` to pull data relating to the aforementioned derived measures for children and their transcripts.
+First, we use the `childesr` function call `get_speaker_statistics` to pull data relating to the aforementioned derived measures for children and their transcripts.
 Note that we exclusively select the data of children, and exclude their caregivers.
@@
 
-The data contained in CHILDES is populated from a diverse array of studies that adhered to varying conditions and paradigms.
-This point is particularly salient in our gender analysis due to potential non-independence issues that result from longitudinal studies.
-For this reason, we fit a linear mixed effects model (`mod_data`, shown in the plot below) to remedy these concerns.^[All code and analyses are available at https://github.com/langcog/childes-db-paper]
-The plot below displays the average MTLD scores for various children at different ages, split by gender, with a line corresponding to our mixed effects model.
-
-<!-- AS: is this sufficient for mixed effects model explanation? -->
+The data contained in CHILDES is populated from a diverse array of studies refelcting varying circumstances of data collection.
+This point is particularly salient in our gender analysis due to potential non-independence issues that may emerge from the inclusion of many transcripts from longitudinal studies.
+To rememdy these concerns, we fit a linear mixed effects model with gender * age (treated as a quadratic predictor) as fixed effects, child identity as a random intercept, and age + gender by language as a random slope.^[All code and analyses are available at https://github.com/langcog/childes-db-paper]
+<!--  mod <- lmer(measure ~ target_child_sex * poly(age_years, 2) - 1 +
+                    (1 | target_child_id) +
+                    (target_child_sex + age_years | language),
+  data = data) -->
+<!-- SM: not sure I got the random effects structure right -->
+The plot below displays the average MTLD scores for various children at different ages, split by gender, with a line corresponding to the prediction of our fit mixed effects model.
 
@@
 ```
-
-This plot reveals a slight gender difference in linguistic productivity in young children (displaying a moderate female advantage), replicating the results of @eriksson2012differences.
-The goal of this analysis was to showcase an example of the the steps involved in using `childesr` for exploring the CHILDES dataset, as well as some of the potential hazards in working with a diverse set of corpora that were collected through various experimental paradigms.
-
-<!-- SM: not terribly clear what the hazards were: sparsity and autocorrelation? AS: mostly the fact that there are longitudinal studies in here-->
+This plot reveals a slight gender difference in linguistic productivity in young children, replicating the moderate female advantage found by @eriksson2012differences.
+The goal of this analysis was to showcase an example of the the steps involved in using `childesr` for exploring the CHILDES dataset, as well as some of the potential hazards -- sparsity and non-independance-- that emerge in working with a diverse set of corpora, many of which were collected in longitudinal studies.
 

### 2018-04-18 paper.Rmd
plot change

@@
 
-```{r}
+```{r, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
 # plot error between clan counts and our counts for speaker unigrams
@@
 
-```{r}
+```{r, eval=FALSE, include=FALSE}
 # plot error between clan counts and our counts for mlu-w

### 2018-04-18 paper.Rmd
say something about MTLD

@@
 
-ggplot(freq_data %>%
+ggplot(mlu_data %>%
          mutate(error = abs(clan - db) / clan),
@@
 
-Our `transcript_by_speaker` table contains multiple derived measures of lexical diversity -- here we used MTLD [@mccarthy2005assessment], a measure that is relatively robust to transcript length.
+Our `transcript_by_speaker` table contains multiple derived measures of lexical diversity -- here we used MTLD [@mccarthy2005assessment].
+MTLD is derived from the average length of continuous text units above a certain type-token ratio, making it relatively robust to transcript length.
 We start by filtering those children for which a sex was defined, who speak a language in our subset, and are in the appropriate age range.
@@
   group_by(target_child_id, target_child_age, target_child_sex, language) %>%
-  filter(mtld != 0) %>%
   summarise(measure = mean(mtld)) %>%
@@
 ```
+
 This plot reveals a slight gender difference in linguistic productivity in young children, replicating the moderate female advantage found by @eriksson2012differences.

### 2018-04-18 paper.Rmd
minor line edits

--- 2018-04-18 paper.Rmd
@@
 
-We have created childes-db to address a number of needs in our own teaching and research, but there are still a number of limitations that point the way to future improvements.
+childes-db addresses a number of needs that have emerged in our own research and tecahing, but there are still a number of limitations that point the way to future improvements.
 For example, currently childes-db operates only on transcript data, without links to the underlying media files; these linkages facilitate deeper analyses of phonology, prosody, social interaction, and many other aspects of acquisition.

### 2018-04-18 paper.Rmd
exclude german child

--- 2018-04-18 paper.Rmd
@@
 We then average their MTLD score across transcripts after having grouped the data by child, age, sex, and language.
+One child in particular contained transcripts that were a collection of his most complex utterances (recorded by the parents); this child was excluded from the analysis.
 
@@
          target_child_sex = factor(target_child_sex, levels = c("male","female"))) %>%
-  filter(age_years < max_age)
+  filter(age_years < max_age, target_child_name != 9387)
 ```

### 2018-04-18 paper.Rmd
dan pass

@@
   The Child Language Data Exchange System (CHILDES) has played a critical role in research on child language development, particularly in characterizing the early language learning environment.
-  Access to these data can be both complex for novices and also difficult to automate for advanced users, however.
+  Access to these data can be both complex for novices and difficult to automate for advanced users, however.
   To address these issues, we introduce *childes-db*, a database-formatted mirror of CHILDES that improves data accessibility and usability by offering novel interfaces, including browsable web applications and an R application programming interface (API).
@@
 
-How children acquire language is one of the core puzzles in cognitive science, developmental psychology, and linguistics.
 What are the representations that children learn about language, and how do they emerge from the interaction of learning mechanisms and environmental input?
 Developing facility with language requires learning a great many interlocking components -- meaningful distinctions between sounds (phonology), names of particular objects and actions (word learning), meaningful sub-word structure (morphology), rules for how to organize words together (syntax), and context-dependent and context-independent aspects of meaning (semantics and pragmatics).
-Key to learning all of these systems is the contribution of the child's input -- exposure to linguistic and non-linguistic data -- in the early environment.
+Key to learning all of these systems is the contribution of the child's input--exposure to linguistic and non-linguistic data--in the early environment.
 While in-lab experiments can shed light on linguistic knowledge and some of the implicated learning mechanisms, characterizing this early environment requires additional research methods and resources.
@@
 One of the key methods that has emerged to address this gap is the collection and annotation of speech to and by children, often in the context of the home.
-Starting with Roger Brown's [-@brown1973] work on Adam, Eve, and Sarah, audio recordings -- and more recently video recordings -- have been augmented with rich, searchable annotations that have allowed researchers to address a number of questions regarding the language learning environment.
-Focusing on language learning in naturalistic contexts also reveals that children have in many cases productive and receptive abilities exceeding those demonstrated in experimental contexts; owing to challenges in lab-based experiments with young children, often their most revealing and sophisticated uses of language emerge in naturalistic play.
+Starting with Roger Brown's [-@brown1973] work on Adam, Eve, and Sarah, audio recordings--and more recently video recordings--have been augmented with rich, searchable annotations that have allowed researchers to address a number of questions regarding the language learning environment.
+Focusing on language learning in naturalistic contexts also reveals that children have in many cases productive and receptive abilities exceeding those demonstrated in experimental contexts.
+Often, children's most revealing and sophisticated uses of language emerge in naturalistic play.
 
 While corpora of early language acquisition are extremely useful, creating such corpora requires significant resources.
-Collecting and transcribing audio and video is costly and extremely time consuming -- even orthographic transcription (i.e., transcriptions with minimal phonetic detail) can take ten times the length of the original recording to transcribe [@macwhinney2000].
+Collecting and transcribing audio and video is costly and extremely time consuming--even orthographic transcription (i.e., transcriptions with minimal phonetic detail) can take ten times the duration of the original recording to transcribe [@macwhinney2000].
 Automated, machine learning-based methods like automatic speech recognition (ASR) have provided only modest gains in efficiency.
@@
 Established in 1984 to address this issue, the Child Language Data Exchange System (CHILDES) aims to make transcripts and recording of children's language available for researchers as free, public datasets [@macwhinney1985; @macwhinney2000; @macwhinney2014].
-CHILDES archives tens of thousands of transcripts and associated media across 20+ languages, making it a critical resource for characterizing children’s early productive language use as well as their language environment.
+CHILDES archives tens of thousands of transcripts and associated media across 20+ languages, making it a critical resource for characterizing both children’s early productive language use and their language environment.
 As the first major effort to consolidate and share transcripts of child language, CHILDES has been a pioneer in the move to curate and disseminate large-scale behavioral datasets publicly.
@@
 Nonetheless, there are some outstanding challenges working with CHILDES, both for students and for advanced users.
-The CHILDES ecosystem uses a specialized file format (CHAT), which is stored as plain text but includes structured annotations grouped into tiers (stored on separate lines) so that information about utterances can be stored with accompanying information such as the phonological, morphological, or syntactic structure of the utterance.
+The CHILDES ecosystem uses a specialized file format (CHAT), which is stored as plain text but includes structured annotations grouped into tiers (stored on separate lines).
+These tiers allow information about utterances to be stored with accompanying information such as the phonological, morphological, or syntactic structure of the utterance.
 These files are usually analyzed using a command-line program (CLAN) that allows users to count word frequencies, compute statistics (e.g., mean length of utterance, or MLU), and execute complex searches against the data.
@@
 Fourth, the CHILDES corpus itself is a moving target; computational work using the entire corpus at one time point may include a different set of data than subsequent work due to newly-contributed corpora or revised annotations of existing ones.
-Currently, there is no simple way for researchers to document exactly what version of the corpus has been used short of creating a full mirror of the data.
+Currently, there is no simple way for researchers to document exactly which version of the corpus has been used short of creating a full mirror of the data.
 These factors together lead to a lack of *computational reproducibility*, a major problem that keeps researchers from verifying or building on published research [@donoho2010;@stodden2016].
@@
          language %in% included_languages) %>%
-  group_by(target_child_id, target_child_age, target_child_sex, language) %>%
+  group_by(target_child_id, target_child_age, 
+           target_child_sex, language) %>%
   summarise(measure = mean(mtld)) %>%
@@
   mutate(age_years = target_child_age / 365,
-         target_child_sex = factor(target_child_sex, levels = c("male","female"))) %>%
-  filter(age_years < max_age, target_child_name != 9387)
+         target_child_sex = factor(target_child_sex, 
+                                   levels = c("male","female"))) %>%
+  filter(age_years < max_age, target_child_id != 9387)
 ```
@@
 <!-- The aim is to reduce time spent getting data out of CHILDES and allow more course time for thinking about interesting questions that could be answered with NLP and child language data.
-<!-- SM: somewhat redundant with the above --> -->
+<!-- SM: somewhat redundant with the above -->
 <!-- Moreover, by writing our API in the R programming language, students will gain access to a general workflow can be integrated into an analysis pipeline that includes state of the art NLP tools.
@@
 there's very little NLP support in R b/c it's string handling is slow.
---> -->
+-->
 
@@
 # References
-```{r create_r-references}
+```{r create_r-references, warning = FALSE}
 r_refs(file = "childesdb.bib")

### 2018-04-18 paper.Rmd
kyle pass. minor wording changes and small changes to dplyr code. some questions added as in-line comments

@@
 
-color_counts <- get_types(collection = "Eng-NA", role = c("Mother", "Father"),
-                          age = c(12,60), type = colors)
+color_counts <- get_types(collection = "Eng-NA", 
+                          role = c("Mother", "Father"),
+                          age = c(12,60), 
+                          type = colors)
 ```
@@
 
-# Find in the total number of tokens produced by these speakers
+# Find the total number of tokens produced by these speakers
 parents <- parent_ids %>%
@@
 We then group the data into 6-month age bins, and compute the proportion of tokens that comprise each color for each child in each 6-month bin.
+<!-- KM: The back-to-back summarise() calls feel a little opaque.
+I think there is an implicit group_by() step?
+-->
 For comparability with the web app analysis, these proportions are converted to parts per million words.
@@
 Comparable cross-linguistic analysis of naturalistic production data has not been conducted, however, and these differences are easy to explore using `childesr`.
-By pulling data from the `transcript_by_speaker` table, a user has access to the multitude of derived linguistic measures that are often used to evaluate a child's grammatical development.
-In this worked example we walk through a sample analysis that probes gender differences in early lexical diversity.
+By pulling data from the `transcript_by_speaker` table, a user has access to a set of derived linguistic measures that are often used to evaluate a child's grammatical development.
+In this worked example, we walk through a sample analysis that explores gender differences in early lexical diversity.
 
 First, we use the `childesr` function call `get_speaker_statistics` to pull data relating to the aforementioned derived measures for children and their transcripts.
-Note that we exclusively select the data of children, and exclude their caregivers.
+Note that we exclusively select the children's production data, and exclude their caregivers' speech.
 
@@
 
-This command retrieves data from all collections and corpora, including those languages for which there are very sparse data.
-In order to make any substantial inferences from our analysis, we begin by filtering languages for which there are a large number of transcripts ($>500$).
-We also restrict our analysis to children under the age of four years.
+This `childesr` call retrieves data from all collections and corpora, including those languages for which there are very sparse data.
+In order to make any substantial inferences from our analysis, we begin by filtering the dataset to include only languages for which there are a large number of transcripts ($>500$).
+We also restrict our analysis to children under the age of four years. <!-- KM: justify age range restriction? -->
 
@@
   filter(target_child_age < max_age * 365) %>%       
-  group_by(language) %>%
-  summarise(n = n()) %>%
+  count(language) %>% 
   filter(n > number_of_transcripts_threshold) %>%
@@
 MTLD is derived from the average length of continuous text units above a certain type-token ratio, making it relatively robust to transcript length.
-We start by filtering those children for which a sex was defined, who speak a language in our subset, and are in the appropriate age range.
-We then average their MTLD score across transcripts after having grouped the data by child, age, sex, and language.
-One child in particular contained transcripts that were a collection of his most complex utterances (recorded by the parents); this child was excluded from the analysis.
+We start by filtering to include only those children for which a sex was defined in the transcript, who speak a language in our subset of languages with a large number of transcripts, and who are in the appropriate age range.
+We then compute an average MTLD score for each child at each age point by aggregating across transcripts while keeping information about the child's sex and language.
+Note that one child in particular contained transcripts that were a collection of his most complex utterances (recorded by the parents); this child was excluded from the analysis.
 
@@
 
-The data contained in CHILDES is populated from a diverse array of studies refelcting varying circumstances of data collection.
+The data contained in CHILDES is populated from a diverse array of studies reflecting varying circumstances of data collection.
 This point is particularly salient in our gender analysis due to potential non-independence issues that may emerge from the inclusion of many transcripts from longitudinal studies.
-To rememdy these concerns, we fit a linear mixed effects model with gender * age (treated as a quadratic predictor) as fixed effects, child identity as a random intercept, and age + gender by language as a random slope.^[All code and analyses are available at https://github.com/langcog/childes-db-paper]
+To account for non-indepenence, we fit a linear mixed effects model with a $gender * age$ (treated as a quadratic predictor) interaction as fixed effects, child identity as a random intercept, and $age + gender$ by language as a random slope.^[All code and analyses are available at https://github.com/langcog/childes-db-paper]
 <!--  mod <- lmer(measure ~ target_child_sex * poly(age_years, 2) - 1 +
@@
   data = data) -->
-<!-- SM: not sure I got the random effects structure right -->
+<!-- SM: not sure I got the random effects structure right. KM: I couldn't get the model to converge with the (target_child_sex + age_years | language) random slope-->
 The plot below displays the average MTLD scores for various children at different ages, split by gender, with a line corresponding to the prediction of our fit mixed effects model.
@@
 This plot reveals a slight gender difference in linguistic productivity in young children, replicating the moderate female advantage found by @eriksson2012differences.
-The goal of this analysis was to showcase an example of the the steps involved in using `childesr` for exploring the CHILDES dataset, as well as some of the potential hazards -- sparsity and non-independance-- that emerge in working with a diverse set of corpora, many of which were collected in longitudinal studies.
+The goal of this analysis was to showcase an example of using `childesr` to explore the CHILDES dataset.
+We also highlighted some of the potential pitfalls -- sparsity and non-independance-- that emerge in working with a diverse set of corpora, many of which were collected in longitudinal studies.
 
@@
 <!-- The aim is to reduce time spent getting data out of CHILDES and allow more course time for thinking about interesting questions that could be answered with NLP and child language data.
-<!-- SM: somewhat redundant with the above -->
+<!-- SM: somewhat redundant with the above.
+KM: agreed.
+-->
 <!-- Moreover, by writing our API in the R programming language, students will gain access to a general workflow can be integrated into an analysis pipeline that includes state of the art NLP tools.
@@
 there's very little NLP support in R b/c it's string handling is slow.
+KM: Fair point.
+I was thinking of the tidytext package (https://www.tidytextmining.com/tidytext.html), which I've used with some success.
 -->
 
-For example, an instructor could create a programming assignment with the goal of reproducing the key findings in the case studies presented above -- color words or gender.
+For example, an instructor could create a programming assignment with the specific goal of reproducing the key findings in the case studies presented above -- color words or gender.
 Depending on the students' knowledge of R, the instructor could decide how much of the `childesr` starter code to provide before asking students to generate their own plots and write-ups.
-Moreover, the instructor could take advantage of recent advances in tools for teaching literate programming such as R Markdown [@xie2015dynamic], building in pre-written blocks of executable code that demonstrate key programming concepts for students and including incomplete or blank code blocks that students can use to write their own code.
+The instructor could then easily compare students' code and plots to the expected output to measure learning progress. 
+In addition to specific programming assignments, the instructor could use the `childes-db` and `childesr` workflow as a tool for facilitating student research projects that are designed to address new research questions.
+<!-- Moreover, the instructor could take advantage of recent advances in tools for teaching literate programming such as R Markdown [@xie2015dynamic], building in pre-written blocks of executable code that demonstrate key programming concepts for students and including incomplete or blank code blocks that students can use to write their own code.
 <!-- SM: cool, but scope creep?
+KM: yeah, probably be out of scope.
 -->
@@
 
-childes-db addresses a number of needs that have emerged in our own research and tecahing, but there are still a number of limitations that point the way to future improvements.
+childes-db addresses a number of needs that have emerged in our own research and teaching, but there are still a number of limitations that point the way to future improvements.
 For example, currently childes-db operates only on transcript data, without links to the underlying media files; these linkages facilitate deeper analyses of phonology, prosody, social interaction, and many other aspects of acquisition.

### 2018-04-19 paper.Rmd
edits

@@
 Developing facility with language requires learning a great many interlocking components -- meaningful distinctions between sounds (phonology), names of particular objects and actions (word learning), meaningful sub-word structure (morphology), rules for how to organize words together (syntax), and context-dependent and context-independent aspects of meaning (semantics and pragmatics).
-Key to learning all of these systems is the contribution of the child's input--exposure to linguistic and non-linguistic data--in the early environment.
+Key to learning all of these systems is the contribution of the child's input -- exposure to linguistic and non-linguistic data -- in the early environment.
 While in-lab experiments can shed light on linguistic knowledge and some of the implicated learning mechanisms, characterizing this early environment requires additional research methods and resources.
@@
 One of the key methods that has emerged to address this gap is the collection and annotation of speech to and by children, often in the context of the home.
-Starting with Roger Brown's [-@brown1973] work on Adam, Eve, and Sarah, audio recordings--and more recently video recordings--have been augmented with rich, searchable annotations that have allowed researchers to address a number of questions regarding the language learning environment.
+Starting with Roger Brown's [-@brown1973] work on Adam, Eve, and Sarah, audio recordings -- and more recently video recordings -- have been augmented with rich, searchable annotations that have allowed researchers to address a number of questions regarding the language learning environment.
 Focusing on language learning in naturalistic contexts also reveals that children have in many cases productive and receptive abilities exceeding those demonstrated in experimental contexts.
@@
 While corpora of early language acquisition are extremely useful, creating such corpora requires significant resources.
-Collecting and transcribing audio and video is costly and extremely time consuming--even orthographic transcription (i.e., transcriptions with minimal phonetic detail) can take ten times the duration of the original recording to transcribe [@macwhinney2000].
+Collecting and transcribing audio and video is costly and extremely time consuming -- even orthographic transcription (i.e., transcriptions with minimal phonetic detail) can take ten times the duration of the original recording to transcribe [@macwhinney2000].
 Automated, machine learning-based methods like automatic speech recognition (ASR) have provided only modest gains in efficiency.
@@
 Individual studies are too numerous to list, but classics include studies of morphological over-regularization [@marcus1992], distributional learning [@redington1998], word segmentation [@goldwater2009], the role of frequency in word learning [@goodman2008], and many others.
-Some studies analyze individual examples in depth [e.g., @snyder2007], other track multiple child-caregiver dyads [e.g., @meylan2017], and still others use the aggregate properties of child or caregiver speech [e.g., @redingtonEtAl1998; @montagEtAl2015].
+Some studies analyze individual examples in depth [e.g., @snyder2007], others track multiple child-caregiver dyads [e.g., @meylan2017], and still others use the aggregate properties of child or caregiver speech [e.g., @redingtonEtAl1998; @montagEtAl2015].
 <!-- These are just a few of the thousands of studies that make use of CHILDES corpora. -->
@@
 Next, we describe the `childesr` API, which provides a set of R functions for programmatic access to the data.
-We conclude by presenting several worked examples of specific uses of the system--both web apps and the R API--for research and teaching.
+We conclude by presenting several worked examples of specific uses of the system -- both web apps and the R API -- for research and teaching.
 
@@
 The current implementation of childes-db emphasizes the computation of lexical statistics, and consequently focuses on reproducing the words, utterances, and speaker information in CHILDES transcripts.
-We have yet to parse and make available the following information:
+For this reason, we do not preserve all of the metadata available in CHILDES, such as:
 
@@
 
-We will prioritize the addition of these information sources in response to community feedback; extending coverage to the above information sources will not require changes to the existing database schema.
+We will prioritize the addition of these information sources and others in response to community feedback.
+<!--AS: removed this, as adding this info would require adding new columns: extending coverage to the above information sources will not require changes to the existing database schema.
+-->
 
@@
 The plot is displayed and updated in real-time, and the underlying data are also available for download alongside the plot to facilitate customization.
-All of these analyses may also be reproduced using the `childesr` package, but the web apps are intended for the casual user who seeks to easy extract developmental indices quickly and without any technical overhead.
+All of these analyses may also be reproduced using the `childesr` package, but the web apps are intended for the casual user who seeks to easily extract developmental indices quickly and without any technical overhead.
 <!-- We next walk through the three web apps provided on our website and describe their functionality in more detail.  -->
@@
 The `childesr` package abstracts away the details of connecting to and querying  the database.
-Users can take advantage of the tools developed in the popular `dplyr` package [@R-dplyr], which make manipulating large datasets quick and easy.
+Users can take advantage of the tools developed in the popular `dplyr` package [@R-dplyr], which makes manipulating large datasets quick and easy.
 We describe the commands that the package provides and then give several worked examples of using the package for analyses.
@@
 
-To demonstrate how one might discover such non-stationary, we take as a case study the frequency with which children hear the color words of English (e.g. "blue", "green").
+To demonstrate how one might discover such non-stationarity, we take as a case study the frequency with which children hear the color words of English (e.g. "blue", "green").
 Color words tend to be learned relatively late by children, potentially in part due to the abstractness of the meanings to which they refer [see @wagner2013].
@@
 
-We can analyze these learning trajectories using `childesr` by breaking the process into five steps: (1) define our words of interest, (2) find the frequencies with with which children hear these words, (3) find the proportion of the *total words* children hear that these frequencies account for, (4) aggregate across transcripts and children to determine the error in our estimates of these proportions, and (5) plot the results.
+We can analyze these learning trajectories using `childesr` by breaking the process into five steps: (1) define our words of interest, (2) find the frequencies with which children hear these words, (3) find the proportion of the *total words* children hear that these frequencies account for, (4) aggregate across transcripts and children to determine the error in our estimates of these proportions, and (5) plot the results.
 
@@
 For demonstration, we look only at the types produced by the speakers in each corpus tagged as Mother and Father.
-We also restrict ourselves to children from 1--5 years old (12--60 months), and look only at the American English corpora.
+We also restrict ourselves to children from 1--5 years old (12--60 months), and look only at the North American English corpora.
 
@@
 
-We now join these two pieces of information together--how many times each speaker produced each color word, and how many total words they produced.
+We now join these two pieces of information together -- how many times each speaker produced each color word, and how many total words they produced.
 We then group the data into 6-month age bins, and compute the proportion of tokens that comprise each color for each child in each 6-month bin.
@@
 For example, currently childes-db operates only on transcript data, without links to the underlying media files; these linkages facilitate deeper analyses of phonology, prosody, social interaction, and many other aspects of acquisition.
+<!-- AS: why are these links important?
+Does that mean users have to copy them from childesr and navigate to a website?
+Also, phonology prosody etc is available through tiers so a little confused -->
 Further, we have focused on including the most common and widely-used tiers of CHAT annotation into the database first, but our plan is eventually to include the full range of tiers.

### 2018-04-19 paper.Rmd
gender model now converges

@@
 library(ggthemes)
+library(optimx)
 knitr::opts_chunk$set(cache = TRUE, echo = FALSE)
@@
 
-```{r warning=FALSE}
+```{r echo=FALSE}
 mod <- lmer(measure ~ target_child_sex * poly(age_years, 2) - 1 +
@@
                     (target_child_sex + age_years | language),
-  data = data)
+  data = data, REML = TRUE, control = lmerControl(
+                           optimizer ='optimx', optCtrl=list(method='nlminb')))
 
@@
 ggplot(data, aes(x = age_years, y = measure,
-                 color = target_child_sex)) +
-  geom_point(alpha = .3) +
+                 fill = target_child_sex, colour = target_child_sex)) +
+  geom_point(alpha = .3, size=1, stroke=0) +
   geom_line(data = mod_data,
@@
 childes-db addresses a number of needs that have emerged in our own research and teaching, but there are still a number of limitations that point the way to future improvements.
-For example, currently childes-db operates only on transcript data, without links to the underlying media files; these linkages facilitate deeper analyses of phonology, prosody, social interaction, and many other aspects of acquisition.
+For example, childes-db currently operates only on transcript data, without links to the underlying media files; in the future, adding such links may facilitate further computational and manual analyses of phonology, prosody, social interaction, and other phenomena by providing easy access to the video and audio data.
 <!-- AS: why are these links important?

### 2018-04-19 paper.Rmd
conclusion edits

--- 2018-04-19 paper.Rmd
@@
 
-We have presented childes-db, a database formatted mirror of the CHILDES dataset.
+We have presented `childes-db`, a database formatted mirror of the CHILDES dataset.
 This database -- together with the `R` API and web apps -- facilitates the use of child language data.
@@
 
-childes-db addresses a number of needs that have emerged in our own research and teaching, but there are still a number of limitations that point the way to future improvements.
-For example, currently childes-db operates only on transcript data, without links to the underlying media files; these linkages facilitate deeper analyses of phonology, prosody, social interaction, and many other aspects of acquisition.
-<!-- AS: why are these links important?
-Does that mean users have to copy them from childesr and navigate to a website?
-Also, phonology prosody etc is available through tiers so a little confused -->
+`childes-db` addresses a number of needs that have emerged in our own research and teaching, but there are still a number of limitations that point the way to future improvements.
+For example, childes-db currently operates only on transcript data, without links to the underlying media files.
+In the future, adding such links may facilitate further computational and manual analyses of phonology, prosody, social interaction, and other phenomena by providing easy access to the video and audio data.
 Further, we have focused on including the most common and widely-used tiers of CHAT annotation into the database first, but our plan is eventually to include the full range of tiers.

### 2018-04-19 paper.Rmd

--- 2018-04-19 paper.Rmd
@@
 For non-child participants (caregivers and others), the record additionally contains an identifier for the corresponding target child, such that data corresponding to children and their caregivers can be easily associated.
+
+Adopting a tabular format entails trade-offs.
+Although we have simplified access to the underlying CHILDES dataset through simplified storage mechanisms, the associated metadata may not be preserved without incurring a cost of sparsity.
+This is due to the relatively unique and idiosyncratic annotations provided in CHAT that only apply to small subsets of utterances, or are only offered by some transcripts.
 
@@
 For example, childes-db currently operates only on transcript data, without links to the underlying media files; in the future, adding such links may facilitate further computational and manual analyses of phonology, prosody, social interaction, and other phenomena by providing easy access to the video and audio data.
-<!-- AS: why are these links important?
-Does that mean users have to copy them from childesr and navigate to a website?
-Also, phonology prosody etc is available through tiers so a little confused -->
 Further, we have focused on including the most common and widely-used tiers of CHAT annotation into the database first, but our plan is eventually to include the full range of tiers.

### 2018-04-19 paper.Rmd
remove tradeoff

@@
 For non-child participants (caregivers and others), the record additionally contains an identifier for the corresponding target child, such that data corresponding to children and their caregivers can be easily associated.
-
-Adopting a tabular format entails trade-offs.
-Although we have simplified access to the underlying CHILDES dataset through simplified storage mechanisms, the associated metadata may not be preserved without incurring a cost of sparsity.
-This is due to the relatively unique and idiosyncratic annotations provided in CHAT that only apply to small subsets of utterances, or are only offered by some transcripts.
 

### 2018-04-20 paper.Rmd
switched random effects to gender * age_years (to include interaction term). Still converges

@@
 This point is particularly salient in our gender analysis due to potential non-independence issues that may emerge from the inclusion of many transcripts from longitudinal studies.
-To account for non-indepenence, we fit a linear mixed effects model with a $gender * age$ (treated as a quadratic predictor) interaction as fixed effects, child identity as a random intercept, and $age + gender$ by language as a random slope.^[All code and analyses are available at https://github.com/langcog/childes-db-paper]
+To account for non-indepenence, we fit a linear mixed effects model with a $gender * age$ (treated as a quadratic predictor) interaction as fixed effects, child identity as a random intercept, and $gender * age$ by language as a random slope.^[All code and analyses are available at https://github.com/langcog/childes-db-paper]
 <!--  mod <- lmer(measure ~ target_child_sex * poly(age_years, 2) - 1 +
@@
                     (1 | target_child_id) +
-                    (target_child_sex + age_years | language),
+                    (target_child_sex * age_years | language),
   data = data, REML = TRUE, control = lmerControl(
@@
 mod_data$measure <- predict(mod, newdata = mod_data,
-                            re.form = ~(target_child_sex + age_years | language))
+                            re.form = ~(target_child_sex * age_years | language))
 ```

### 2018-04-20 paper.Rmd
remove interactions + add leo name

@@
 We then compute an average MTLD score for each child at each age point by aggregating across transcripts while keeping information about the child's sex and language.
-Note that one child in particular contained transcripts that were a collection of his most complex utterances (recorded by the parents); this child was excluded from the analysis.
+Note that one child in particular, "Leo" in the eponymous corpus, contained transcripts that were a collection of his most complex utterances (recorded by the parents); this child was excluded from the analysis.
 
@@
 data <- stats %>%
-  filter(!is.na(target_child_sex),
+  filter(!is.na(target_child_sex), target_child_name != "Leo",
          language %in% included_languages) %>%
@@
                                    levels = c("male","female"))) %>%
-  filter(age_years < max_age, target_child_id != 9387)
+  filter(age_years < max_age)
 ```
@@
 ```{r echo=FALSE}
-mod <- lmer(measure ~ target_child_sex * poly(age_years, 2) - 1 +
+mod <- lmer(measure ~ target_child_sex + poly(age_years, 2) - 1 +
                     (1 | target_child_id) +
-                    (target_child_sex * age_years | language),
+                    (target_child_sex + age_years | language),
   data = data, REML = TRUE, control = lmerControl(
@@
 mod_data$measure <- predict(mod, newdata = mod_data,
-                            re.form = ~(target_child_sex * age_years | language))
+                            re.form = ~(target_child_sex + age_years | language))
 ```

### 2018-04-21 paper.Rmd
line edits and polish

@@
   Access to these data can be both complex for novices and difficult to automate for advanced users, however.
-  To address these issues, we introduce *childes-db*, a database-formatted mirror of CHILDES that improves data accessibility and usability by offering novel interfaces, including browsable web applications and an R application programming interface (API).
-  Along with a versioned infrastructure that facilitates reproducibility of past analyses, these interfaces lower barriers to analyzing naturalistic parent-child language, allowing for a wider range of researchers in language and cognitive development to easily leverage CHILDES in their work.
+  To address these issues, we introduce `childes-db`, a database-formatted mirror of CHILDES that improves data accessibility and usability by offering novel interfaces, including browsable web applications and an R application programming interface (API).
+  Along with versioned infrastructure that facilitates reproducibility of past analyses, these interfaces lower barriers to analyzing naturalistic parent-child language, allowing for a wider range of researchers in language and cognitive development to easily leverage CHILDES in their work.
 
@@
 One of the key methods that has emerged to address this gap is the collection and annotation of speech to and by children, often in the context of the home.
-Starting with Roger Brown's [-@brown1973] work on Adam, Eve, and Sarah, audio recordings -- and more recently video recordings -- have been augmented with rich, searchable annotations that have allowed researchers to address a number of questions regarding the language learning environment.
-Focusing on language learning in naturalistic contexts also reveals that children have in many cases productive and receptive abilities exceeding those demonstrated in experimental contexts.
-Often, children's most revealing and sophisticated uses of language emerge in naturalistic play.
+Starting with Roger Brown's [-@brown1973] work on Adam, Eve, and Sarah, audio recordings -- and more recently video recordings -- have been augmented with rich, searchable annotations to allow researchers to address a number of questions regarding the language learning environment.
+Focusing on language learning in naturalistic contexts also reveals that children have, in many cases, productive and receptive abilities exceeding those demonstrated in experimental contexts.
+Often, children's most revealing and sophisticated uses of language emerge in the course of naturalistic play.
 
-While corpora of early language acquisition are extremely useful, creating such corpora requires significant resources.
-Collecting and transcribing audio and video is costly and extremely time consuming -- even orthographic transcription (i.e., transcriptions with minimal phonetic detail) can take ten times the duration of the original recording to transcribe [@macwhinney2000].
+While corpora of early language acquisition are extremely useful, creating them requires significant resources.
+Collecting and transcribing audio and video is costly and extremely time consuming -- even orthographic transcription (i.e., transcriptions with minimal phonetic detail) can take ten times the duration of the original recording [@macwhinney2000].
 Automated, machine learning-based methods like automatic speech recognition (ASR) have provided only modest gains in efficiency.
-Such systems are limited both by the less-than-ideal acoustic properties of home recordings and also by the poor fit of language models built on adult-directed, adult-produced language samples to child-directed and child-produced speech.
-Thus, researchers' desire for data in analyses of child language corpora can very quickly outstrip their resources.
+Such systems are limited both by the less-than-ideal acoustic properties of home recordings, and also by the poor fit of language models built on adult-directed, adult-produced language samples to child-directed and child-produced speech.
+Thus, researchers' desires for data in analyses of child language corpora can very quickly outstrip their resources.
 <!-- data pooling is necessary) -->
 
-Established in 1984 to address this issue, the Child Language Data Exchange System (CHILDES) aims to make transcripts and recording of children's language available for researchers as free, public datasets [@macwhinney1985; @macwhinney2000; @macwhinney2014].
-CHILDES archives tens of thousands of transcripts and associated media across 20+ languages, making it a critical resource for characterizing both children’s early productive language use and their language environment.
+Established in 1984 to address this issue, the Child Language Data Exchange System (CHILDES) aims to make transcripts and recordings relevant to the study of child language acquisition available to researchers as free, public datasets [@macwhinney1985; @macwhinney2000; @macwhinney2014].
+CHILDES now archives tens of thousands of transcripts and associated media across 20+ languages, making it a critical resource for characterizing both children’s early productive language use and their language environment.
 As the first major effort to consolidate and share transcripts of child language, CHILDES has been a pioneer in the move to curate and disseminate large-scale behavioral datasets publicly.
@@
 Individual studies are too numerous to list, but classics include studies of morphological over-regularization [@marcus1992], distributional learning [@redington1998], word segmentation [@goldwater2009], the role of frequency in word learning [@goodman2008], and many others.
-Some studies analyze individual examples in depth [e.g., @snyder2007], others track multiple child-caregiver dyads [e.g., @meylan2017], and still others use the aggregate properties of child or caregiver speech [e.g., @redingtonEtAl1998; @montagEtAl2015].
+Some studies analyze individual examples in depth [e.g., @snyder2007], others track multiple child-caregiver dyads [e.g., @meylan2017], and still others use the aggregate properties of all child or caregiver speech pooled across corpora [e.g., @redington1998; @montagEtAl2015].
 <!-- These are just a few of the thousands of studies that make use of CHILDES corpora. -->
@@
 Nonetheless, there are some outstanding challenges working with CHILDES, both for students and for advanced users.
-The CHILDES ecosystem uses a specialized file format (CHAT), which is stored as plain text but includes structured annotations grouped into tiers (stored on separate lines).
+The CHILDES ecosystem uses a specialized file format (CHAT), which is stored as plain text but includes structured annotations grouped into tiers stored on separate lines.
 These tiers allow information about utterances to be stored with accompanying information such as the phonological, morphological, or syntactic structure of the utterance.
@@
 First, effort is wasted in implementing the same features again and again.
-Second, this process introduces a major source of errors and inconsistencies in data handling due to difficulties dealing with the many special cases in the CHAT standard.
-Third, these parsing scripts are rarely shared -- and when when they are, they typically break with subsequent revisions to the dataset -- leading to much greater difficulty in reproducing the exact numerical results from papers using CHILDES [see e.g., @meylan2017 for an example].
-Fourth, the CHILDES corpus itself is a moving target; computational work using the entire corpus at one time point may include a different set of data than subsequent work due to newly-contributed corpora or revised annotations of existing ones.
-Currently, there is no simple way for researchers to document exactly which version of the corpus has been used short of creating a full mirror of the data.
+Second, this process can introduce errors and inconsistencies in data handling due to difficulties dealing with the many special cases in the CHAT standard.
+Third, these parsing scripts are rarely shared -- and when when they are, they typically break with subsequent revisions to the dataset -- leading to much greater difficulty in reproducing the exact numerical results from previous published research that used CHILDES [see e.g., @meylan2017 for an example].
+Fourth, the CHILDES corpus itself is a moving target: computational work using the entire corpus at one time point may include a different set of data than subsequent work due as corpora are added and revised.
+Currently, there is no simple way for researchers to document exactly which version of the corpus has been used, short of creating a full mirror of the data.
 These factors together lead to a lack of *computational reproducibility*, a major problem that keeps researchers from verifying or building on published research [@donoho2010;@stodden2016].
@@
 In the current manuscript, we describe a system for extending the functionality of CHILDES to address these issues.
-Our system, childes-db, is a database-formatted mirror of CHILDES that allows access through an application programming interface (API).
+Our system, `childes-db`, is a database-formatted mirror of CHILDES that allows access through an application programming interface (API).
 This infrastructure allows the creation of web applications for browsing and easily visualizing the data, facilitating classroom use of the dataset.
@@
 
-We begin by describing the architecture of childes-db and the web applications that we provide.
-Next, we describe the `childesr` API, which provides a set of R functions for programmatic access to the data.
+We begin by describing the architecture of `childes-db` and the web applications that we provide.
+Next, we describe the `childesr` API, which provides a set of R functions for programmatic access to the data while abstracting away many of the technical details.
 We conclude by presenting several worked examples of specific uses of the system -- both web apps and the R API -- for research and teaching.
@@
 
-As described above, CHILDES is most often approached as a set of distinct CHAT files, which are then parsed by users (often using CLAN).
-In contrast to this parsing approach, which entails the sequential processing of strings, childes-db treats CHILDES as a set of linked tables, with records corresponding to intuitive abstractions such as words, utterances, and transcripts.
+As described above, CHILDES is most often approached as a set of distinct CHAT files, which are then parsed by users, often using CLAN.
+In contrast to this parsing approach, which entails the sequential processing of strings, `childes-db` treats CHILDES as a set of linked tables, with records corresponding to intuitive abstractions such as words, utterances, and transcripts [see @klineCLANtoR for an earlier example of deriving tabular representations of CHILDES].
 Users of data analysis languages like R or Julia, libraries like Pandas, or those familiar with Structured Query Language (SQL) will be familiar with operations on tables such as filtering (subsetting), sorting, aggregation (grouping), and joins (merges).
 These operations obviate the need for users to consider the specifics of the CHAT representation -- instead they simply request the entities they need for their research and allow the API to take care of the formatting details.
-We begin by orienting readers to the design of the system via a top-level description and motivation for the design of the database schema, details of the database’s current technical implementation, and our versioning scheme.
+We begin by orienting readers to the design of the system via a top-level description and motivation for the design of the database schema, then provide details on the database’s current technical implementation and the versioning scheme.
 Users primarily interested in accessing the database can skip these details and focus on access through the `childesr` API and the web apps.
@@
 
-At its core, childes-db is a database consisting of a set of linked tabular data stores where records correspond to linguistic entities like words, utterances, and sampling units like transcriptions and corpora.
-The smallest unit of abstraction tracked by the database is a token, treated here as the standard (citation) orthographic form of a word.
-Using the standardized written form of the word facilitates the computation of lexical frequency statistics, for comparison or aggregation across children or time periods.
-Deviations from the citation form -- reflecting the deviations from citation forms that are particularly common in the course of language development -- are kept as a separate (possibly null) field associated with each token.
+At its core, `childes-db` is a database consisting of a set of linked tabular data stores where records correspond to linguistic entities like words, utterances, and sampling units like transcriptions and corpora.
+The smallest unit of abstraction tracked by the database is a *token*, treated here as the standard (or citation) orthographic form of a word.
+Using the standardized written form of the word facilitates the computation of lexical frequency statistics for comparison or aggregation across children or time periods.
+Deviations from the citation form -- which are particularly common in the course of language development and often of interest to researchers -- are kept as a separate (possibly null) field associated with each token.
 
-```{r database-schema, fig.pos="t", out.width = "95%", fig.cap="Database schema for childes-db.
+```{r database-schema, fig.pos="t", out.width = "95%", fig.cap="Database schema for `childes-db`.
 Tokens are linked to superordinate groupings of utterances, transcripts, corpora, and collections (red arrows).
@@
 
-Many of the other tables in the database are hierarchical collections built out of tokens -- *utterance*, *transcript*, *corpus*, and *collection* -- which store attributes appropriate for each level of description.
+Many of the other tables in the database are hierarchical collections built out of tokens -- *utterance*, *transcript*, *corpus*, and *collection* -- that store attributes appropriate for each level of description.
 Every entity includes attributes that link it to all higher-order collections, e.g., an utterance lists the transcript, corpus, and collection to which it belongs.
-An *utterance* contains one or more words and includes fields such as the utterance type (declarative, question, etc.), total number of tokens, and the total number of morphemes (if the morphological structure is available in the original CHAT).
-A *transcript* consists of one or more utterances and includes the date collected, the name of the target child, and age in days (if defined), and the filename from CHILDES.
-A *corpus* consists of one or more transcripts, corresponding to well-known collections like the Brown or Providence corpora.
+An *utterance* contains one or more words and includes fields such as the utterance type such as *declarative* or *interrogative*, total number of tokens, and the total number of morphemes if the morphological structure is available in the original CHAT file.
+A *transcript* consists of one or more utterances and includes the date collected, the name of the target child, and age in days if defined, and the filename from CHILDES.
+A *corpus* consists of one or more transcripts, corresponding to well-known collections like the Brown [@brown1973] or Providence [@demuth2006word] corpora.
 Finally, a *collection* is a superordinate collection of corpora generally corresponding to a geographic region, following the convention in CHILDES.
@@
 
-Participants (e.g., children and caregivers) are represented separately from the token hierarchy because it is common for the same children to appear in multiple transcripts.
+Participants -- generally children and caregivers -- are represented separately from the token hierarchy because it is common for the same children to appear in multiple transcripts.
 A participant identifier is associated with every word and utterance, including a name, role, 3-letter CHILDES identifier (CHI = child, MOT = mother, FAT = father, etc.), and the range of ages (or age of corresponding child) for which they are observed.
@@
 
-childes-db is implemented as a MySQL database, an industry-standard, open-source relational database server that can be accessed directly from a wide range of programming languages.
+`childes-db` is stored as a MySQL database, an industry-standard, open-source relational database server that can be accessed directly from a wide range of programming languages.
 The childes-db project provides hosted, read-only databases for direct access and for `childesr` (described below) as well as compressed .sql exports for local installation.
-While the former is appropriate for most users, local installation can provide performance gains (allowing a user to access the database on their machine or on their local network) as well as allowing users to store derived information in the same database.
+While the former is appropriate for most users, local installation can provide performance gains by allowing a user to access the database on their machine or on their local network, as well as allowing users to store derived information in the same database.
 
@@
 AS: I don't know if I would call it a "contribution" -- we just refactored it to make it more convenient for our purposes -->
-Logic implemented in Python converts the linear, multi-tier parse into a tabular format appropriate for childes-db.
+Logic implemented in Python converts the linear, multi-tier parse into a tabular format appropriate for `childes-db`.
 This logic includes decisions that we review below regarding what information sources are captured in the current release of the database and which are left for future development.
 
-The data imported into childes-db is subject to data integrity checks to ensure that our import of the corpora is accurate and preferable over ad-hoc parsers developed by many individual researchers.
+The data imported into `childes-db` is subject to data integrity checks to ensure that our import of the corpora is accurate and preferable over ad-hoc parsers developed by many individual researchers.
 <!-- MCF: more testing details would be useful -->
@@
 
-The current implementation of childes-db emphasizes the computation of lexical statistics, and consequently focuses on reproducing the words, utterances, and speaker information in CHILDES transcripts.
-For this reason, we do not preserve all of the metadata available in CHILDES, such as:
+The current implementation of `childes-db` emphasizes the computation of lexical statistics, and consequently focuses on reproducing the words, utterances, and speaker information in CHILDES transcripts.
+For this reason, we do not preserve all of the information available in CHILDES, such as:
 
@@
 
-We first discuss the childes-db web apps and then introduce the `childesr` `R` package.
+We first discuss the `childes-db` web apps and then introduce the `childesr` `R` package.
 
@@
 All of this is doable without the requirement of understanding command-line tools or any kind of programming knowledge as had been the case with CLAN.^[The LuCiD toolkit [@chang2017] provides related functionality for a number of common analyses.
-In contrast to those tools, which focus on filling gaps not covered by CLAN -- e.g., the use of n-gram models, incremental sentence generation, and distributional word classification -- our web apps focus on covering the same common tasks as CLAN, but making the outputs into browsable visualizations.]
+In contrast to those tools, which focus on filling gaps not covered by CLAN -- e.g., the use of $n$-gram models, incremental sentence generation, and distributional word classification -- our web apps focus on covering the same common tasks as CLAN, but making the outputs into browsable visualizations.]
 
@@
 The user may specify what collection, corpus, child, age range, caregiver, etc., should be included in a given analysis.
-The plot is displayed and updated in real-time, and the underlying data are also available for download alongside the plot to facilitate customization.
+The plot is displayed and updated in real-time, and the underlying data are also available for download alongside the plot.
 All of these analyses may also be reproduced using the `childesr` package, but the web apps are intended for the casual user who seeks to easily extract developmental indices quickly and without any technical overhead.
@@
 In this web app, inspired by ChildFreq [@baath2010], we provide users the ability to search for any word spoken by a participant in the CHILDES corpora and track the usage of that word by a child or caregiver over time.
-Because of the various toggles available to the user that can subset the data, a user may display frequency curves for a single child in the Brown corpus or all Spanish speaking children, if desired.
+Because of the various toggles available to the user that can subset the data, a user may word frequencies curves for a single child in the Brown corpus or all Spanish speaking children, if desired.
 In addition, users can plot frequency curves belonging to caregivers alongside their child for convenient side-by-side comparisons.
@@
 The syntactic complexity and lexical diversity of children's speech are similarly critical metrics for acquisition researchers [@miller1981;@watkins1995].
-There are a number of well-established measures of children's speech that operationalize complexity and diversity.
-These developmental metrics have many applications in speech-language pathology (SLP), where measures outside of the normal range may be indicative of speech, language, or communication disorders.
+There are a number of well-established measures of children's speech that operationalize complexity and diversity, and have many applications in speech-language pathology (SLP), where measures outside of the normal range may be indicative of speech, language, or communication disorders.
 
@@
 - MTLD [measure of textual lexical diversity; @malvernRichards1997],
-- HD-D [lexical diversity via the hypergeometric distribution; @mccarthy2010])
+- HD-D [lexical diversity via the hypergeometric distribution; @mccarthy2010]
 <!-- AS: removed voc-d as HDD is idealized version of voc-d -->
@@
 This web app is intended to provide a basic overview regarding the scale and temporal extent of various corpora in CHILDES, as well as giving researchers insight into the aggregate characteristics of CHILDES.
-For example, examining the aggregate statistics reveals that coverage peaks at around 30 months.
+For example, examining the aggregate statistics reveals that coverage in CHILDES peaks at around 30 months.
 <!-- SM: note that the graphs don't currently include a grouping factor, so adding multiple corpora generates one line, so you can't actually compare two corpora in the same visualization -->
@@
 
-Although the interactive analysis tools described above cover some of the most common use cases of CHILDES data, researchers interested in more detailed and flexible analyses will want to interface directly with the data in childes-db.
+Although the interactive analysis tools described above cover some of the most common use cases of CHILDES data, researchers interested in more detailed and flexible analyses will want to interface directly with the data in `childes-db`.
 Making use of the `R` programming language [@R-base], we provide the `childesr` package.
@@
 \noindent Each of these functions take arguments that restrict the query to a particular subset of the data (e.g.
-by collection, by corpus, by speaker role, by target child age, etc.) and for a particular version of the data.
+by collection, by corpus, by speaker role, by target child age, etc.) and returns the output in the form of a table.
+All functions support the specification of the database version to use.
 For more detailed documentation, see the package repository (http://github.com/langcog/childesr).
@@
 
-In this section we give a number of examples of how childes-db can be used in both research and teaching, using both the web apps and the R API.
+In this section we give a number of examples of how `childes-db` can be used in both research and teaching, using both the web apps and the R API.
 Note that all of these examples use `dplyr` syntax [@R-dplyr]; several accessible introductions to this framework are available online [e.g., @wickham2016].
@@
 
-Our `transcript_by_speaker` table contains multiple derived measures of lexical diversity -- here we used MTLD [@mccarthy2005assessment].
-MTLD is derived from the average length of continuous text units above a certain type-token ratio, making it relatively robust to transcript length.
+Our `transcript_by_speaker` table contains multiple derived measures of lexical diversity -- here we use MTLD [@mccarthy2005assessment].
+MTLD is derived from the average length of orthographic words that are above a pre-specified type-token ratio, making it more robust to transcript length than simple TTR.
 We start by filtering to include only those children for which a sex was defined in the transcript, who speak a language in our subset of languages with a large number of transcripts, and who are in the appropriate age range.
 We then compute an average MTLD score for each child at each age point by aggregating across transcripts while keeping information about the child's sex and language.
-Note that one child in particular, "Leo" in the eponymous corpus, contained transcripts that were a collection of his most complex utterances (recorded by the parents); this child was excluded from the analysis.
+Note that one child in particular, "Leo" in the eponymous German corpus, contained transcripts that were a collection of his most complex utterances (as caregivers were instructed to record); this child was excluded from the analysis.
 
@@
 This point is particularly salient in our gender analysis due to potential non-independence issues that may emerge from the inclusion of many transcripts from longitudinal studies.
-To account for non-indepenence, we fit a linear mixed effects model with a $gender * age$ (treated as a quadratic predictor) interaction as fixed effects, child identity as a random intercept, and $gender * age$ by language as a random slope.^[All code and analyses are available at https://github.com/langcog/childes-db-paper]
+To account for non-indepenence, we fit a linear mixed effects model with a $gender * age$ (treated as a quadratic predictor) interaction as fixed effects, child identity as a random intercept, and $gender + age$ by language as a random slope, the maximal converging random effects structure [@barr2013random].^[All code and analyses are available at https://github.com/langcog/childes-db-paper]
 <!--  mod <- lmer(measure ~ target_child_sex * poly(age_years, 2) - 1 +
@@
 ```{r echo=FALSE}
-mod <- lmer(measure ~ target_child_sex + poly(age_years, 2) - 1 +
+mod <- lmer(measure ~ target_child_sex * poly(age_years, 2) - 1 +
                     (1 | target_child_id) +
@@
 The goal of this analysis was to showcase an example of using `childesr` to explore the CHILDES dataset.
-We also highlighted some of the potential pitfalls -- sparsity and non-independance-- that emerge in working with a diverse set of corpora, many of which were collected in longitudinal studies.
+We also highlighted some of the potential pitfalls -- sparsity and non-independance -- that emerge in working with a diverse set of corpora, many of which were collected in longitudinal studies.
 
@@
 childes-db addresses a number of needs that have emerged in our own research and teaching, but there are still a number of limitations that point the way to future improvements.
-For example, childes-db currently operates only on transcript data, without links to the underlying media files; in the future, adding such links may facilitate further computational and manual analyses of phonology, prosody, social interaction, and other phenomena by providing easy access to the video and audio data.
+For example, `childes-db` currently operates only on transcript data, without links to the underlying media files; in the future, adding such links may facilitate further computational and manual analyses of phonology, prosody, social interaction, and other phenomena by providing easy access to the video and audio data.
 Further, we have focused on including the most common and widely-used tiers of CHAT annotation into the database first, but our plan is eventually to include the full range of tiers.

### 2018-04-21 paper.Rmd
typo

@@
 library(tidyverse)
-library(childesr)
+#library(childesr)
+source("/Users/alessandro/Github/childesr/R/childesr.R")
+library(magrittr)
+library(RJSONIO)
 library(directlabels)
@@
 The goal of this analysis was to showcase an example of using `childesr` to explore the CHILDES dataset.
-We also highlighted some of the potential pitfalls -- sparsity and non-independance -- that emerge in working with a diverse set of corpora, many of which were collected in longitudinal studies.
+We also highlighted some of the potential pitfalls -- sparsity and non-independence -- that emerge in working with a diverse set of corpora, many of which were collected in longitudinal studies.
 

### 2018-04-22 gender_model.Rmd
several variants of the gender model

---
title: "Mixed-Effects Model for MTLD by Gender"
output: html_notebook
---

```{r load_packages, echo = FALSE, warning=FALSE}
library(papaja)
library(tidyverse)
library(childesr)
library(directlabels)
library(png)
library(here)
library(lme4)
library(ggthemes)
library(optimx)
knitr::opts_chunk$set(cache = TRUE, echo = FALSE)
```

```{r echo=TRUE}
stats <- get_speaker_statistics(role = "Target_Child")
```

```{r echo=TRUE}
number_of_transcripts_threshold <- 500
max_age <- 4

included_languages <- stats %>%
  filter(target_child_age < max_age * 365) %>%       
  count(language) %>% 
  filter(n > number_of_transcripts_threshold) %>%
  pull(language)
```

```{r echo=TRUE}
data <- stats %>%
  filter(!is.na(target_child_sex), target_child_name != "Leo",
         language %in% included_languages) %>%
  group_by(target_child_id, target_child_age, 
           target_child_sex, language) %>%
  summarise(measure = mean(mtld)) %>%
  ungroup() %>%
  mutate(age_years = target_child_age / 365,
         target_child_sex = factor(target_child_sex, 
                                   levels = c("male","female"))) %>%
  filter(age_years < max_age)
```

```{r echo=TRUE, warning=FALSE}

draw_gender_plot <- function(data, mod_data){
  gender_plot = ggplot(data, aes(x = age_years, y = measure,
                 fill = target_child_sex, colour = target_child_sex)) +
  geom_point(alpha = .3, size=1, stroke=0) +
  geom_line(data = mod_data,
            aes(x = age_years, y = measure, color = target_child_sex)) +
  facet_wrap(~language) +
  ggthemes::theme_few() +
  ggthemes::scale_color_ptol(name = "Sex") +
  xlim(0,max_age) +
  ylim(0,15) +
  xlab("Age (years)") +
  ylab("MTLD") +
  theme(legend.position = "bottom")
  print(gender_plot)
}  
```

# Model 1: (currently in paper)

```{r echo=TRUE, warning=TRUE}
mod <- lmer(measure ~ target_child_sex * poly(age_years, 2) - 1 +
                    (1 | target_child_id) +
                    (target_child_sex + age_years | language),
  data = data, REML = TRUE, control = lmerControl(
                           optimizer ='optimx', optCtrl=list(method='nlminb')))

mod_data <- expand.grid(age_years = seq(0, max_age, .1),
                                          target_child_sex = c("male","female"),
                                          language = included_languages) %>%
  as_data_frame()

mod_data$measure <- predict(mod, newdata = mod_data,
                            re.form = ~(target_child_sex + age_years | language))
summary(mod)
```
```{r echo=TRUE, warning=TRUE}
draw_gender_plot(data, mod_data)
```

Random effects structure does not include the interaction of gender and age in years. Worse, the fitted model doesn't pass through the origin. This suggests high MTLD values for children at 0 years

# Model 2: adding interaction in random effects

```{r echo=TRUE, warning=TRUE}
mod <- lmer(measure ~ target_child_sex * poly(age_years, 2) - 1 +
                    (1 | target_child_id) +
                    (target_child_sex * age_years | language),
  data = data, REML = TRUE, control = lmerControl(
                           optimizer ='optimx', optCtrl=list(method='nlminb')))
summary(mod)

mod_data <- expand.grid(age_years = seq(0, max_age, .1),
                                          target_child_sex = c("male","female"),
                                          language = included_languages) %>%
  as_data_frame()

mod_data$measure <- predict(mod, newdata = mod_data,
                            re.form = ~(target_child_sex * age_years | language))
```

```{r echo=TRUE, warning=TRUE}
draw_gender_plot(data, mod_data)
```

Random effects structure seems to be converging in this case (we were receiving non-convergence warnings previously).

# Model 3: Minimal Linear Model

```{r echo=TRUE, warning=TRUE}
mod <- lm(measure ~ 0 + target_child_sex * age_years,
  data = data)
summary(mod)

mod_data <- expand.grid(age_years = seq(0, max_age, .1),
                                          target_child_sex = c("male","female"),
                                          language = included_languages) %>%
  as_data_frame()

mod_data$measure <- predict(mod, newdata = mod_data)
draw_gender_plot(data, mod_data)
```

Because there is a main effect of gender (a factor), intercepts are nonzero.

# Model 4: Linear Model with no main effect of gender

```{r echo=TRUE, warning=TRUE}
mod <- lm(measure ~ 0 + age_years:target_child_sex,
  data = data)
summary(mod)

mod_data <- expand.grid(age_years = seq(0, max_age, .1),
                                          target_child_sex = c("male","female"),
                                          language = included_languages) %>%
  as_data_frame()

mod_data$measure <- predict(mod, newdata = mod_data)
draw_gender_plot(data, mod_data)
```

By removing the main effect of gender and just modeling its interaction with age, we achieve 0 intercepts and differing slopes by gender

```{r echo=TRUE, warning=TRUE}
mod_data$measure[0:10]
```

# Model 5: Linear Model with no main effect of gender, polynomial age

```{r echo=TRUE, warning=TRUE}
mod <- lm(measure ~ 0 + poly(age_years,2):target_child_sex,
  data = data)
summary(mod)

mod_data <- expand.grid(age_years = seq(0, max_age, .1),
                                          target_child_sex = c("male","female"),
                                          language = included_languages) %>%
  as_data_frame()

mod_data$measure <- predict(mod, newdata = mod_data)
draw_gender_plot(data, mod_data)
```

No idea what is going on here.

# Model 6: Linear Model, polynomial age only

```{r echo=TRUE, warning=TRUE}
mod <- lm(measure ~ poly(age_years,2),
  data = data)
summary(mod)

mod_data <- expand.grid(age_years = seq(0, max_age, .1),
                                          target_child_sex = c("male","female"),
                                          language = included_languages) %>%
  as_data_frame()

mod_data$measure <- predict(mod, newdata = mod_data)
draw_gender_plot(data, mod_data)
```
Intercept is close to zero but nonzero:
```{r echo=TRUE, warning=TRUE}
mod_data$measure[0:10]
```

# Model 7: Linear Model, polynomial age with specified intercept

```{r echo=TRUE, warning=TRUE}
mod <- lm(measure ~ poly(age_years,2) -1,
  data = data)
summary(mod)

mod_data <- expand.grid(age_years = seq(0, max_age, .1),
                                          target_child_sex = c("male","female"),
                                          language = included_languages) %>%
  as_data_frame()

mod_data$measure <- predict(mod, newdata = mod_data)
draw_gender_plot(data, mod_data)
```

Similar to Model 5, specifying an intercept of 0 with a polynomial term results in these weird fits

# Model 8: Mixed-effects model, no polynomials

```{r echo=TRUE, warning=TRUE}
mod <- lmer(measure ~ 0 +  target_child_sex:target_child_sex +
                    (1 | target_child_id) +
                    (0 + age_years:target_child_sex | language),
  data = data, REML = TRUE, control = lmerControl(
                           optimizer ='optimx', optCtrl=list(method='nlminb')))
summary(mod)

mod_data <- expand.grid(age_years = seq(0, max_age, .1),
                                          target_child_sex = c("male","female"),
                                          language = included_languages) %>%
  as_data_frame()

mod_data$measure <- predict(mod, newdata = mod_data,
                            re.form = ~(0 + age_years:target_child_sex | language))
```

```{r echo=TRUE, warning=TRUE}
draw_gender_plot(data, mod_data)
```

Intercept is close to zero but nonzero:
```{r echo=TRUE, warning=TRUE}
mod_data$measure[0:10]
```

### 2018-08-17 paper.Rmd
remove stroop, change age_months calculations

@@
   left_join(parents) %>%
-  mutate(age_months = target_child_age / 30.5,
+  mutate(age_months = target_child_age,
          age_bin = as.integer(floor(age_months / 6) * 6),
@@
 included_languages <- stats %>%
-  filter(target_child_age < max_age * 365) %>%       
+  filter(target_child_age < max_age * 12) %>%       
   count(language) %>% 
@@
   ungroup() %>%
-  mutate(age_years = target_child_age / 365,
+  mutate(age_years = target_child_age / 12,
          target_child_sex = factor(target_child_sex, 

### 2018-08-17 paper.Rmd
revised manuscript according to R1 suggestions

--- 2018-04-21 paper.Rmd
@@
 
-The content of CHILDES changes as additional corpora are added or transcriptions are updated; as of time of writing, these changes are not systematically tracked.
+The content of CHILDES changes as additional corpora are added or transcriptions are updated; as of time of writing, these changes are not systematically tracked in a public repository\footnote{Specific versions of the database, tracked using the version control system Git, can be obtained by emailing the maintainers of the CHILDES project.
+While tracking line-level changes with Git provides detailed information about what has changed, our method allows researchers to access the relevant version programatically.}.
 To facilitate reproducibility of past analyses, we introduce a simple versioning system by adding a new complete parse of the current state of CHILDES every six months or as warranted by changes in CHILDES.
@@
 
+`childes-db` also focuses strictly on the contents of CHILDES, and does not include material in related TalkBank projects such as PhonBank, AphasiaBank, or DementiaBank.
 We will prioritize the addition of these information sources and others in response to community feedback.
@@
 To this end we have created powerful yet easy-to-use interactive web applications that enable users to visualize various dimensions of the CHILDES corpus: frequency counts, mean lengths of utterance, type-token ratios, and more.
-All of this is doable without the requirement of understanding command-line tools or any kind of programming knowledge as had been the case with CLAN.^[The LuCiD toolkit [@chang2017] provides related functionality for a number of common analyses.
+All of this is doable without the requirement of understanding command-line tools.^[The LuCiD toolkit [@chang2017] provides related functionality for a number of common analyses.
 In contrast to those tools, which focus on filling gaps not covered by CLAN -- e.g., the use of $n$-gram models, incremental sentence generation, and distributional word classification -- our web apps focus on covering the same common tasks as CLAN, but making the outputs into browsable visualizations.]
@@
 We store these in the `colors` variable, and then use the `get_types` function from `childesr` to get the type frequency of each of these words in all of the corpora in CHILDES.
+All other functions are provided by base R or the `tidyverse` package. 
 For demonstration, we look only at the types produced by the speakers in each corpus tagged as Mother and Father.
@@
 The goal of this analysis was to showcase an example of using `childesr` to explore the CHILDES dataset.
-We also highlighted some of the potential pitfalls -- sparsity and non-independence -- that emerge in working with a diverse set of corpora, many of which were collected in longitudinal studies.
+We also highlighted some of the potential pitfalls -- sparsity and non-independance -- that emerge in working with a diverse set of corpora, many of which were collected in longitudinal studies.
 

### 2018-08-17 paper.Rmd
don't show warning in color counts

--- 2018-08-17 paper.Rmd
@@
 
-```{r join-colors-and-tokens, echo = TRUE}
+```{r join-colors-and-tokens, echo = TRUE, warning=F}
 count_estimates <- color_counts %>%

### 2018-08-19 paper.Rmd
minor formatting changes to text/code/plots

--- 2018-08-17 paper.Rmd
@@
 
-```{r load_packages, echo = FALSE, warning=FALSE}
+```{r load_packages, include=FALSE}
 library(papaja)
@@
 library(optimx)
-knitr::opts_chunk$set(cache = TRUE, echo = FALSE)
+library(grid)
+knitr::opts_chunk$set(cache = TRUE, echo = FALSE, message = FALSE,
+                      warning = FALSE, error = FALSE)
 ```
@@
 
-png::readPNG(here::here("images/schema.png")) %>% grid::grid.raster()
+readPNG(here("images/schema.png")) %>% grid.raster()
 ```
@@
 `childes-db` is stored as a MySQL database, an industry-standard, open-source relational database server that can be accessed directly from a wide range of programming languages.
-The childes-db project provides hosted, read-only databases for direct access and for `childesr` (described below) as well as compressed .sql exports for local installation.
+The `childes-db` project provides hosted, read-only databases for direct access and for `childesr` (described below) as well as compressed .sql exports for local installation.
 While the former is appropriate for most users, local installation can provide performance gains by allowing a user to access the database on their machine or on their local network, as well as allowing users to store derived information in the same database.
@@
 
-```{r, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
+```{r, eval=FALSE, include=FALSE}
 # plot error between clan counts and our counts for speaker unigrams
@@
   geom_density() +
-  xlim(0,.25)
+  xlim(0, .25)
 ```
@@
 
-# Interfaces for Accessing childes-db
+# Interfaces for Accessing `childes-db`
 
-We first discuss the `childes-db` web apps and then introduce the `childesr` `R` package.
+We first discuss the `childes-db` web apps and then introduce the `childesr` R package.
 
@@
 
-The ability to easily browse and explore the CHILDES corpora is a cornerstone of the childes-db project.
+The ability to easily browse and explore the CHILDES corpora is a cornerstone of the `childes-db` project.
 To this end we have created powerful yet easy-to-use interactive web applications that enable users to visualize various dimensions of the CHILDES corpus: frequency counts, mean lengths of utterance, type-token ratios, and more.
@@
 
-Our web apps are built using Shiny, an R package that enables easy app construction using `R`.
+Our web apps are built using Shiny, an R package that enables easy app construction using R.
 Underneath the hood, each web app is making calls to our `childesr` API and subsequently plots the data using the popular R plotting package `ggplot2`.
@@
 
-png::readPNG(here::here("images/freq_app.png")) %>% grid::grid.raster()
+readPNG(here("images/freq_app.png")) %>% grid.raster()
 ```
@@
 
-png::readPNG(here::here("images/derived_measures_app.png")) %>% grid::grid.raster()
+readPNG(here("images/derived_measures_app.png")) %>% grid.raster()
 ```
@@
 
-png::readPNG(here::here("images/population_viewer_app.png")) %>% grid::grid.raster()
+readPNG(here("images/population_viewer_app.png")) %>% grid.raster()
 ```
@@
 Although the interactive analysis tools described above cover some of the most common use cases of CHILDES data, researchers interested in more detailed and flexible analyses will want to interface directly with the data in `childes-db`.
-Making use of the `R` programming language [@R-base], we provide the `childesr` package.
-`R` is an open-source, extensible statistical computing environment that is rapidly growing in popularity across fields and is increasing in use in child language research [e.g.
+Making use of the R programming language [@R-base], we provide the `childesr` package.
+R is an open-source, extensible statistical computing environment that is rapidly growing in popularity across fields and is increasing in use in child language research [e.g.
 @song2015; @norrman2015].
@@
 
-# Using childes-db: Worked Examples
+# Using `childes-db`: Worked Examples
 
@@
 These frequency estimates are used both in the development of theory [e.g., frequent words are learned earlier; @goodman2008], and in the construction of age-appropriate experimental stimuli.
-One benefit of the childes-db interface is that it allows for easy analysis of how the frequencies of words change over development.
+One benefit of the `childes-db` interface is that it allows for easy analysis of how the frequencies of words change over development.
 Many of our theories in which children learn the structure of language from its statistical properties implicitly assume that these statistics are *stationary*, i.e.
@@
 
-#### Using childesr
+#### Using `childesr`
 
@@
 For this analysis, we will define our words of interest as the basic color words of English (except for gray, which children hear very rarely).
-We store these in the `colors` variable, and then use the `get_types` function from `childesr` to get the type frequency of each of these words in all of the corpora in CHILDES.
+We store these in the `colors` variable, and then use the `get_types()` function from `childesr` to get the type frequency of each of these words in all of the corpora in CHILDES.
 For demonstration, we look only at the types produced by the speakers in each corpus tagged as Mother and Father.
@@
                           role = c("Mother", "Father"),
-                          age = c(12,60), 
+                          age = c(12, 60),
                           type = colors)
@@
 To normalize correctly (i.e., to ask what proportion of the input children hear consists of these color words), we need to know how many total words these children hear from their parents in these transcripts.
-To do this, we use the `get_speaker_statistics` function, which will return a total number of tokens (`num_tokens`) for each of these speakers.
+To do this, we use the `get_speaker_statistics()` function, which will return a total number of tokens (`num_tokens`) for each of these speakers.
 
@@
 We then group the data into 6-month age bins, and compute the proportion of tokens that comprise each color for each child in each 6-month bin.
-<!-- KM: The back-to-back summarise() calls feel a little opaque.
-I think there is an implicit group_by() step?
--->
 For comparability with the web app analysis, these proportions are converted to parts per million words.
@@
   group_by(age_bin, color, target_child_id, transcript_id) %>%
-  summarise(count = sum(count), num_tokens = sum(num_tokens)) %>%
-  summarise(count = sum(count), num_tokens = sum(num_tokens)) %>%  
-  mutate(parts = count / num_tokens * 1e6)
+  summarise(transcript_count = sum(count),
+            transcript_num_tokens = sum(num_tokens)) %>%
+  group_by(age_bin, color, target_child_id) %>%
+  summarise(child_count = sum(transcript_count),
+            child_num_tokens = sum(transcript_num_tokens)) %>%
+  mutate(parts = child_count / child_num_tokens * 1e6)
 ```
@@
 ```{r color-plot, fig.height = 5, fig.width = 10, fig.cap ="Color frequency as a function of age.
-Points represent means across transcripts, error bars represent 95% confidence intervals computed by nonparametric bootstrap", cache = TRUE, message = FALSE, warning = FALSE}
+Points represent means across transcripts, error bars represent 95% confidence intervals computed by nonparametric bootstrap.", cache = TRUE}
 ggplot(count_estimates_with_error, aes(x = age_bin, y = empirical_stat,
@@
   scale_y_continuous(name = "Frequency (per million words)") +
-  geom_dl(method = list(dl.trans(x=x +.2), "last.qp", cex=1)) +
+  geom_dl(method = list(dl.trans(x = x + .2), "last.qp", cex = 1)) +
   theme(legend.position = "none")
@@
 
-First, we use the `childesr` function call `get_speaker_statistics` to pull data relating to the aforementioned derived measures for children and their transcripts.
+First, we use the `childesr` function call `get_speaker_statistics()` to pull data relating to the aforementioned derived measures for children and their transcripts.
 Note that we exclusively select the children's production data, and exclude their caregivers' speech.
@@
 ```{r echo=TRUE}
-stats <- get_speaker_statistics(role = "Target_Child")
+speaker_stats <- get_speaker_statistics(role = "Target_Child")
 ```
@@
 
-included_languages <- stats %>%
+included_languages <- speaker_stats %>%
   filter(target_child_age < max_age * 12) %>%
@@
 ```{r echo=TRUE}
-data <- stats %>%
+mtld_data <- speaker_stats %>%
   filter(!is.na(target_child_sex), target_child_name != "Leo",
          language %in% included_languages) %>%
-  group_by(target_child_id, target_child_age, 
-           target_child_sex, language) %>%
+  group_by(target_child_id, target_child_age, target_child_sex, language) %>%
   summarise(measure = mean(mtld)) %>%
@@
          target_child_sex = factor(target_child_sex,
-                                   levels = c("male","female"))) %>%
+                                   levels = c("male", "female"))) %>%
   filter(age_years < max_age)
@@
 
-```{r echo=FALSE}
+```{r}
 mod <- lmer(measure ~ target_child_sex * poly(age_years, 2) - 1 +
@@
               (target_child_sex + age_years | language),
-  data = data, REML = TRUE, control = lmerControl(
-                           optimizer ='optimx', optCtrl=list(method='nlminb')))
+            data = mtld_data, REML = TRUE, control = lmerControl(
+              optimizer = "optimx", optCtrl = list(method = "nlminb")))
 
 mod_data <- expand.grid(age_years = seq(0, max_age, .1),
-                                          target_child_sex = c("male","female"),
+                        target_child_sex = c("male", "female"),
                         language = included_languages) %>%
@@
 
-mod_data$measure <- predict(mod, newdata = mod_data,
-                            re.form = ~(target_child_sex + age_years | language))
+mod_data$measure <- predict(
+  mod, newdata = mod_data, re.form = ~(target_child_sex + age_years | language)
+)
 ```
 
-```{r echo=FALSE, warning=FALSE}
-ggplot(data, aes(x = age_years, y = measure,
-                 fill = target_child_sex, colour = target_child_sex)) +
-  geom_point(alpha = .3, size=1, stroke=0) +
-  geom_line(data = mod_data,
-            aes(x = age_years, y = measure, color = target_child_sex)) +
+```{r, fig.asp = 0.9}
+language_codes <- c("deu" = "German", "eng" = "English", "fra" = "French",
+                    "jpn" = "Japanese", "nld" = "Dutch", "spa" = "Spanish",
+                    "zho" = "Chinese")
+
+plt_mtld_data <- mtld_data %>%
+  mutate(language = language_codes[language])
+plt_mod_data <- mod_data %>%
+  mutate(language = language_codes[language])
+
+ggplot(plt_mtld_data, aes(x = age_years, y = measure, fill = target_child_sex,
+                          colour = target_child_sex)) +
+  geom_point(alpha = .3, size = 1, stroke = 0) +
+  geom_line(aes(x = age_years, y = measure, color = target_child_sex),
+            data = plt_mod_data) +
   facet_wrap(~language) +
-  ggthemes::theme_few() +
-  ggthemes::scale_color_ptol(name = "Sex") +
-  xlim(0,max_age) +
-  ylim(0,15) +
+  theme_few() +
+  scale_color_ptol(name = "Sex") +
+  scale_fill_ptol(guide = FALSE) +
+  xlim(0, max_age) +
+  ylim(0, 15) +
   xlab("Age (years)") +
   ylab("MTLD") +
-  theme(legend.position = "bottom")
+  theme(legend.position = "top",
+        legend.box.spacing = unit(0, "mm"))
 ```
@@
 
-## Teaching with childes-db
+## Teaching with `childes-db`
 
@@
 2. allow students to reflect or do a pair-and-share discussion with their neighbor,
-3. show the trajectory of a single lexical item while explaining key parts of the visualization (see Panel A of Figure \ref{fig:in-class-demo-plot}),
+3. show the trajectory of a single lexical item while explaining key parts of the visualization (see Panel a of Figure \ref{fig:in-class-demo-plot}),
 4. elicit hypotheses from students about the kinds of words that children are likely to produce,
-5. make real-time queries to the web application to add students' suggestions and talk through the updated plots (Panels B and C of Figure \ref{fig:in-class-demo-plot}), and
-6. finish by entering a pre-selected set of words that communicate the important takeaway point (Panel D of Figure \ref{fig:in-class-demo-plot}) .
+5. make real-time queries to the web application to add students' suggestions and talk through the updated plots (Panels b and c of Figure \ref{fig:in-class-demo-plot}), and
+6. finish by entering a pre-selected set of words that communicate the important takeaway point (Panel d of Figure \ref{fig:in-class-demo-plot}).
 
 ```{r in-class-demo-plot, fig.pos="h", out.width = "95%", fig.cap="Worked example of using the web applications for in-class teaching.
-Panels A-D show how an instructor could dynamically build a plot during a lecture to demonstrate a key concept in language acquisition."}
+Panels a-d show how an instructor could dynamically build a plot during a lecture to demonstrate a key concept in language acquisition."}
 
-png::readPNG(here::here("images/in_class_teach_viz.png")) %>% grid::grid.raster()
+readPNG(here("images/in_class_teach_viz.png")) %>% grid.raster()
 ```
@@
 We have presented `childes-db`, a database formatted mirror of the CHILDES dataset.
-This database -- together with the `R` API and web apps -- facilitates the use of child language data.
+This database -- together with the R API and web apps -- facilitates the use of child language data.
 For teachers, students, and casual explorers, the web apps allow browsing and demonstration.
@@
 
-childes-db addresses a number of needs that have emerged in our own research and teaching, but there are still a number of limitations that point the way to future improvements.
+`childes-db` addresses a number of needs that have emerged in our own research and teaching, but there are still a number of limitations that point the way to future improvements.
 For example, `childes-db` currently operates only on transcript data, without links to the underlying media files; in the future, adding such links may facilitate further computational and manual analyses of phonology, prosody, social interaction, and other phenomena by providing easy access to the video and audio data.
@@
 # References
-```{r create_r-references, warning = FALSE}
+```{r create_r-references}
 r_refs(file = "childesdb.bib")

### 2018-08-22 paper.Rmd

@@
 For this analysis, we will define our words of interest as the basic color words of English (except for gray, which children hear very rarely).
-We store these in the `colors` variable, and then use the `get_types()` function from `childesr` to get the type frequency of each of these words in all of the corpora in CHILDES.
+We store these in the `colors` variable, and then use the `get_types` function from `childesr` to get the type frequency of each of these words in all of the corpora in CHILDES.
+All other functions are provided by base R or the `tidyverse` package. 
 For demonstration, we look only at the types produced by the speakers in each corpus tagged as Mother and Father.

### 2018-08-22 paper.Rmd
Revert "fixed outstanding text change"
This reverts commit 4ff124982150176eaf755543545a5c964f6c6db9, reversing
changes made to d9ab2260238999401e1a5f66997395177efd3729.

--- 2018-08-17 paper.Rmd
@@
 
-```{r join-colors-and-tokens, echo = TRUE, warning=F}
+```{r join-colors-and-tokens, echo = TRUE}
 count_estimates <- color_counts %>%

### 2018-08-23 paper.Rmd
merged all previous changes from merge issue in Git; small grammar fixes; clarified role of CLANtoR; clarified description of CHAT tiers

--- 2018-08-19 paper.Rmd
@@
 author_note: |
+   Co-first authorship indicated with $*$.
 Thanks to Brian MacWhinney for advice and guidance, and to Melissa Kline for her work on ClanToR, which formed a starting point for our work.
@@
   Access to these data can be both complex for novices and difficult to automate for advanced users, however.
-  To address these issues, we introduce `childes-db`, a database-formatted mirror of CHILDES that improves data accessibility and usability by offering novel interfaces, including browsable web applications and an R application programming interface (API).
+  To address these issues, we introduce `childes-db`, a database-formatted mirror of CHILDES that improves data accessibility and usability by offering novel interfaces, including browsable web applications and an `R` application programming interface (API).
   Along with versioned infrastructure that facilitates reproducibility of past analyses, these interfaces lower barriers to analyzing naturalistic parent-child language, allowing for a wider range of researchers in language and cognitive development to easily leverage CHILDES in their work.
@@
 Individual studies are too numerous to list, but classics include studies of morphological over-regularization [@marcus1992], distributional learning [@redington1998], word segmentation [@goldwater2009], the role of frequency in word learning [@goodman2008], and many others.
-Some studies analyze individual examples in depth [e.g., @snyder2007], others track multiple child-caregiver dyads [e.g., @meylan2017], and still others use the aggregate properties of all child or caregiver speech pooled across corpora [e.g., @redington1998; @montagEtAl2015].
+Some studies analyze individual examples in depth [e.g., @snyder2007], some track multiple child-caregiver dyads [e.g., @meylan2017], and still others use the aggregate properties of all child or caregiver speech pooled across corpora [e.g., @redington1998; @montagEtAl2015].
 <!-- These are just a few of the thousands of studies that make use of CHILDES corpora. -->
@@
 Nonetheless, there are some outstanding challenges working with CHILDES, both for students and for advanced users.
-The CHILDES ecosystem uses a specialized file format (CHAT), which is stored as plain text but includes structured annotations grouped into tiers stored on separate lines.
-These tiers allow information about utterances to be stored with accompanying information such as the phonological, morphological, or syntactic structure of the utterance.
+The CHILDES ecosystem uses a specialized file format (CHAT), which is stored as plain text but includes structured annotations grouped into parallel information ``tiers'' on separate lines.
+These tiers allow for a searchable plaintext transcript of an utterance to be stored along with structured annotations of its phonological, morphological, or syntactic content.
 These files are usually analyzed using a command-line program (CLAN) that allows users to count word frequencies, compute statistics (e.g., mean length of utterance, or MLU), and execute complex searches against the data.
@@
 CLAN software is an excellent tool for interactive exploration, but -- as a free-standing application -- it can be tricky to build into a processing pipeline written in Python or R.
-Thus, researchers who would like to ingest the entire corpus (or some large subset) into a computational analysis typically write their own parsers of the CHAT format to extract the subset of the data they would like to use [e.g., @redington1998;@yang2013;@meylan2017].
+Thus, researchers who would like to ingest the entire corpus (or some large subset) into a computational analysis typically write their own parsers of the CHAT format to extract the subset of the data they would like to use [e.g., @klineCLANtoR; @redington1998;@yang2013;@meylan2017].
 
@@
 Third, these parsing scripts are rarely shared -- and when when they are, they typically break with subsequent revisions to the dataset -- leading to much greater difficulty in reproducing the exact numerical results from previous published research that used CHILDES [see e.g., @meylan2017 for an example].
-Fourth, the CHILDES corpus itself is a moving target: computational work using the entire corpus at one time point may include a different set of data than subsequent work due as corpora are added and revised.
+Fourth, the CHILDES corpus itself is a moving target: computational work using the entire corpus at one time point may include a different set of data than subsequent work as corpora are added and revised.
 Currently, there is no simple way for researchers to document exactly which version of the corpus has been used, short of creating a full mirror of the data.
@@
 We begin by describing the architecture of `childes-db` and the web applications that we provide.
-Next, we describe the `childesr` API, which provides a set of R functions for programmatic access to the data while abstracting away many of the technical details.
-We conclude by presenting several worked examples of specific uses of the system -- both web apps and the R API -- for research and teaching.
+Next, we describe the `childesr` API, which provides a set of `R` functions for programmatic access to the data while abstracting away many of the technical details.
+We conclude by presenting several worked examples of specific uses of the system -- both web apps and the `R` API -- for research and teaching.
 
@@
 As described above, CHILDES is most often approached as a set of distinct CHAT files, which are then parsed by users, often using CLAN.
-In contrast to this parsing approach, which entails the sequential processing of strings, `childes-db` treats CHILDES as a set of linked tables, with records corresponding to intuitive abstractions such as words, utterances, and transcripts [see @klineCLANtoR for an earlier example of deriving tabular representations of CHILDES].
-Users of data analysis languages like R or Julia, libraries like Pandas, or those familiar with Structured Query Language (SQL) will be familiar with operations on tables such as filtering (subsetting), sorting, aggregation (grouping), and joins (merges).
+In contrast to this parsing approach, which entails the sequential processing of strings, `childes-db` treats CHILDES as a set of linked tables, with records corresponding to intuitive abstractions such as words, utterances, and transcripts [see @klineCLANtoR for an earlier example of deriving a singular tabular representation of a CHILDES transcript].
+Users of data analysis languages like `R` or Julia, libraries like Pandas, or those familiar with Structured Query Language (SQL) will be familiar with operations on tabular representations of data such as filtering (subsetting), sorting, aggregation (grouping), and joins (merges).
 These operations obviate the need for users to consider the specifics of the CHAT representation -- instead they simply request the entities they need for their research and allow the API to take care of the formatting details.
@@
 Using the standardized written form of the word facilitates the computation of lexical frequency statistics for comparison or aggregation across children or time periods.
-Deviations from the citation form -- which are particularly common in the course of language development and often of interest to researchers -- are kept as a separate (possibly null) field associated with each token.
+Deviations from the citation form -- which are particularly common in the course of language development and often of special interest to researchers -- are kept as a separate (possibly null) field associated with each token.
 
@@
 
-Many of the other tables in the database are hierarchical collections built out of tokens -- *utterance*, *transcript*, *corpus*, and *collection* -- that store attributes appropriate for each level of description.
+Many of the other tables in the database describe hierarchical collections built out of tokens -- *utterance*, *transcript*, *corpus*, and *collection* -- and store attributes appropriate for each level of description.
 Every entity includes attributes that link it to all higher-order collections, e.g., an utterance lists the transcript, corpus, and collection to which it belongs.
-An *utterance* contains one or more words and includes fields such as the utterance type such as *declarative* or *interrogative*, total number of tokens, and the total number of morphemes if the morphological structure is available in the original CHAT file.
-A *transcript* consists of one or more utterances and includes the date collected, the name of the target child, and age in days if defined, and the filename from CHILDES.
-A *corpus* consists of one or more transcripts, corresponding to well-known collections like the Brown [@brown1973] or Providence [@demuth2006word] corpora.
+An *utterance* contains one or more word tokens and includes fields such as the utterance type (e.g.,  *declarative*, *interrogative*, etc.), total number of tokens, and the total number of morphemes if the morphological structure is available in the original CHAT file.
+A *transcript* consists of one or more utterances and includes the date collected, the name of the target child, the age in days if defined, and the filename from CHILDES.
+A *corpus* consists of one or more transcripts, corresponding to well-known collections like the Brown [@brown1973] or Providence [@demuth2006word] corpus.
 Finally, a *collection* is a superordinate collection of corpora generally corresponding to a geographic region, following the convention in CHILDES.
@@
 Participants -- generally children and caregivers -- are represented separately from the token hierarchy because it is common for the same children to appear in multiple transcripts.
-A participant identifier is associated with every word and utterance, including a name, role, 3-letter CHILDES identifier (CHI = child, MOT = mother, FAT = father, etc.), and the range of ages (or age of corresponding child) for which they are observed.
+A participant identifier is associated with every word and utterance, including a name, role, 3-letter CHILDES identifier (CHI = child, MOT = mother, FAT = father, etc.), and the range of ages for which they are observed (or age of corresponding child, in the case of caregivers).
 For non-child participants (caregivers and others), the record additionally contains an identifier for the corresponding target child, such that data corresponding to children and their caregivers can be easily associated.
@@
 
-`childes-db` is stored as a MySQL database, an industry-standard, open-source relational database server that can be accessed directly from a wide range of programming languages.
-The `childes-db` project provides hosted, read-only databases for direct access and for `childesr` (described below) as well as compressed .sql exports for local installation.
+`childes-db` is stored as a MySQL database, an industry-standard, open-source relational database that can be accessed directly from a wide range of programming languages.
+The `childes-db` project provides access to hosted, read-only databases on a publicly-accessible server for direct access and `childesr` (described below).
+The project also hosts compressed .sql exports for local installation.
 While the former is appropriate for most users, local installation can provide performance gains by allowing a user to access the database on their machine or on their local network, as well as allowing users to store derived information in the same database.
@@
 
-The content of CHILDES changes as additional corpora are added or transcriptions are updated; as of time of writing, these changes are not systematically tracked.
+The content of CHILDES changes as additional corpora are added or transcriptions are updated; as of time of writing, these changes are not systematically tracked in a public repository.\footnote{Specific versions of the database, tracked using the version control system Git, can be obtained by emailing the maintainers of the CHILDES project.
+While tracking line-level changes with Git provides detailed information about what has changed, our method allows researchers to access the relevant version programatically by simply adding an argument to a function call.}
 To facilitate reproducibility of past analyses, we introduce a simple versioning system by adding a new complete parse of the current state of CHILDES every six months or as warranted by changes in CHILDES.
@@
 To support reproduction of results with previous versions of the database, we continue to host recent versions (up to the last three years / six versions) through our `childesr` API so that researchers can run analyses against specific historical versions of the database.
-For versions more than three years old, we host compressed .sql files that users may download and serve using a local installation of MySQL server.
+For versions more than three years old, we host compressed .sql files that users may download and serve using a local installation of MySQL server (for which we provide instructions).
 <!-- SM: How easy is it to implement versioning in the Shiny apps?
@@
 
+\noindent At present, `childes-db` focuses strictly on the contents of CHILDES, and does not include material in related TalkBank projects such as PhonBank, AphasiaBank, or DementiaBank.
 We will prioritize the addition of these information sources and others in response to community feedback.
@@
 
-We first discuss the `childes-db` web apps and then introduce the `childesr` R package.
+We first discuss the `childes-db` web apps and then introduce the `childesr` `R` package.
 
@@
 To this end we have created powerful yet easy-to-use interactive web applications that enable users to visualize various dimensions of the CHILDES corpus: frequency counts, mean lengths of utterance, type-token ratios, and more.
-All of this is doable without the requirement of understanding command-line tools or any kind of programming knowledge as had been the case with CLAN.^[The LuCiD toolkit [@chang2017] provides related functionality for a number of common analyses.
-In contrast to those tools, which focus on filling gaps not covered by CLAN -- e.g., the use of $n$-gram models, incremental sentence generation, and distributional word classification -- our web apps focus on covering the same common tasks as CLAN, but making the outputs into browsable visualizations.]
+All of this is doable without the requirement of understanding command-line tools.^[The LuCiD toolkit [@chang2017] provides related functionality for a number of common analyses.
+In contrast to those tools, which focus on filling gaps not covered by CLAN -- e.g., the use of $n$-gram models, incremental sentence generation, and distributional word classification -- our web apps focus on covering the same common tasks as CLAN, but yielding visualizations for the web browser.]
 
-Our web apps are built using Shiny, an R package that enables easy app construction using R.
-Underneath the hood, each web app is making calls to our `childesr` API and subsequently plots the data using the popular R plotting package `ggplot2`.
+Our web apps are built using Shiny, a software package that enables easy app construction using `R`.
+Underneath the hood, each web app is making calls to our `childesr` API and subsequently plots the data using the popular `R` plotting package `ggplot2`.
 A user's only task is to configure exactly what should be plotted through a series of buttons, sliders, and text boxes.
@@
 As with the Frequency Counts app, caregivers' lexical diversity measures can be plotted alongside children's.
-
 We have currently implemented the following measures:
@@
 
-Many times a researcher will want to investigate the statistics of corpora (e.g., their size, number of utterances, number of tokens) before choosing a target corpus or set of corpora for a project.
-This web app is intended to provide a basic overview regarding the scale and temporal extent of various corpora in CHILDES, as well as giving researchers insight into the aggregate characteristics of CHILDES.
+In many cases a researcher may want to view the statistics and properties of corpora (e.g., their size, number of utterances, number of tokens) before choosing a target corpus or set of corpora for an analysis.
+This web app is intended to provide a basic overview regarding the scale and temporal extent of various corpora in CHILDES, as well as give researchers insight into the aggregate characteristics of CHILDES.
 For example, examining the aggregate statistics reveals that coverage in CHILDES peaks at around 30 months.
@@
 Although the interactive analysis tools described above cover some of the most common use cases of CHILDES data, researchers interested in more detailed and flexible analyses will want to interface directly with the data in `childes-db`.
-Making use of the R programming language [@R-base], we provide the `childesr` package.
-R is an open-source, extensible statistical computing environment that is rapidly growing in popularity across fields and is increasing in use in child language research [e.g.
+Making use of the `R` programming language [@R-base], we provide the `childesr` package.
+`R` is an open-source, extensible statistical computing environment that is rapidly growing in popularity across fields and is increasing in use in child language research [e.g.
 @song2015; @norrman2015].
@@
 Users can take advantage of the tools developed in the popular `dplyr` package [@R-dplyr], which makes manipulating large datasets quick and easy.
-We describe the commands that the package provides and then give several worked examples of using the package for analyses.
+We describe the commands that the package provides and then give several worked examples of analyses using the package.
 
-The `childesr` package is easily installed via CRAN, the comprehensive R archive network.
+The `childesr` package is easily installed via CRAN, the comprehensive `R` archive network.
 To install, simply type: `install.packages("childesr")`. After installation, users have access to functions that can be used to retrieve tabular data from the database:
@@
 
-In this section we give a number of examples of how `childes-db` can be used in both research and teaching, using both the web apps and the R API.
+In this section we give a number of examples of how `childes-db` can be used in both research and teaching, using both the web apps and the `R` API.
 Note that all of these examples use `dplyr` syntax [@R-dplyr]; several accessible introductions to this framework are available online [e.g., @wickham2016].
@@
 
-```{r color-shiny, out.width = "500px", fig.cap = "An example of using the Frequency shiny app to explore how children's color input changes over development"}
+```{r color-shiny, out.width = "500px", fig.cap = "An example of using the Frequency shiny app to explore how children's color input changes over development."}
 knitr::include_graphics("images/color-shiny.png")
@@
 We store these in the `colors` variable, and then use the `get_types()` function from `childesr` to get the type frequency of each of these words in all of the corpora in CHILDES.
+All other functions are provided by base `R` or the `tidyverse` package. 
 For demonstration, we look only at the types produced by the speakers in each corpus tagged as Mother and Father.
@@
   group_by(age_bin, color, target_child_id, transcript_id) %>%
-  summarise(transcript_count = sum(count),
-            transcript_num_tokens = sum(num_tokens)) %>%
+  summarise(transcript_count = sum(count), transcript_num_tokens = sum(num_tokens)) %>%
   group_by(age_bin, color, target_child_id) %>%
-  summarise(child_count = sum(transcript_count),
-            child_num_tokens = sum(transcript_num_tokens)) %>%
+  summarise(child_count = sum(transcript_count), child_num_tokens = sum(transcript_num_tokens)) %>%
   mutate(parts = child_count / child_num_tokens * 1e6)
@@
 ```{r color-plot, fig.height = 5, fig.width = 10, fig.cap ="Color frequency as a function of age.
-Points represent means across transcripts, error bars represent 95% confidence intervals computed by nonparametric bootstrap.", cache = TRUE}
+Points represent means across transcripts, error bars represent 95% confidence intervals computed by nonparametric bootstrap.
+Y-axes are free because non-stationarity is evaluated within each color word, while their absolute frequencies vary widely.", cache = TRUE}
 ggplot(count_estimates_with_error, aes(x = age_bin, y = empirical_stat,
@@
   scale_y_continuous(name = "Frequency (per million words)") +
-  geom_dl(method = list(dl.trans(x = x + .2), "last.qp", cex = 1)) +
+  geom_dl(method = list(dl.trans(x = x +.2), "last.qp", cex = 1)) +
   theme(legend.position = "none")
@@
          target_child_sex = factor(target_child_sex, 
-                                   levels = c("male", "female"))) %>%
+                                   levels = c("male","female"))) %>%
   filter(age_years < max_age)
@@
 
-```{r}
+```{r echo=FALSE}
 mod <- lmer(measure ~ target_child_sex * poly(age_years, 2) - 1 +
@@
   data = mtld_data, REML = TRUE, control = lmerControl(
-              optimizer = "optimx", optCtrl = list(method = "nlminb")))
+                           optimizer ='optimx', optCtrl=list(method='nlminb')))
 
@@
 
-mod_data$measure <- predict(
-  mod, newdata = mod_data, re.form = ~(target_child_sex + age_years | language)
-)
+mod_data$measure <- predict(mod, newdata = mod_data, re.form = ~(target_child_sex + age_years | language))
 ```
@@
 
-## Teaching with `childes-db`
+## Teaching with childes-db
 
@@
 2. allow students to reflect or do a pair-and-share discussion with their neighbor,
-3. show the trajectory of a single lexical item while explaining key parts of the visualization (see Panel a of Figure \ref{fig:in-class-demo-plot}),
+3. show the trajectory of a single lexical item while explaining key parts of the visualization (see Panel A of Figure \ref{fig:in-class-demo-plot}),
 4. elicit hypotheses from students about the kinds of words that children are likely to produce,
-5. make real-time queries to the web application to add students' suggestions and talk through the updated plots (Panels b and c of Figure \ref{fig:in-class-demo-plot}), and
-6. finish by entering a pre-selected set of words that communicate the important takeaway point (Panel d of Figure \ref{fig:in-class-demo-plot}).
+5. make real-time queries to the web application to add students' suggestions and talk through the updated plots (Panels B and C of Figure \ref{fig:in-class-demo-plot}), and
+6. finish by entering a pre-selected set of words that communicate the important takeaway point (Panel D of Figure \ref{fig:in-class-demo-plot}) .
 
 ```{r in-class-demo-plot, fig.pos="h", out.width = "95%", fig.cap="Worked example of using the web applications for in-class teaching.
-Panels a-d show how an instructor could dynamically build a plot during a lecture to demonstrate a key concept in language acquisition."}
+Panels A-D show how an instructor could dynamically build a plot during a lecture to demonstrate a key concept in language acquisition."}
 
@@
 We have presented `childes-db`, a database formatted mirror of the CHILDES dataset.
-This database -- together with the R API and web apps -- facilitates the use of child language data.
+This database -- together with the `R` API and web apps -- facilitates the use of child language data.
 For teachers, students, and casual explorers, the web apps allow browsing and demonstration.

### 2018-08-28 paper.Rmd
captions, hold figures, spelling corrections

@@
 library(grid)
-knitr::opts_chunk$set(cache = TRUE, echo = FALSE, message = FALSE,
+knitr::opts_chunk$set(fig.pos = "H", cache = TRUE, echo = FALSE, message = FALSE,
                        warning = FALSE, error = FALSE)
@@
 The content of CHILDES changes as additional corpora are added or transcriptions are updated; as of time of writing, these changes are not systematically tracked in a public repository.\footnote{Specific versions of the database, tracked using the version control system Git, can be obtained by emailing the maintainers of the CHILDES project.
-While tracking line-level changes with Git provides detailed information about what has changed, our method allows researchers to access the relevant version programatically by simply adding an argument to a function call.}
+While tracking line-level changes with Git provides detailed information about what has changed, our method allows researchers to access the relevant version programmatically by simply adding an argument to a function call.}
 To facilitate reproducibility of past analyses, we introduce a simple versioning system by adding a new complete parse of the current state of CHILDES every six months or as warranted by changes in CHILDES.
@@
 In this web app, inspired by ChildFreq [@baath2010], we provide users the ability to search for any word spoken by a participant in the CHILDES corpora and track the usage of that word by a child or caregiver over time.
-Because of the various toggles available to the user that can subset the data, a user may word frequencies curves for a single child in the Brown corpus or all Spanish speaking children, if desired.
+Because of the various toggles available to the user that can subset the data, a user may view word frequency curves for a single child in the Brown corpus or all Spanish speaking children, if desired.
 In addition, users can plot frequency curves belonging to caregivers alongside their child for convenient side-by-side comparisons.
@@
 
-```{r frequency-app2, fig.pos="h", out.width = "95%", fig.cap="Frequency Counts."}
+```{r frequency-app2, fig.pos="h", out.width = "95%", fig.cap="The Frequency Counts application allows users to track the frequency of words across various subgroups of children."}
 
@@
 
-```{r derived-measures-app2, fig.pos="h", out.width = "95%", fig.cap="Derived Measures."}
+```{r derived-measures-app2, fig.pos="h", out.width = "95%", fig.cap="The Derived Measures application allows users to view several measures of children’s speech in CHILDES that operationalize complexity and diversity."}
 
@@
 
-```{r population-viewer-app, fig.pos="h", out.width = "95%", fig.cap="Population Viewer."}
+```{r population-viewer-app, fig.pos="h", out.width = "95%", fig.cap="The Population Viewer application allows users to investigate the statistics of corpora in CHILDES."}
 
@@
 
-```{r, fig.asp = 0.9}
+```{r, fig.asp = 0.9, fig.cap="MTLD (Measure of Textual Lexical Diversity) scores as a function of age.
+Points represent mean scores for individual children across transcripts.
+Females have a slight advantage in linguistic productivity over males.", out.extra = ''}
 language_codes <- c("deu" = "German", "eng" = "English", "fra" = "French",
